---
title: "Livestock diet items"
author:
  - name: "Lolita Muller"
    email: "m.lolita@cgiar.org"
    orcid: "0009-0002-8345-7219"
  - name: "Peter Steward"
    email: "p.steward@cgiar.org"
    orcid: "0000-0003-3985-4911"
  - name: "Namita Joshi"
    email: "n.joshi@cgiar.org"
    orcid: "0000-0002-1623-6910"
  - name: "Todd Rosenstock"
    email: "t.rosenstock@cgiar.org"
    orcid: "0000-0002-1958-9500"
date: "2025-02-19"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show

---

## Introduction


### Objectives
This analysis implements a comprehensive, automated workflow for processing livestock diet data with the following key goals:

1. **Download and Load Metadata:**  
   Retrieve the latest livestock diet metadata from an S3 bucket, ensuring that the data are up-to-date.

2. **Harmonize Ingredient Units:**  
   Standardize various weight, volume, and time units (e.g., converting kg to grams, liters to milliliters, and weeks to days) to ensure consistent measurements across studies.

3. **Link Diet Ingredients to Feedapedia:**  
   Merge diet ingredient data with external nutritional and digestibility information from Feedapedia by matching corresponding feed ingredient codes.

4. **Integrate Basal Diets:**  
   Combine basal (control) diet components with experimental diets, while preserving the identity of the basal ingredients through appropriate flagging.

5. **Impute Missing Amounts Using Feed Intake Data:**  
   Where ingredient amounts are missing, infer values from feed intake records and flag these instances, thus improving dataset completeness.

6. **Calculate Relative Ingredient Proportions:**  
   Compute the percentage contribution of each ingredient within a complete diet, ensuring that only diets with consistent units are used for accurate calculations.

7. **Merge Nutritional and Digestibility Information:**  
   Create a comprehensive diet-level table that integrates total feed amounts with nutritional and digestibility metrics from both the original metadata and Feedapedia.

8. **Track Data Filtering and Potential Loss:**  
   Monitor the impact of each processing step—from unit harmonization to the filtering of inconsistent data—and document any data loss or special cases (such as ad libitum feeding).

By automating these steps, this workflow not only streamlines data preparation for livestock nutrition research but also ensures reproducibility and robust analyses across diverse studies.


## 1) Set up 
Download the packages and download the data from the s3 bucket.

```{r,message=FALSE,warning=FALSE}
# Install and load 'pacman' package manager if not already installed

library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)
library(readxl)
library(tidyr)
library(arrow)



# Set up connection to S3 bucket
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("skinny_cow_2022", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
livestock_metadata <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))


#total number of studies 
total_studies<- length(unique(livestock_metadata$Animals.Diet$B.Code))
total_ingredients<- length(unique(livestock_metadata$Animals.Diet$D.Item))
total_diets<- length(unique(livestock_metadata$Animals.Diet$A.Level.Name))



```



## 2) Harmonize units 


```{r,message=FALSE,warning=FALSE}
# diet ingredients: breakdown of all ingredients within each diet
ingredients <- livestock_metadata$Animals.Diet %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item.AOM,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    D.Unit.Time,
    D.Unit.Animals,
    DC.Is.Dry,
    D.Ad.lib,
    D.Is.Group,
    D.Item.Root.Other.Comp.Proc_All
  ) %>%
  mutate(D.Amount = as.numeric(D.Amount))

# Function to harmonize diet ingredient units
harmonize_units <- function(df) {
  df %>%
    mutate(
      # Record original units (optional, for debugging/traceability)
      Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"),

      # Harmonize weight and volume units to a consistent base
      D.Amount = case_when(
        D.Unit.Amount %in% c("kg", "mg", "g/100g", "l", "kg/t", "g/300l", 
                             "mg/kg", "kg/100kg", "kg/kg metabolic weight", 
                             "kg/100kg body weight", "g/kg metabolic weight (0.75)", 
                             "g/L") ~ {
          case_when(
            D.Unit.Amount == "kg" ~ D.Amount * 1000,                      # kg -> g
            D.Unit.Amount == "mg" ~ D.Amount / 1000,                     # mg -> g
            D.Unit.Amount == "g/100g" ~ D.Amount * 10,                   # g/100g -> g/kg
            D.Unit.Amount == "l" ~ D.Amount * 1000,                      # l -> ml (assumed g ~ ml in water)
            D.Unit.Amount == "kg/t" ~ D.Amount * 1000,                   # kg/t -> g/kg
            D.Unit.Amount == "g/300l" ~ D.Amount / 300,                  # g/300l -> g/L
            D.Unit.Amount == "mg/kg" ~ D.Amount / 1000,                  # mg/kg -> g/kg
            D.Unit.Amount == "kg/100kg body weight" ~ D.Amount * 10,     # -> g/kg body weight
            D.Unit.Amount == "kg/kg metabolic weight" ~ D.Amount * 1000, # -> g/kg metabolic weight
            TRUE ~ D.Amount
          )
        },
        TRUE ~ D.Amount
      ),

      # Update D.Unit.Amount to reflect new base
      D.Unit.Amount = case_when(
        D.Unit.Amount == "kg"  ~ "g",
        D.Unit.Amount == "mg"  ~ "g",
        D.Unit.Amount == "g/100g" ~ "g/kg",
        D.Unit.Amount == "l"  ~ "ml",
        D.Unit.Amount == "kg/t" ~ "g/kg",
        D.Unit.Amount == "g/300l" ~ "g/L",
        D.Unit.Amount == "mg/kg" ~ "g/kg",
        D.Unit.Amount == "kg/100kg body weight" ~ "g/kg body weight",
        D.Unit.Amount == "kg/kg metabolic weight" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg metabolic weight (0.75)" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/L" ~ "g/L",
        TRUE ~ D.Unit.Amount
      ),

      # Harmonize time units to day
      D.Amount = case_when(
        D.Unit.Time == "week" ~ D.Amount / 7,
        D.Unit.Time == "month" ~ D.Amount / 30,
        D.Unit.Time == "2 weeks" ~ D.Amount / 14,
        D.Unit.Time == "4 days interval" ~ D.Amount / 4,
        D.Unit.Time == "2x/day" ~ D.Amount * 2,
        TRUE ~ D.Amount
      ),
       D.Unit.Time = ifelse(!is.na(D.Unit.Time), "day", D.Unit.Time)
    ) %>%
    # NEW: If D.Amount is NA, set unit fields to NA as well
    mutate(
      D.Unit.Amount   = ifelse(is.na(D.Amount), NA_character_, D.Unit.Amount),
      D.Unit.Time     = ifelse(is.na(D.Amount), NA_character_, D.Unit.Time),
      D.Unit.Animals  = ifelse(is.na(D.Amount), NA_character_, D.Unit.Animals)
    )
}

# Apply the harmonization
ingredients <- harmonize_units(ingredients)

# Load replicate info to adjust for "all animals in replicate"
reps <- livestock_metadata$MT.Out %>%
  mutate(T.Animals = as.numeric(T.Animals)) %>%
  distinct(B.Code, A.Level.Name, .keep_all = TRUE)

ingredients <- ingredients %>%
  left_join(reps %>% dplyr::select(B.Code, A.Level.Name, T.Animals), by=c("B.Code","A.Level.Name")) %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Animals == "all animals in replicate" & !is.na(T.Animals) ~ D.Amount / T.Animals,
      TRUE ~ D.Amount
    ),
    D.Unit.Animals = ifelse(D.Unit.Animals == "all animals in replicate", "individual", D.Unit.Animals)
  )

# Refresh 'Units' column after these adjustments (optional)
ingredients <- ingredients %>%
  mutate(Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"))


```


```{r,message=FALSE,warning=FALSE}


# Define a list of acceptable units
acceptable_units <- c("g/kg","g","%","ml","L","kg","g/L",
                      "g/kg metabolic weight (0.75)","% Diet","kg/kg metabolic weight",
                      "g/kg body weight","ml/kg","% Body Mass","% Concentrate",
                      "kg/100kg body weight")

# Filter rows to keep only acceptable units
ingredients_harmonized <- ingredients %>%
  filter(D.Unit.Amount %in% acceptable_units | is.na(D.Unit.Amount)) %>%
  mutate(D.Item = ifelse(is.na(D.Item), "Unspecified", D.Item))

#tracking paper loss because of harmnozied units
total_studies_harmonized<- length(unique(ingredients_harmonized$B.Code))
total_ingredients_harmonized<- length(unique(ingredients_harmonized$D.Item))
total_diets_harmonized<- length(unique(ingredients_harmonized$A.Level.Name))


```

## 3) Link ingredients to feedipedia 

Feedipedia is an online database that provides comprehensive information on feed ingredients used in livestock nutrition. It aggregates data on the nutritional composition, digestibility, and other relevant properties of various feeds.We will use it to infer data when info is missing from our data extraction. 

```{r,message=FALSE,warning=FALSE}

## download data from feedapedia (generated with the "feedapedia scrapping script available on github)

# Use the raw URL of the file on GitHub
url <- "https://github.com/ERAgriculture/ERL/raw/main/downloaded_data/feedipedia.parquet"

# Create a temporary file location with a .parquet extension
temp_file <- tempfile(fileext = ".parquet")

# Download the file in binary mode
download.file(url, destfile = temp_file, mode = "wb")

# Read the Parquet file
feedipedia <- read_parquet(temp_file)

## download AOM with the correspondance between feedipedia codes and diet items 

# Define the raw URL for the Excel file
url <- "https://raw.githubusercontent.com/peetmate/era_codes/main/era_master_sheet.xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, destfile = temp_file, mode = "wb")

# Read only the sheet called "AOM"
mastersheet <- read_excel(temp_file, sheet = "AOM") %>%
  filter(L5=="Feed Ingredient")

mastersheet$Feedipedia <- as.numeric(sub(".*node/", "", mastersheet$Feedipedia))


```


```{r,message=FALSE,warning=FALSE}

# Merge the mastersheet feedipedia node numbers into the ingredients data
ingredients <- ingredients_harmonized %>%
  left_join(
    mastersheet %>% select(Edge_Value, Feedipedia),
    by = c("D.Item" = "Edge_Value")
  )%>%
  distinct()

#proportion of ingredients without feedapedia codes 
missing_feedapedia <- ingredients %>%
  filter(is.na(Feedipedia))%>%
  pull(D.Item) %>%
  unique() %>%
  length()

proportion_missing_feedipedia<-missing_feedapedia/length(unique(ingredients$D.Item))

```


## 4) Integrate basal diets 

Basal diets serve as the control and foundation for all treatments in livestock experiments. They are present in both the control diet and in every treatment diet that includes additional ingredients. This section integrates basal diet ingredients into the treatment diets while preserving the original basal formulation, and it flags any ingredient that originates from the basal diet.

```{r,message=FALSE,warning=FALSE}
# Identify basal diets
basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) == "base") %>%
  mutate(from_basal = TRUE)

# Identify non-basal diets
non_basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) != "base") %>%
  mutate(from_basal = FALSE)

# Duplicate basal diets for each non-basal diet within the same B.Code
basal_expanded <- non_basal_diets %>%
  dplyr::select(B.Code, A.Level.Name) %>%
  distinct() %>%
  inner_join(basal_diets, by="B.Code") %>%
  mutate(
    A.Level.Name = A.Level.Name.x, # carry over the non-basal A.Level.Name
    from_basal   = TRUE
  ) %>%
  dplyr::select(-A.Level.Name.x, -A.Level.Name.y)

# Combine everything into one integrated dataset
ingredients <- bind_rows(
  non_basal_diets,   # original non-basal
  basal_expanded,    # basal items added to each non-basal diet
  basal_diets        # keep the purely basal diet for control
)



```


## 5) Infer ingredient amount from feed intake data 

Occasionally, the ingredient amounts may be missing from the ingredient table or simply noted as ad libitum feeding. However, since the actual intake is sometimes recorded in the outcome data, we can extract the feed intake value and infer it as the amount of the ingredient administered to the animal.

```{r,message=FALSE,warning=FALSE}

#number of initial ingredient with diet amount 
ingredient_amount_initial <- ingredients %>%
  filter(!is.na(D.Amount)) %>%
  nrow()


# Extract feed intake data
feed_intake <- livestock_metadata$Data.Out %>%
  select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw, ED.Mean.T, ED.Error)

# Filter feed intake data for ingredients only (not entire diet)
feed_intake_ingredients <- feed_intake %>%
  filter(ED.Intake.Item != A.Level.Name)

# Merge to fill missing D.Amount values
ingredients <- ingredients %>%
  left_join(feed_intake_ingredients %>% 
              select(B.Code, A.Level.Name, ED.Intake.Item, ED.Mean.T), 
            by = c("B.Code", "A.Level.Name", "D.Item" = "ED.Intake.Item")) %>%
  mutate(
    # Flag if the value is being inferred from feed intake
    from_feed_intake = ifelse(is.na(D.Amount) & !is.na(ED.Mean.T), TRUE, FALSE),

    # If D.Amount is missing, fill with ED.Mean.T
    D.Amount = ifelse(is.na(D.Amount), ED.Mean.T, D.Amount)
  ) %>%
  select(-ED.Mean.T) %>%  # Remove ED.Mean.T column after use
  distinct()  # Remove duplicates if any

ingredient_amount_after_feed_intake <- ingredients %>%
  filter(!is.na(D.Amount)) %>%
  nrow()

```
whats the unit of feed intake ??

## 6) Table of all diet ingredients amounts within each diet 
```{r,message=FALSE,warning=FALSE}


# This 'ingredients' now contains all diets + basal integrated.
# Table 1: Final diet amounts with harmonized units (including basal)
datatable(
  ingredients, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Diet Ingredients"
)

```

## 7) Relative % of each ingredient in the diet

This table presents the relative percentage of each ingredient within a diet. To ensure accurate proportion calculations, we first filtered the dataset to include only diets with consistent units. Diets with incompatible or non-harmonizable units were excluded, as their proportions could not be reliably determined.

```{r,message=FALSE,warning=FALSE}
# We now have a single integrated dataset: 'ingredients'.
# 1) Filter to diets that have a single unit type per group (consistent units).
# 2) Compute total amounts and percentages.

inconsistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  filter(Unit_Count > 1) %>%
  ungroup()

# Step 1: Flag only consistent-unit diets
consistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  # Count only the non-NA units
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  # Keep groups with 0 or 1 distinct unit (ignoring NA)
  filter(Unit_Count <= 1) %>%
  ungroup()

#porportion of inconsistent diets 
proportion_inconsistent_diets<- length(unique(inconsistent_diets$A.Level.Name))/length(unique(ingredients$A.Level.Name))

# Step 2: Sum total amounts per (B.Code, A.Level.Name, D.Unit.Amount)
# Transform consistent_diets into a summary of totals, removing any NA in amount or unit
total_amounts <- consistent_diets %>%
  filter(!is.na(D.Amount), !is.na(D.Unit.Amount)) %>%
  group_by(B.Code, A.Level.Name, D.Unit.Amount) %>%
  summarise(Total_Amount = sum(D.Amount, na.rm = TRUE), .groups = "drop")
# Display the resulting table


# Create an interactive table for "total_amounts"
datatable(
  total_amounts, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)



# Step 3: Calculate percentages
diet_percentages <- consistent_diets %>%
  left_join(total_amounts, by=c("B.Code","A.Level.Name","D.Unit.Amount")) %>%
  mutate(
    Percentage_of_Diet = ifelse(Total_Amount > 0,
                                (D.Amount / Total_Amount)*100,
                                NA_real_)
  ) %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    Percentage_of_Diet,
    from_basal
  )

# Table 2: Relative percentage of each ingredient in each diet


# Create an interactive table for "total_amounts"
datatable(
  diet_percentages, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)






```

```{r,message=FALSE,warning=FALSE}
# Check if each diet sums close to 100%
validation <- diet_percentages %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(Total_Percentage = sum(Percentage_of_Diet, na.rm=TRUE)) %>%
  ungroup()

# Display diets and their total percentages
datatable(
  validation, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Validation table"
)


```

The lines that have 0 as validation are the ones that were feeding ad libitum for which we do not have a precise amount so it's impossible to calculate the relative proportion of the ingredients.

## 8) Table with nutrition and digestibility for the entire diet

Within the nutrition and digestibility tables, the "is entire diet" column distinguishes the scope of the data. A value of FALSE indicates that the nutrition or digestibility information pertains solely to the specific diet ingredient, whereas a value of TRUE signifies that the information represents the nutritional or digestibility profile of the entire diet. 

## 8.1) Add nutrition and digestibility info at the ingredient level 


```{r,message=FALSE,warning=FALSE}

#### 1) Prepare Digestibility data
Digestibility_ingredient <- livestock_metadata$Animals.Diet.Digest %>%
  filter(is_entire_diet=="FALSE")

#### 2) Prepare Nutrition data
Nutrition_ingredient <- livestock_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet=="FALSE")

#### 3) Join to ingredients table

ingredients_dd_dc <- ingredients %>%
  left_join(Digestibility_ingredient, 
            by = c("B.Code", "D.Item"),
            suffix = c("", "_digest")) %>%
  left_join(Nutrition_ingredient, 
            by = c("B.Code", "D.Item"),
            suffix = c("", "_nutri"))


```


## 8.2) Add nutrition and digestibility info at the diet level 

```{r,message=FALSE,warning=FALSE}

#### 1) Prepare Digestibility data
Digestibility_diet <- livestock_metadata$Animals.Diet.Digest %>%
  filter(is_entire_diet=="TRUE")%>%
  rename(A.Level.Name = D.Item) 

#### 2) Prepare Nutrition data
Nutrition_diet <- livestock_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet=="TRUE")%>%
  rename(A.Level.Name = D.Item) 

#### 3) Join to total_amounts

complete_diet_dd_dc <- total_amounts %>%
  left_join(Digestibility_diet, 
            by = c("B.Code", "A.Level.Name"),
            suffix = c("", "_digest")) %>%
  left_join(Nutrition_diet, 
            by = c("B.Code", "A.Level.Name"),
            suffix = c("", "_nutri"))


```


## 9) Add nutrition and digestibility information from feedipedia parquet at the ingredient level

```{r,message=FALSE,warning=FALSE}
### --- 1. Categorize Feedipedia Variables ----------------------------------

digestibility_keywords <- c(
  "digestibility", "degradability", "energy digestibility", 
  "DE", "ME", "NE", "AMEn", "OM digestibility", "DM digestibility",
  "ruminants", "rabbit", "pig", "poultry", "horse", "salmonids",
  "hydrolysis"
)

# Assign category based on the presence of digestibility-related keywords
feedipedia <- feedipedia %>%
  mutate(category = if_else(
    str_detect(Variable, regex(paste(digestibility_keywords, collapse = "|"), ignore_case = TRUE)), 
    "digestibility", 
    "nutrition"
  ))

### --- 2. Prepare Long-Format Feedipedia Data for Each Ingredient ------------

# Convert feedipedia_node to numeric and select needed columns
feedipedia_long <- feedipedia %>%
  mutate(feedipedia_node = as.numeric(feedipedia_node)) %>%
  select(feedipedia_node, Variable, Unit, Avg, category)

# Join the feedipedia data to the ingredients data by matching Feedipedia node number
ingredients_feed <- ingredients_dd_dc %>%
  left_join(feedipedia_long, by = c("Feedipedia" = "feedipedia_node"))

# For each ingredient (grouped by B.Code and Feedipedia), separate nutrition and digestibility rows,
# and add a row number for alignment.
nutr_long <- ingredients_feed %>%
  filter(category == "nutrition") %>%
  group_by(B.Code, Feedipedia) %>%
  mutate(DC.row = row_number()) %>%
  ungroup() %>%
  select(B.Code, Feedipedia, DC.row,
         DC.Variable = Variable, DC.Value = Avg, DC.Unit = Unit)

dig_long <- ingredients_feed %>%
  filter(category == "digestibility") %>%
  group_by(B.Code, Feedipedia) %>%
  mutate(DD.row = row_number()) %>%
  ungroup() %>%
  select(B.Code, Feedipedia, DD.row,
         DD.Variable = Variable, DD.Value = Avg, DD.Unit = Unit)

# Combine the nutrition and digestibility subsets by matching B.Code, Feedipedia, and the row numbers.
combined_feed <- full_join(nutr_long, dig_long, 
                           by = c("B.Code", "Feedipedia", "DC.row" = "DD.row")) %>%
  select(-DC.row)  # remove the temporary row number

### --- 3. Merge Combined Feedipedia Data Back with Ingredients and Pivot to Long Format ----

# Merge the combined feed information back with the original ingredients data.
# Merge Feedipedia data into the existing ingredients dataset
ingredients_dd_dc_feed <- ingredients_dd_dc %>%
  left_join(combined_feed, by = c("B.Code", "Feedipedia")) %>%
  mutate(
    DC.Value = coalesce(DC.Value.x, DC.Value.y),
    DC.Unit = coalesce(DC.Unit.x, DC.Unit.y),
    DD.Value = coalesce(DD.Value.x, DD.Value.y),
    DD.Unit = coalesce(DD.Unit.x, DD.Unit.y),

    # Add Variables explicitly
    DC.Variable = coalesce(DC.Variable.x, DC.Variable.y),
    DD.Variable = coalesce(DD.Variable.x, DD.Variable.y),

    nutrition_source = case_when(
      !is.na(DC.Value.x) ~ "extraction",
      !is.na(DC.Value.y) ~ "feedipedia",
      TRUE ~ NA_character_
    ),
    digestibility_source = case_when(
      !is.na(DD.Value.x) ~ "extraction",
      !is.na(DD.Value.y) ~ "feedipedia",
      TRUE ~ NA_character_
    )
  ) %>%
  select(-ends_with(".x"), -ends_with(".y"))



```

## 10) infer nutrition from our data 

```{r,message=FALSE,warning=FALSE}

# -----------------------------
# Impute missing nutrition values
# -----------------------------
nutrition_imputed <- ingredients_dd_dc_feed %>%
  # Consider only rows with non-missing nutrition values (extracted from our data source)
   filter(!is.na(DC.Value))%>%
  filter(nutrition_source=="extraction")%>%
  group_by(D.Item,DC.Variable, DC.Unit) %>%  # Group by ingredient and its nutrition unit
  summarise(mean_nutrition = mean(DC.Value, na.rm = TRUE))



# Merge back to the ingredients data and impute missing nutrition values
ingredients_dd_dc_feed <- ingredients_dd_dc_feed %>%
  left_join(nutrition_imputed, by = c("D.Item","DC.Variable","DC.Unit")) %>%
  mutate(
    # If DC_Value is missing, fill it with the computed mean
    DC.Value = if_else(is.na(DC.Value), mean_nutrition, DC.Value),
    # Flag the source as "extraction" if we imputed the value
    nutrition_source = if_else(is.na(nutrition_source) & is.na(DC.Value) == FALSE, "extraction estimate", nutrition_source)
  ) %>%
  select(-mean_nutrition)

# -----------------------------
# Impute missing digestibility values
# -----------------------------
digest_imputed <- ingredients_dd_dc_feed %>%
  filter(!is.na(DD.Value)) %>%
    filter(digestibility_source=="extraction")%>%

  group_by(D.Item,DD.Variable, DD.Unit) %>%  # Group by ingredient and its digestibility unit
  summarise(mean_digest = mean(DD.Value, na.rm = TRUE))

# Merge back to the ingredients data and impute missing digestibility values
ingredients_dd_dc_feed_complete <- ingredients_dd_dc_feed %>%
  left_join(digest_imputed, by = c("D.Item","DD.Variable","DD.Unit")) %>%
  mutate(
    DD.Value = if_else(is.na(DD.Value), mean_digest, DD.Value),
    digestibility_source = if_else(is.na(digestibility_source) & is.na(DD.Value) == FALSE, "extraction estimate", digestibility_source)
  ) %>%
  select(-mean_digest)



```



## 11) Tracking data table 
```{r,message=FALSE,warning=FALSE}


# 2. Percentage of studies with at least one ad libitum feeding
b_codes_with_ad_lib <- unique(ingredients$B.Code[ingredients$D.Ad.lib == TRUE])
# Calculate percentage
ad_lib_percent <- (length(b_codes_with_ad_lib) / n_distinct(ingredients$B.Code)) * 100

# 3. Count the number of unique ingredients
total_ingredients <- n_distinct(ingredients_dd_dc$D.Item)

# 4. Calculate % of ingredients with nutrition and digestibility info
ingredients_with_info <- ingredients_dd_dc %>%
  group_by(D.Item) %>%
  summarise(
    has_nutrition = any(!is.na(DC.Value)),
    has_digestibility = any(!is.na(DD.Value))
  ) %>%
  mutate(has_any_info = has_nutrition | has_digestibility)

perc_nutrition_dd_dc <- (sum(ingredients_with_info$has_nutrition) / total_ingredients) * 100
perc_digestibility_dd_dc <- (sum(ingredients_with_info$has_digestibility) / total_ingredients) * 100
perc_both_dd_dc <- (sum(ingredients_with_info$has_any_info) / total_ingredients) * 100

# 5. After Feedipedia Scraping
ingredients_with_info_feed <- ingredients_dd_dc_feed %>%
  group_by(D.Item) %>%
  summarise(
    has_nutrition = any(!is.na(DC.Value)),
    has_digestibility = any(!is.na(DD.Value))
  ) %>%
  mutate(has_any_info = has_nutrition | has_digestibility)

perc_nutrition_dd_dc_feed <- (sum(ingredients_with_info_feed$has_nutrition) / total_ingredients) * 100
perc_digestibility_dd_dc_feed <- (sum(ingredients_with_info_feed$has_digestibility) / total_ingredients) * 100
perc_both_dd_dc_feed <- (sum(ingredients_with_info_feed$has_any_info) / total_ingredients) * 100

# 6. After Inferring from Own Data
ingredients_with_info_complete <- ingredients_dd_dc_feed_complete %>%
  group_by(D.Item) %>%
  summarise(
    has_nutrition = any(!is.na(DC.Value)),
    has_digestibility = any(!is.na(DD.Value))
  ) %>%
  mutate(has_any_info = has_nutrition | has_digestibility)

perc_nutrition_dd_dc_feed_complete <- (sum(ingredients_with_info_complete$has_nutrition) / total_ingredients) * 100
perc_digestibility_dd_dc_feed_complete <- (sum(ingredients_with_info_complete$has_digestibility) / total_ingredients) * 100
perc_both_dd_dc_feed_complete <- (sum(ingredients_with_info_complete$has_any_info) / total_ingredients) * 100

# 7. At the Diet Level
total_diets <- n_distinct(complete_diet_dd_dc$A.Level.Name)

diets_with_info <- complete_diet_dd_dc %>%
  group_by(A.Level.Name) %>%
  summarise(
    has_nutrition = any(!is.na(DC.Value)),
    has_digestibility = any(!is.na(DD.Value))
  ) %>%
  mutate(has_any_info = has_nutrition | has_digestibility)

perc_nutrition_dd_dc_diet <- (sum(diets_with_info$has_nutrition) / total_diets) * 100
perc_digestibility_dd_dc_diet <- (sum(diets_with_info$has_digestibility) / total_diets) * 100
perc_both_dd_dc_diet <- (sum(diets_with_info$has_any_info) / total_diets) * 100

# 8. Create and Display Tracking Table
tracking_table <- data.frame(
  Step = c(
    "Total number of unique studies (B.Code) in the initial dataset",
    "Total number of unique studies after harmonization",
    "Proportion of ingredients missing Feedipedia information",
    "Proportion of diets with inconsistent units",
    "Percentage of studies with at least one ad libitum diet"
  ),
  Value = round(c(
    total_studies,
    total_studies_harmonized,
    proportion_missing_feedipedia,
    proportion_inconsistent_diets,
    ad_lib_percent
  ), 2) # Round to 2 decimals for readability
)

# Display the tracking table interactively
datatable(
  tracking_table, 
  options = list(
    pageLength = 10,         
    scrollX = TRUE,          
    autoWidth = TRUE,        
    searchHighlight = TRUE   
  ),
  caption = "Tracking Table: Data Processing and Missing Info Percentage"
)

tracking_table_nutrition_digestibility <- data.frame(
  Category = c("Nutrition", "Digestibility"),
  Initial = c(100 - perc_nutrition_dd_dc, 100 - perc_digestibility_dd_dc),
  After_Feedipedia_Scraping = c(100 - perc_nutrition_dd_dc_feed, 100 - perc_digestibility_dd_dc_feed),
  After_Inference_from_Data = c(100 - perc_nutrition_dd_dc_feed_complete, 100 - perc_digestibility_dd_dc_feed_complete)
)

# Display the second table
datatable(
  tracking_table_nutrition_digestibility, 
  options = list(pageLength = 2, scrollX = TRUE, autoWidth = TRUE, searchHighlight = TRUE),
  caption = "Tracking Table: Nutrition and Digestibility Improvements"
)

```

#12) Number for emissions calculations on new extractions

```{r}
#total number of livestock papers 
length(unique(livestock_metadata$Pub.Out$B.Code))


# total number of cattle and sheeps and goats 
prod_out <- livestock_metadata[["Prod.Out"]]

# Filter and count unique b.code for Sheep, Goat, or Cattle
library(dplyr)

prod_out %>%
  filter(P.Product %in% c("Sheep", "Goat", "Cattle")) %>%
  distinct(B.Code) %>%
  nrow()


#gross energy papers


# Step 1: Get b.codes for Sheep, Goat, or Cattle
target_bcodes <- livestock_metadata[["Prod.Out"]] %>%
  filter(P.Product %in% c("Sheep", "Goat", "Cattle")) %>%
  pull(B.Code) %>%
  unique()

# Step 2: Filter other_table for those b.codes and variable == "GE", then count unique b.codes
livestock_metadata$Animals.Diet.Comp %>%
  filter(B.Code %in% target_bcodes, DC.Variable == "GE") %>%
  distinct(B.Code) %>%
  nrow()


```





#13 Number for emission calculations in courageous camel 

```{r}


# Set up connection to S3 bucket
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("courageous_camel_2024_draft_2025", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
courageous_camel <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))

```



