---
title: "Livestock diet items"
author:
  - name: "Peter Steward"
    email: "p.steward@cgiar.org"
    orcid: "0000-0003-3985-4911"
  - name: "Namita Joshi"
    email: "n.joshi@cgiar.org"
    orcid: "0000-0002-1623-6910"
  - name: "Lolita Muller"
    email: "m.lolita@cgiar.org"
    orcid: "0009-0002-8345-7219"
  - name: "Todd Rosenstock"
    email: "t.rosenstock@cgiar.org"
    orcid: "0000-0002-1958-9500"
date: "2024-12-15"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide

---

## Introduction

### Background
Understanding livestock diets is crucial for improving animal nutrition, optimizing feed efficiency, and assessing environmental impacts. However, raw diet composition data is often messy, with inconsistent units and missing harmonized nutritional information. Standardizing and integrating these datasets enables more accurate comparisons and facilitates meta-analysis across studies.

### Objective
This analysis aims to:
1. **Download and process livestock diet metadata** from an online repository.
2. **Harmonize dietary ingredient units** 
3. **Integrate basal diets** into experimental treatments.
4. **Calculate the relative proportion of each ingredient** within each diet.
5. **Generate a final table** with summarized nutritional and digestibility information for complete diets.
6. **Infer the data** from feedapedia and when missing infer from current data for nutrition and digestibility.

By automating this workflow, we streamline livestock nutrition research and ensure consistent data preparation for further analysis.


## 1) Set up 

```{r,message=FALSE,warning=FALSE}
# Install and load 'pacman' package manager if not already installed

library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)


# Set up connection to S3 bucket
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("skinny_cow_2022", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
livestock_metadata <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))



```



## 2) Harmonize units 

```{r,message=FALSE,warning=FALSE}
# diet ingredients: breakdown of all ingredients within each diet
ingredients <- livestock_metadata$Animals.Diet %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item.AOM,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    D.Unit.Time,
    D.Unit.Animals,
    DC.Is.Dry,
    D.Ad.lib,
    D.Is.Group,
    D.Item.Root.Other.Comp.Proc_All
  ) %>%
  mutate(D.Amount = as.numeric(D.Amount))

# Function to harmonize diet ingredient units
harmonize_units <- function(df) {
  df %>%
    mutate(
      # Record original units (optional, for debugging/traceability)
      Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"),

      # Harmonize weight and volume units to a consistent base
      D.Amount = case_when(
        D.Unit.Amount %in% c("kg", "mg", "g/100g", "l", "kg/t", "g/300l", 
                             "mg/kg", "kg/100kg", "kg/kg metabolic weight", 
                             "kg/100kg body weight", "g/kg metabolic weight (0.75)", 
                             "g/L") ~ {
          case_when(
            D.Unit.Amount == "kg" ~ D.Amount * 1000,                      # kg -> g
            D.Unit.Amount == "mg" ~ D.Amount / 1000,                     # mg -> g
            D.Unit.Amount == "g/100g" ~ D.Amount * 10,                   # g/100g -> g/kg
            D.Unit.Amount == "l" ~ D.Amount * 1000,                      # l -> ml (assumed g ~ ml in water)
            D.Unit.Amount == "kg/t" ~ D.Amount * 1000,                   # kg/t -> g/kg
            D.Unit.Amount == "g/300l" ~ D.Amount / 300,                  # g/300l -> g/L
            D.Unit.Amount == "mg/kg" ~ D.Amount / 1000,                  # mg/kg -> g/kg
            D.Unit.Amount == "kg/100kg body weight" ~ D.Amount * 10,     # -> g/kg body weight
            D.Unit.Amount == "kg/kg metabolic weight" ~ D.Amount * 1000, # -> g/kg metabolic weight
            TRUE ~ D.Amount
          )
        },
        TRUE ~ D.Amount
      ),

      # Update D.Unit.Amount to reflect new base
      D.Unit.Amount = case_when(
        D.Unit.Amount == "kg"  ~ "g",
        D.Unit.Amount == "mg"  ~ "g",
        D.Unit.Amount == "g/100g" ~ "g/kg",
        D.Unit.Amount == "l"  ~ "ml",
        D.Unit.Amount == "kg/t" ~ "g/kg",
        D.Unit.Amount == "g/300l" ~ "g/L",
        D.Unit.Amount == "mg/kg" ~ "g/kg",
        D.Unit.Amount == "kg/100kg body weight" ~ "g/kg body weight",
        D.Unit.Amount == "kg/kg metabolic weight" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg metabolic weight (0.75)" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/L" ~ "g/L",
        TRUE ~ D.Unit.Amount
      ),

      # Harmonize time units to day
      D.Amount = case_when(
        D.Unit.Time == "week" ~ D.Amount / 7,
        D.Unit.Time == "month" ~ D.Amount / 30,
        D.Unit.Time == "2 weeks" ~ D.Amount / 14,
        D.Unit.Time == "4 days interval" ~ D.Amount / 4,
        D.Unit.Time == "2x/day" ~ D.Amount * 2,
        TRUE ~ D.Amount
      ),
       D.Unit.Time = ifelse(!is.na(D.Unit.Time), "day", D.Unit.Time)
    ) %>%
    # NEW: If D.Amount is NA, set unit fields to NA as well
    mutate(
      D.Unit.Amount   = ifelse(is.na(D.Amount), NA_character_, D.Unit.Amount),
      D.Unit.Time     = ifelse(is.na(D.Amount), NA_character_, D.Unit.Time),
      D.Unit.Animals  = ifelse(is.na(D.Amount), NA_character_, D.Unit.Animals)
    )
}

# Apply the harmonization
ingredients <- harmonize_units(ingredients)

# Load replicate info to adjust for "all animals in replicate"
reps <- livestock_metadata$MT.Out %>%
  mutate(T.Animals = as.numeric(T.Animals)) %>%
  distinct(B.Code, A.Level.Name, .keep_all = TRUE)

ingredients <- ingredients %>%
  left_join(reps %>% dplyr::select(B.Code, A.Level.Name, T.Animals), by=c("B.Code","A.Level.Name")) %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Animals == "all animals in replicate" & !is.na(T.Animals) ~ D.Amount / T.Animals,
      TRUE ~ D.Amount
    ),
    D.Unit.Animals = ifelse(D.Unit.Animals == "all animals in replicate", "individual", D.Unit.Animals)
  )

# Refresh 'Units' column after these adjustments (optional)
ingredients <- ingredients %>%
  mutate(Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"))


```


```{r,message=FALSE,warning=FALSE}


# Define a list of acceptable units
acceptable_units <- c("g/kg","g","%","ml","L","kg","g/L",
                      "g/kg metabolic weight (0.75)","% Diet","kg/kg metabolic weight",
                      "g/kg body weight","ml/kg","% Body Mass","% Concentrate",
                      "kg/100kg body weight")

# Filter rows to keep only acceptable units
ingredients <- ingredients %>%
  filter(D.Unit.Amount %in% acceptable_units | is.na(D.Unit.Amount)) %>%
  mutate(D.Item = ifelse(is.na(D.Item), "Unspecified", D.Item))


```


## 3) Integrate basal diets 

```{r,message=FALSE,warning=FALSE}
# Identify basal diets
basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) == "base") %>%
  mutate(from_basal = TRUE)

# Identify non-basal diets
non_basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) != "base") %>%
  mutate(from_basal = FALSE)

# Duplicate basal diets for each non-basal diet within the same B.Code
basal_expanded <- non_basal_diets %>%
  dplyr::select(B.Code, A.Level.Name) %>%
  distinct() %>%
  inner_join(basal_diets, by="B.Code") %>%
  mutate(
    A.Level.Name = A.Level.Name.x, # carry over the non-basal A.Level.Name
    from_basal   = TRUE
  ) %>%
  dplyr::select(-A.Level.Name.x, -A.Level.Name.y)

# Combine everything into one integrated dataset
ingredients <- bind_rows(
  non_basal_diets,   # original non-basal
  basal_expanded,    # basal items added to each non-basal diet
  basal_diets        # keep the purely basal diet for control
)

# This 'ingredients' now contains all diets + basal integrated.
# Table 1: Final diet amounts with harmonized units (including basal)
datatable(
  ingredients, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Diet Ingredients"
)

```

## 4) Relative % of each ingredient in the diet

```{r,message=FALSE,warning=FALSE}
# We now have a single integrated dataset: 'ingredients'.
# 1) Filter to diets that have a single unit type per group (consistent units).
# 2) Compute total amounts and percentages.

inconsistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(Unit_Count = n_distinct(D.Unit.Amount)) %>%
  filter(Unit_Count > 1) %>%
  ungroup()

# Step 1: Flag only consistent-unit diets
consistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  # Count only the non-NA units
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  # Keep groups with 0 or 1 distinct unit (ignoring NA)
  filter(Unit_Count <= 1) %>%
  ungroup()



# Step 2: Sum total amounts per (B.Code, A.Level.Name, D.Unit.Amount)
# Transform consistent_diets into a summary of totals, removing any NA in amount or unit
total_amounts <- consistent_diets %>%
  filter(!is.na(D.Amount), !is.na(D.Unit.Amount)) %>%
  group_by(B.Code, A.Level.Name, D.Unit.Amount) %>%
  summarise(Total_Amount = sum(D.Amount, na.rm = TRUE), .groups = "drop")
# Display the resulting table


# Create an interactive table for "total_amounts"
datatable(
  total_amounts, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)



# Step 3: Calculate percentages
diet_percentages <- consistent_diets %>%
  left_join(total_amounts, by=c("B.Code","A.Level.Name","D.Unit.Amount")) %>%
  mutate(
    Percentage_of_Diet = ifelse(Total_Amount > 0,
                                (D.Amount / Total_Amount)*100,
                                NA_real_)
  ) %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    Percentage_of_Diet,
    from_basal
  )

# Table 2: Relative percentage of each ingredient in each diet


# Create an interactive table for "total_amounts"
datatable(
  diet_percentages, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)






```

```{r,message=FALSE,warning=FALSE}
# Check if each diet sums close to 100%
validation <- diet_percentages %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(Total_Percentage = sum(Percentage_of_Diet, na.rm=TRUE)) %>%
  ungroup()

# Display diets and their total percentages
datatable(
  validation, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Validation table"
)


```


## 5) Table with nutrition and digestibility for the entire diet

```{r,message=FALSE,warning=FALSE}

#### 1) Prepare Digestibility data
Digestibility <- livestock_metadata$Animals.Diet.Digest %>%
  rename(A.Level.Name = D.Item) 

#### 2) Prepare Nutrition data
Nutrition <- livestock_metadata$Animals.Diet.Comp %>%
  rename(A.Level.Name = D.Item) 

#### 3) Join to total_amounts

complete_diet <- total_amounts %>%
  left_join(Digestibility, 
            by = c("B.Code", "A.Level.Name"),
            suffix = c("", "_digest")) %>%
  left_join(Nutrition, 
            by = c("B.Code", "A.Level.Name"),
            suffix = c("", "_nutri"))



datatable(
  complete_diet, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Nutrition and digestibility information at the diet levels"
)

```
- porblem i sentire diet = false in nutrition no way to know from which ingredient it is coming from because only diet column in the nutrition table 

6) Tracking paper loss and ad libitum data 
```{r,message=FALSE,warning=FALSE}


# Initialize the tracking table
tracking_table <- data.frame(
  Step = character(),
  Unique_BCodes = integer(),
  Lost_BCodes = integer(),
  Additional_Info = character(),
  stringsAsFactors = FALSE
)

# Step 1: Initial unique B.Codes count
initial_count <- length(unique(livestock_metadata$Animals.Diet$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "Initial Dataset", 
                                                   Unique_BCodes = initial_count, 
                                                   Lost_BCodes = 0,
                                                   Additional_Info = "Raw dataset before processing"))

# Step 2: After harmonization
post_harmonization <- length(unique(ingredients$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "After Harmonization", 
                                                   Unique_BCodes = post_harmonization, 
                                                   Lost_BCodes = initial_count - post_harmonization,
                                                   Additional_Info = "Standardized units across dataset"))

# Step 3: After filtering acceptable units
post_units_filter <- length(unique(ingredients$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "After Filtering Acceptable Units", 
                                                   Unique_BCodes = post_units_filter, 
                                                   Lost_BCodes = post_harmonization - post_units_filter,
                                                   Additional_Info = "Removed studies with inconsistent units"))

# Step 4: After integrating basal diets
post_basal_integration <- length(unique(ingredients$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "After Integrating Basal Diets", 
                                                   Unique_BCodes = post_basal_integration, 
                                                   Lost_BCodes = post_units_filter - post_basal_integration,
                                                   Additional_Info = "Integrated basal diets into experimental diets"))

# Step 5: After filtering for consistent units
post_consistent_units <- length(unique(consistent_diets$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "After Filtering for Consistent Units", 
                                                   Unique_BCodes = post_consistent_units, 
                                                   Lost_BCodes = post_basal_integration - post_consistent_units,
                                                   Additional_Info = "Removed diets with multiple inconsistent units"))

# Step 6: After removing NA amounts for percentage calculation
post_percentage_calc <- length(unique(total_amounts$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "After Removing NA Amounts for Percentage Calculation", 
                                                   Unique_BCodes = post_percentage_calc, 
                                                   Lost_BCodes = post_consistent_units - post_percentage_calc,
                                                   Additional_Info = "Filtered out missing diet proportions"))

# Step 7: Number of studies with ad libitum diets
num_b_codes_with_ad_libitum <- length(unique(b_codes_with_ad_libitum$B.Code))
tracking_table <- rbind(tracking_table, data.frame(Step = "Studies with Ad Libitum Diets", 
                                                   Unique_BCodes = num_b_codes_with_ad_libitum, 
                                                   Lost_BCodes = NA,
                                                   Additional_Info = "Studies where at least one ingredient is ad libitum"))

# Step 8: Missing digestibility/nutrition information
total_diets <- n_distinct(complete_diet$A.Level.Name)
prop_no_digestibility <- mean(diet_status$no_digestibility) * 100
prop_no_nutrition <- mean(diet_status$no_nutrition) * 100
prop_no_info <- mean(diet_status$no_info) * 100

tracking_table <- rbind(tracking_table, data.frame(Step = "Diets with No Digestibility Info", 
                                                   Unique_BCodes = NA, 
                                                   Lost_BCodes = NA,
                                                   Additional_Info = paste0(round(prop_no_digestibility, 1), "% of diets")))
tracking_table <- rbind(tracking_table, data.frame(Step = "Diets with No Nutrition Info", 
                                                   Unique_BCodes = NA, 
                                                   Lost_BCodes = NA,
                                                   Additional_Info = paste0(round(prop_no_nutrition, 1), "% of diets")))
tracking_table <- rbind(tracking_table, data.frame(Step = "Diets with No Digestibility or Nutrition Info", 
                                                   Unique_BCodes = NA, 
                                                   Lost_BCodes = NA,
                                                   Additional_Info = paste0(round(prop_no_info, 1), "% of diets")))

# Display the tracking table interactively
datatable(
  tracking_table, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Comprehensive Tracking Table: Study Filtering and Data Loss"
)



```

Add how many papers dont have nutrition and or digestibility info with NAS

## 6) Infer amounts from feed intake data in outcomes 

```{r}


feed_intake<-livestock_metadata$Data.Out%>%
  select(B.Code,A.Level.Name,ED.Intake.Item,ED.Intake.Item.Raw,ED.Mean.T,ED.Error,ED.Product.Comp)
```

