---
title: "Livestock diet items"
author:
  - name: "Lolita Muller"
    email: "m.lolita@cgiar.org"
    orcid: "0009-0002-8345-7219"
  - name: "Peter Steward"
    email: "p.steward@cgiar.org"
    orcid: "0000-0003-3985-4911"
  - name: "Namita Joshi"
    email: "n.joshi@cgiar.org"
    orcid: "0000-0002-1623-6910"
  - name: "Todd Rosenstock"
    email: "t.rosenstock@cgiar.org"
    orcid: "0000-0002-1958-9500"
date: "2025-02-19"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show

---

## Introduction


## 1) Set up 
This section installs and loads required packages, connects to the S3 bucket containing ERA data, and loads both the livestock and camel metadata into memory.

```{r,message=FALSE,warning=FALSE}
# Install and load 'pacman' package manager if not already installed

library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)
library(readxl)
library(tidyr)
library(arrow)
library(ggplot2)
library(data.table)
library(purrr)

# Set up connection to S3 bucket
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("skinny_cow_2022", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
livestock_metadata <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))


```


```{r,message=FALSE,warning=FALSE}
# Set up connection to S3 bucket for camel data 
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("courageous_camel_2024", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
camel <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))

```


```{r,message=FALSE,warning=FALSE}
#merge camel and livestock 


# Define column lists (as character vectors)
diet_cols <- c("B.Code", "A.Level.Name", "D.Item", "D.Type", "D.Amount",
               "D.Unit.Amount", "D.Unit.Time", "D.Unit.Animals", "DC.Is.Dry", 
               "D.Ad.lib", "D.Is.Group")

mt_cols <- c("B.Code", "A.Level.Name", "T.Animals")

comp_cols <- c("B.Code", "D.Item", "is_entire_diet", "DC.Variable", 
               "DC.Value", "DC.Unit")

digest_cols <- c("B.Code", "D.Item", "is_entire_diet", "DD.Variable", 
                 "DD.Value", "DD.Unit")

data_cols <- c("B.Code", "A.Level.Name", "ED.Intake.Item", "ED.Intake.Item.Raw", 
               "ED.Mean.T", "ED.Error","Out.Subind", "Out.Unit", "Out.WG.Start", "Out.WG.Unit")

prod_cols <- c("B.Code", "P.Product")

# Extract livestock tables and add Source column
diet_livestock   <- livestock_metadata$Animals.Diet[, ..diet_cols][, Source := "livestock"]
mt_livestock     <- livestock_metadata$MT.Out[, ..mt_cols][, Source := "livestock"]
comp_livestock   <- livestock_metadata$Animals.Diet.Comp[, ..comp_cols][, Source := "livestock"]
digest_livestock <- livestock_metadata$Animals.Diet.Digest[, ..digest_cols][, Source := "livestock"]
data_livestock   <- livestock_metadata$Data.Out[, ..data_cols][, Source := "livestock"]
prod_livestock   <- livestock_metadata$Prod.Out[, ..prod_cols][, Source := "livestock"]

# Extract camel tables and rename columns to match livestock structure
diet_camel <- camel$Animal.Diet[, ..diet_cols][, Source := "camel"]

# Construct MT.Out for camel: get A.Level.Name and T.Animals from Herd
mt_camel <- camel$Herd.Out[, .(
  B.Code, T.Animals = Herd.N
)][, Source := "camel"]

# Rename DN.* to DC.* in camel Diet.Comp
comp_camel <- camel$Animal.Diet.Comp[, .(
  B.Code, D.Item, is_entire_diet, 
  DC.Variable = DN.Variable,
  DC.Value = DN.Value,
  DC.Unit = DN.Unit
)][, Source := "camel"]

# Rename DD.* and add digestibility source
digest_camel <- camel$Animal.Diet.Digest[, .(
  B.Code, D.Item, is_entire_diet, 
  DD.Variable = DD.Variable,
  DD.Value = DD.Value,
  DD.Unit = DD.Unit
)][, Source := "camel"]

# Join Data.Out from camel with Herd to get weight info
herd_weight <- camel$Herd.Out[, .(
  B.Code, A.Level.Name = Herd.Level.Name, 
  Out.WG.Start = Herd.Start.Weight, 
  Out.WG.Unit = Herd.Start.Weight.Unit
)]

data_camel <- camel$Data.Out[
  , .(B.Code, ED.Intake.Item, 
      ED.Intake.Item.Raw = ED.Intake.Item, 
      A.Level.Name= T.Name,  # Temporarily keep this as-is for now
      ED.Mean.T, ED.Error, Out.Unit,Out.Subind)
][
  herd_weight, 
  on = .(B.Code), 
  `:=`(P.Product = i.A.Level.Name, 
       Out.WG.Start = i.Out.WG.Start, 
       Out.WG.Unit = i.Out.WG.Unit)
][
  , .(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw, 
      ED.Mean.T, ED.Error, Out.Unit,Out.Subind, Out.WG.Start, Out.WG.Unit)
][, Source := "camel"]


# Extract product info from Herd table in camel
prod_camel <- camel$Herd.Out[, .(B.Code, P.Product = V.Product)][, Source := "camel"]

# Merge camel and livestock datasets
diet_all   <- rbindlist(list(diet_livestock, diet_camel), fill = TRUE)
mt_all     <- rbindlist(list(mt_livestock, mt_camel), fill = TRUE)
comp_all   <- rbindlist(list(comp_livestock, comp_camel), fill = TRUE)
digest_all <- rbindlist(list(digest_livestock, digest_camel), fill = TRUE)
data_all   <- rbindlist(list(data_livestock, data_camel), fill = TRUE)
prod_all   <- rbindlist(list(prod_livestock, prod_camel), fill = TRUE)

# Final merged metadata list
merged_metadata <- list(
  Animals.Diet        = diet_all,
  MT.Out              = mt_all,
  Animals.Diet.Comp   = comp_all,
  Animals.Diet.Digest = digest_all,
  Data.Out            = data_all,
  Prod.Out            = prod_all
)

```

```{r,message=FALSE,warning=FALSE}
# Clean up intermediate objects
rm(
  diet_cols, mt_cols, comp_cols, digest_cols, data_cols, prod_cols,
  diet_livestock, mt_livestock, comp_livestock, digest_livestock, data_livestock, prod_livestock,
  diet_camel, mt_camel, comp_camel, digest_camel, data_camel, prod_camel, herd_weight,comp_all,data_all,diet_all,digest_all,mt_all,prod_all,livestock_metadata,camel)

# (Optional) Run gc() to free memory
gc()
```
Some ingredients appear multiple times with amounts reported in different units. In this section, we prioritize entries that are already in preferred units to minimize the need for conversions.

```{r,message=FALSE,warning=FALSE}
## keep ingredient that appear twice with different units only once 

# Define your unit preferences
preferred_units <- c("g", "kg")
preferred_time <- "day"
preferred_animals <- "individual"

# Prioritize and deduplicate
merged_metadata$Animals.Diet <- merged_metadata$Animals.Diet %>%
  # Add a priority score to sort by best rows
  mutate(
    priority = case_when(
      D.Unit.Amount %in% preferred_units &
        D.Unit.Time == preferred_time &
        D.Unit.Animals == preferred_animals ~ 1,
      TRUE ~ 2  # lower priority
    )
  ) %>%
  group_by(B.Code, A.Level.Name, D.Item) %>%
  slice_min(order_by = priority, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)  # optional: drop helper column

```


Summary of the data 
```{r,message=FALSE,warning=FALSE}

# Define species of interest
species_list <- c("Cattle", "Sheep", "Goat")

# Initialize result containers
species_summary <- list()
bcode_table <- list()

# Loop over each species
for (species in species_list) {
  
  # All B.Codes for this species
  species_bcodes <- merged_metadata$Prod.Out %>%
    filter(P.Product == species) %>%
    pull(B.Code) %>%
    unique()
  
  # Feed intake B.Codes
  feed_data <- merged_metadata$Data.Out %>%
    filter(B.Code %in% species_bcodes, !is.na(ED.Mean.T)) %>%
    filter(Out.Subind=="Feed Intake")%>%
    select(B.Code, D.Item = ED.Intake.Item)
  
  feed_bcodes <- unique(feed_data$B.Code)
  
  # GE B.Codes
  ge_data <- merged_metadata$Animals.Diet.Comp %>%
    filter(B.Code %in% species_bcodes, DC.Variable == "GE", !is.na(DC.Value)) %>%
    select(B.Code, D.Item)
  
  ge_bcodes <- unique(ge_data$B.Code)
  
  # Matched B.Codes (join on both B.Code and D.Item)
  matched_bcodes <- inner_join(feed_data, ge_data, by = c("B.Code", "D.Item")) %>%
    pull(B.Code) %>%
    unique()
  
  # Store species-level summary
  species_summary[[species]] <- tibble(
    Species = species,
    Total_BCodes = length(species_bcodes),
    With_FEED = length(feed_bcodes),
    With_GE = length(ge_bcodes),
    With_BOTH = length(matched_bcodes)
  )
  
  # Store long-format bcode list
  bcode_table[[species]] <- tibble(
    B.Code = c(species_bcodes, feed_bcodes, ge_bcodes, matched_bcodes),
    Category = rep(c("All", "With_FEED", "With_GE", "With_BOTH"), 
                   times = c(length(species_bcodes), length(feed_bcodes), length(ge_bcodes), length(matched_bcodes))),
    Species = species
  ) %>% distinct()
}

# Bind results
species_summary_df <- bind_rows(species_summary)
bcode_table_df <- bind_rows(bcode_table)

```


## 2) Harmonize units of ingredient amounts 
This step standardizes ingredient amount units (e.g., from mg, kg, %, etc.) to a consistent base (usually grams per individual per day), ensuring reliable downstream calculations and comparisons.

```{r,message=FALSE,warning=FALSE}
# diet ingredients: breakdown of all ingredients within each diet
ingredients <- merged_metadata$Animals.Diet %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    D.Unit.Time,
    D.Unit.Animals,
    DC.Is.Dry,
    D.Ad.lib,
    D.Is.Group
  ) %>%
  mutate(D.Amount = as.numeric(D.Amount))

# Function to harmonize diet ingredient units
harmonize_units <- function(df) {
  df %>%
    mutate(
      # Record original units (optional, for debugging/traceability)
      Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"),

      # Harmonize weight and volume units to a consistent base
      D.Amount = case_when(
        D.Unit.Amount %in% c("kg", "mg", "g/100g", "l", "kg/t", "g/300l", 
                             "mg/kg", "kg/100kg", "kg/100kg body weight", 
                             "kg/kg metabolic weight", "g/kg metabolic weight (0.75)", 
                             "g/kg DMI", "g/L", "g/kg DM") ~ {
          case_when(
            D.Unit.Amount == "kg" ~ D.Amount * 1000,                      # kg -> g
            D.Unit.Amount == "mg" ~ D.Amount / 1000,                     # mg -> g
            D.Unit.Amount == "g/100g" ~ D.Amount * 10,                   # g/100g -> g/kg
            D.Unit.Amount == "l" ~ D.Amount * 1000,                      # l -> ml
            D.Unit.Amount == "kg/t" ~ D.Amount * 1000,                   # kg/t -> g/kg
            D.Unit.Amount == "g/300l" ~ D.Amount / 300,                  # g/300l -> g/L
            D.Unit.Amount == "mg/kg" ~ D.Amount / 1000,                  # mg/kg -> g/kg
            D.Unit.Amount %in% c("kg/100kg", "kg/100kg body weight") ~ D.Amount * 10,     # -> g/kg body weight
            D.Unit.Amount == "kg/kg metabolic weight" ~ D.Amount * 1000, # -> g/kg metabolic weight
            D.Unit.Amount %in% c("g/kg body weight (0.75)", "g/kg Body Weight (0.75)") ~ D.Amount, # unchanged, just normalize name
            D.Unit.Amount == "g/kg DMI" ~ D.Amount,
            D.Unit.Amount == "g/kg DM" ~ D.Amount,
            TRUE ~ D.Amount
          )
        },
        TRUE ~ D.Amount
      ),

      # Update D.Unit.Amount to reflect new base
      D.Unit.Amount = case_when(
        D.Unit.Amount == "kg"  ~ "g",
        D.Unit.Amount == "mg"  ~ "g",
        D.Unit.Amount == "g/100g" ~ "g/kg",
        D.Unit.Amount == "l"  ~ "ml",
        D.Unit.Amount == "kg/t" ~ "g/kg",
        D.Unit.Amount == "g/300l" ~ "g/L",
        D.Unit.Amount == "mg/kg" ~ "g/kg",
        D.Unit.Amount %in% c("kg/100kg", "kg/100kg body weight") ~ "g/kg body weight",
        D.Unit.Amount %in% c("g/kg body weight (0.75)", "g/kg Body Weight (0.75)") ~ "g/kg metabolic weight",
        D.Unit.Amount == "kg/kg metabolic weight" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg metabolic weight (0.75)" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg DMI" ~ "g/kg",
        D.Unit.Amount == "g/kg DM" ~ "g/kg",
        D.Unit.Amount == "g/L" ~ "g/L",
        D.Unit.Amount == "g/kg Body Weight" ~ "g/kg body weight",  # case fix
        TRUE ~ D.Unit.Amount
      ),

      # Harmonize time units to day
      D.Amount = case_when(
        D.Unit.Time == "week" ~ D.Amount / 7,
        D.Unit.Time == "month" ~ D.Amount / 30,
        D.Unit.Time == "2 weeks" ~ D.Amount / 14,
        D.Unit.Time == "4 days interval" ~ D.Amount / 4,
        D.Unit.Time == "2x/day" ~ D.Amount * 2,
        TRUE ~ D.Amount
      ),
      D.Unit.Time = ifelse(!is.na(D.Unit.Time), "day", D.Unit.Time)
    ) %>%
    # Set unit fields to NA if amount is NA
    mutate(
      D.Unit.Amount   = ifelse(is.na(D.Amount), NA_character_, D.Unit.Amount),
      D.Unit.Time     = ifelse(is.na(D.Amount), NA_character_, D.Unit.Time),
      D.Unit.Animals  = ifelse(is.na(D.Amount), NA_character_, D.Unit.Animals)
    )
}



# Apply the harmonization
ingredients <- harmonize_units(ingredients)

# Load replicate info to adjust for "all animals in replicate"
reps <- merged_metadata$MT.Out %>%
  mutate(T.Animals = as.numeric(T.Animals)) %>%
  distinct(B.Code, A.Level.Name, .keep_all = TRUE)

ingredients <- ingredients %>%
  left_join(reps %>% dplyr::select(B.Code, A.Level.Name, T.Animals), by=c("B.Code","A.Level.Name")) %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Animals == "all animals in replicate" & !is.na(T.Animals) ~ D.Amount / T.Animals,
      TRUE ~ D.Amount
    ),
    D.Unit.Animals = ifelse(D.Unit.Animals == "all animals in replicate", "individual", D.Unit.Animals)
  )

# Refresh 'Units' column after these adjustments (optional)
ingredients <- ingredients %>%
  mutate(Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"))


```

### 2.1) suspicious dry amounts

This step identifies ingredients where DC.Is.Dry is marked as NA or unspecified. We cross-check these with moisture values from the composition table to infer whether the ingredient is likely dry or wet. We also flag inconsistencies between the ingredient name (e.g., containing "dried") and the recorded dry status.


```{r}
#look at ingredient namaes and mismatch with DC.Is.Dry
dry_name_mismatch <- ingredients %>%
  filter(
    str_detect(tolower(D.Item), "dry|dried"),       # name suggests it's dry
    is.na(DC.Is.Dry) | tolower(DC.Is.Dry) %in% c("no", "unspecified")  # but not marked as dry
  )



# 1. Extract Moisture values from composition data
moisture_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(str_to_lower(DC.Variable) %in% c("moisture", "moist")) %>%
  select(B.Code, D.Item, Moisture = DC.Value, Moisture.Unit = DC.Unit)%>%
  filter(!is.na(Moisture))


moisture_classified <- moisture_data %>%
  mutate(
    Unit.Standard = str_to_lower(Moisture.Unit),
    Moisture = as.numeric(Moisture),  # ensure numeric
    Dry_Prediction = case_when(
      Unit.Standard %in% c("%", "% dm", "g/100g") & Moisture < 15 ~ TRUE,
      Unit.Standard %in% c("%", "% dm", "g/100g") & Moisture > 50 ~ FALSE,
      Unit.Standard == "g/kg" & Moisture < 150 ~ TRUE,
      Unit.Standard == "g/kg" & Moisture > 500 ~ FALSE,
      TRUE ~ NA  # uncertain or mid-range
    )
  ) %>%
  left_join(
    ingredients %>% select(B.Code, D.Item, DC.Is.Dry),
    by = c("B.Code","D.Item")
  )%>%
  distinct()%>%
  select(-"Moisture.Unit")

dm_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(str_to_lower(DC.Variable) == "dm") %>%
  select(B.Code, D.Item, DM = DC.Value, DM.Unit = DC.Unit) %>%
  filter(!is.na(DM))

dm_classified <- dm_data %>%
  mutate(
    Unit.Standard = str_to_lower(DM.Unit),
    DM = as.numeric(DM),
   Dry_Prediction = case_when(
  # Clearly dry
  Unit.Standard %in% c("%", "% dm", "g/100g") & DM > 85 ~ TRUE,
  Unit.Standard == "g/kg" & DM > 850 ~ TRUE,

  # Clearly wet
  Unit.Standard %in% c("%", "% dm", "g/100g") & DM < 50 ~ FALSE,
  Unit.Standard == "g/kg" & DM < 500 ~ FALSE,

  # Mid-range or uncertain
  TRUE ~ NA
  )) %>%
  left_join(
    ingredients %>% select(B.Code, D.Item, DC.Is.Dry),
    by = c("B.Code", "D.Item")
  ) %>%
  distinct() %>%
  select(-"DM.Unit")


# Merge both moisture and DM classifications, keeping both unit standards
dry_prediction <- full_join(
  moisture_classified %>%
    rename(
      Dry_Prediction_Moisture = Dry_Prediction,
      Moisture_Value = Moisture,
      Moisture_Unit = Unit.Standard
    ),
  dm_classified %>%
    rename(
      Dry_Prediction_DM = Dry_Prediction,
      DM_Value = DM,
      DM_Unit = Unit.Standard
    ),
  by = c("B.Code", "D.Item", "DC.Is.Dry")
)

length(unique(dry_prediction$B.Code))
```


### 2.2) Filter data with acceptable units

We filter the dataset to retain only ingredients with acceptable units. Uncommon or unexpected unit formats are excluded to ensure consistency and simplify downstream processing

```{r,message=FALSE,warning=FALSE}


# Define a list of acceptable units
acceptable_units <- c(
  "g/kg", "g", "%", "ml", "L", "kg", "g/L",
  "g/kg metabolic weight", "% Diet", 
  "g/kg body weight", "ml/kg", "% Body Mass", 
  "% Concentrate", "g/kg body weight"  # leave only one if repeated
)


# Filter rows to keep only acceptable units
ingredients <- ingredients %>%
  filter(D.Unit.Amount %in% acceptable_units | is.na(D.Unit.Amount)) %>%
  mutate(D.Item = ifelse(is.na(D.Item), "Unspecified", D.Item))

#tracking paper loss because of harmnozied units
#total_studies_harmonized<- length(unique(ingredients_harmonized$B.Code))
#total_ingredients_harmonized<- length(unique(ingredients_harmonized$D.Item))
#total_diets_harmonized<- length(unique(ingredients_harmonized$A.Level.Name))


```

### 2.3) Convert body weight units by linking animals weight to data
```{r}

# Step 3: Join body weight
ingredients <- ingredients %>%
  left_join(
    merged_metadata$Data.Out %>%
      select(B.Code, A.Level.Name, Out.WG.Start, Out.WG.Unit),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    Out.WG.Unit = tolower(Out.WG.Unit),
    Out.WG.Start = case_when(
      Out.WG.Unit == "g" ~ Out.WG.Start / 1000,
      Out.WG.Unit == "mg" ~ Out.WG.Start / 1e6,
      Out.WG.Unit %in% c("kg", "unspecified", NA) ~ Out.WG.Start,
      TRUE ~ NA_real_
    ),
    Out.WG.Unit = "kg"
  ) %>%
  group_by(A.Level.Name) %>%
  mutate(
    Out.WG.Start = ifelse(is.na(Out.WG.Start),
                          first(na.omit(Out.WG.Start)),
                          Out.WG.Start)
  ) %>%
  ungroup() %>%
  distinct()

# Step 4: Convert ED.Mean.T to kg/day
ingredients <- ingredients %>%
  mutate(
    # Calculate ED.Mean.T in grams based on body weight if weight is available
    ED.Mean.T = case_when(
      D.Unit.Amount == "g/kg body weight" & !is.na(Out.WG.Start) ~ D.Amount * Out.WG.Start,
      D.Unit.Amount == "g/kg metabolic weight" & !is.na(Out.WG.Start) ~ D.Amount * Out.WG.Start^0.75,
      D.Unit.Amount == "% Body Mass" & !is.na(Out.WG.Start) ~ (D.Amount / 100) * Out.WG.Start * 1000,
      TRUE ~ NA_real_
    ),
    
    # Update unit to "g" only if ED.Mean.T was calculated
    D.Unit.Amount = if_else(!is.na(ED.Mean.T), "g", D.Unit.Amount)
  )

```

### 2.4) Estimate amounts using feed intake

For ingredients reported in relative units such as %, % Diet, or g/kg, we can estimate their absolute amounts when total diet feed intake is available. This allows us to convert proportions into standardized quantities.

#### 2.4.1) Harmonization of feed intake data

```{r}
feed_intake <- merged_metadata$Data.Out %>%
  select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw,
         ED.Mean.T, ED.Error, Out.Unit, Out.Subind) %>%
  filter(Out.Subind == "Feed Intake") %>%
  mutate(
    Out.Unit = tolower(trimws(Out.Unit)),

    # Extract unit components
    Out.Unit.Animals = str_extract(Out.Unit, "\\b(individual|replicate|animal replicate)\\b"),
    Out.Unit.Time    = str_extract(Out.Unit, "\\b(day|d|yr|year)\\b"),
    Out.Unit.Amount  = Out.Unit %>%
      str_remove_all("\\b(individual|replicate|animal replicate)\\b") %>%
      str_remove_all("\\b(day|d|yr|year)\\b") %>%
      str_replace_all("/+", "/") %>%
      str_trim() %>%
      str_remove("/$"))


# Step 3: Convert replicate-level intake to individual-level when possible
# Step 3: Convert replicate-level intake to individual-level when possible
feed_intake <- feed_intake %>%
  left_join(
    merged_metadata$MT.Out %>% select(B.Code, A.Level.Name, T.Animals),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    # Ensure numeric type
    ED.Mean.T = as.numeric(ED.Mean.T),
    T.Animals = as.numeric(T.Animals),

    # Convert intake and update unit
    ED.Mean.T = case_when(
      Out.Unit.Animals == "replicate" & !is.na(T.Animals) & T.Animals > 0 ~ ED.Mean.T / T.Animals,
      TRUE ~ ED.Mean.T
    ),
    Out.Unit.Animals = case_when(
      Out.Unit.Animals == "replicate" & !is.na(T.Animals) & T.Animals > 0 ~ "individual",
      TRUE ~ Out.Unit.Animals
    )
  )



# Define preferred units
preferred_units <- c("kg dm", "g dm")
preferred_time <- "day"
preferred_animals <- "individual"

# Create filtered table with prioritization
feed_intake <- feed_intake %>%
  mutate(
    priority = case_when(
      Out.Unit.Amount %in% preferred_units &
        Out.Unit.Time == preferred_time &
        Out.Unit.Animals == preferred_animals &
        !is.na(ED.Error) ~ 1,
      Out.Unit.Amount %in% preferred_units &
        Out.Unit.Time == preferred_time &
        Out.Unit.Animals == preferred_animals ~ 2,
      TRUE ~ 3
    )
  ) %>%
  group_by(B.Code, A.Level.Name, ED.Intake.Item) %>%
  slice_min(order_by = priority, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)  # optional



### keep unique units of ingredients 


# Step 3: Join body weight
feed_intake <- feed_intake %>%
  left_join(
    merged_metadata$Data.Out %>%
      select(B.Code, A.Level.Name, Out.WG.Start, Out.WG.Unit),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    Out.WG.Unit = tolower(Out.WG.Unit),
    Out.WG.Start = case_when(
      Out.WG.Unit == "g" ~ Out.WG.Start / 1000,
      Out.WG.Unit == "mg" ~ Out.WG.Start / 1e6,
      Out.WG.Unit %in% c("kg", "unspecified", NA) ~ Out.WG.Start,
      TRUE ~ NA_real_
    ),
    Out.WG.Unit = "kg"
  ) %>%
  group_by(A.Level.Name) %>%
  mutate(
    Out.WG.Start = ifelse(is.na(Out.WG.Start),
                          first(na.omit(Out.WG.Start)),
                          Out.WG.Start)
  ) %>%
  ungroup() %>%
  distinct()

# Step 4: Convert ED.Mean.T to kg/day
feed_intake <- feed_intake %>%
  mutate(
    Out.Unit = trimws(tolower(Out.Unit)),
    Out.Unit.Amount = trimws(str_remove(Out.Unit.Amount, "/$")),
    ED.Mean.T = case_when(
      Out.Unit.Amount %in% c("g", "g dm") ~ ED.Mean.T / 1000,
      Out.Unit.Amount %in% c("kg", "kg dm") ~ ED.Mean.T,
      Out.Unit.Amount == "g dm/100 g" ~ ED.Mean.T * 10 / 1000,
      Out.Unit.Amount == "g/week" ~ ED.Mean.T / 7 / 1000,
      Out.Unit.Amount == "g dm/kg metabolic weight" ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount == "g dm/kg body weight" ~ (ED.Mean.T * Out.WG.Start) / 1000,
      Out.Unit.Amount == "mg" ~ ED.Mean.T / 1e6,
      Out.Unit.Amount %in% c("g/kg body weight", "g dm/kg body weight") ~ (ED.Mean.T * Out.WG.Start) / 1000,
      Out.Unit.Amount %in% c("g/kg body weight (0.75)", "g dm/kg body weight (0.75)") ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount %in% c("g/kg metabolic weight","g dm/kg metabolic weight") ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount == "kg dm/kg metabolic weight" ~ ED.Mean.T * Out.WG.Start^0.75,
      Out.Unit.Amount %in% c("% body weight", "dm % body weight") ~ (ED.Mean.T / 100) * Out.WG.Start,
      Out.Unit.Amount == "g/g body weight" ~ (ED.Mean.T * Out.WG.Start) / 1000,
      TRUE ~ NA_real_
    ),
    Out.Unit.Amount = case_when(
      Out.Unit.Amount %in% c("g", "g dm", "g dm/100 g", "g/week", "g/kg body weight",
                             "g dm/kg body weight", "g/kg body weight (0.75)",
                             "g dm/kg body weight (0.75)", "g/kg metabolic weight","g dm/kg metabolic weight","g/g body weight",
                             "kg dm/kg metabolic weight", "% body weight", 
                             "dm % body weight", "kg", "kg dm", "mg") ~ "kg",
      TRUE ~ Out.Unit.Amount
    )
  ) %>%
  select(-Out.Unit)

feed_intake <- feed_intake %>%
  group_by(B.Code, A.Level.Name, ED.Intake.Item.Raw) %>%
  arrange(
    desc(!is.na(ED.Mean.T)), 
    desc(Out.Unit.Amount == "kg"),
    desc(Out.Unit.Animals == "individual"),
    desc(Out.Unit.Time == "day")
  ) %>%
  slice(1) %>%
  ungroup()
```

#### 2.4.2) Use of feed intake for ingredient amounts

If feed intake data is there for the entire diet and that the ingredients in ingredients dataset are in % or % of diet or g/kg we can convert 

```{r}
# Step 5: Estimate ingredient absolute amounts for diets with "Entire Diet" intake
ingredients <- ingredients %>%
  # Join only feed intake values where ED.Intake.Item is "Entire Diet"
  left_join(
    feed_intake %>%
      filter(ED.Intake.Item == "Entire Diet") %>%
      select(B.Code, A.Level.Name, Intake_kg_day = ED.Mean.T),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    # Estimate absolute amount (in grams)
    D.Amount = case_when(
      D.Unit.Amount == "%" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "% Diet" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "g/kg" & !is.na(Intake_kg_day) ~ D.Amount * Intake_kg_day,
      TRUE ~ D.Amount
    ),
    
    # Update unit only if we did a conversion
    D.Unit.Amount = case_when(
      D.Unit.Amount %in% c("%", "% Diet", "g/kg") & !is.na(Intake_kg_day) ~ "g",
      TRUE ~ D.Unit.Amount
    )
  ) %>%
  select(-Intake_kg_day)%>%
  select(-ED.Mean.T)


```

## 3) Integrate basal diets 

Basal diets serve as the control and foundation for all treatments in livestock experiments. They are present in both the control diet and in every treatment diet that includes additional ingredients. This section integrates basal diet ingredients into the treatment diets while preserving the original basal formulation, and it flags any ingredient that originates from the basal diet.Basal diets (base diets) are merged into treatment diets to ensure consistency across feeding trials, allowing for a more accurate estimate of total diet composition.

```{r,message=FALSE,warning=FALSE}
# Identify basal diets
basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) == "base") %>%
  mutate(from_basal = TRUE)


# Identify non-basal diets
non_basal_diets <- ingredients %>%
  filter(tolower(str_trim(A.Level.Name)) != "base") %>%
  mutate(from_basal = FALSE)

# Duplicate basal diets for each non-basal diet within the same B.Code
basal_expanded <- non_basal_diets %>%
  dplyr::select(B.Code, A.Level.Name) %>%
  distinct() %>%
  inner_join(basal_diets, by="B.Code") %>%
  mutate(
    A.Level.Name = A.Level.Name.x, # carry over the non-basal A.Level.Name
    from_basal   = TRUE
  ) %>%
  dplyr::select(-A.Level.Name.x, -A.Level.Name.y)

# Combine everything into one integrated dataset
ingredients <- bind_rows(
  non_basal_diets,   # original non-basal
  basal_expanded,    # basal items added to each non-basal diet
  basal_diets        # keep the purely basal diet for control
)



```

Similarly to the previous step, we can estimate the amounts of ingredients originating from basal diets using the total feed intake—provided the units are reported in % or g/kg.
```{r}
# Step 5: Estimate ingredient absolute amounts for diets with "Entire Diet" intake
ingredients <- ingredients %>%
  # Join only feed intake values where ED.Intake.Item is "Entire Diet"
  left_join(
    feed_intake %>%
      filter(ED.Intake.Item == "Entire Diet") %>%
      select(B.Code, A.Level.Name, Intake_kg_day = ED.Mean.T),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    # Estimate absolute amount (in grams)
    D.Amount = case_when(
      D.Unit.Amount == "%" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "% Diet" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "g/kg" & !is.na(Intake_kg_day) ~ D.Amount * Intake_kg_day,
      TRUE ~ D.Amount
    ),
    
    # Update unit only if we did a conversion
    D.Unit.Amount = case_when(
      D.Unit.Amount %in% c("%", "% Diet", "g/kg") & !is.na(Intake_kg_day) ~ "g",
      TRUE ~ D.Unit.Amount
    )
  ) %>%
  select(-Intake_kg_day)

```


## 4) Replace feed intake data where available

We prioritize feed intake data over the amounts reported in the methods section, as feed intake is the key variable needed for emission calculations. When feed intake data is unavailable, we fall back on the harmonized ingredient amounts extracted earlier.

### 4.1) Feed intake data at the diet level 

```{r,message=FALSE,warning=FALSE}
# Step 5: Remove feed intake entries that are actually full-diet labels
feed_intake_ingredients <- feed_intake %>%
  filter(ED.Intake.Item != A.Level.Name | ED.Intake.Item != "Entire Diet")



# Step 6: Merge with ingredients and override with ED.Mean.T when available
# FIXED VERSION
# Merge with ingredients and override with ED.Mean.T when available
ingredients <- ingredients %>%
  left_join(
    feed_intake_ingredients %>%
      select(B.Code, A.Level.Name, ED.Intake.Item, ED.Mean.T, Out.Unit.Amount, Out.Unit.Animals, Out.Unit.Time),
    by = c("B.Code", "A.Level.Name", "D.Item" = "ED.Intake.Item")
  ) %>%
  mutate(
    from_feed_intake = !is.na(ED.Mean.T),
    D.Amount = if_else(from_feed_intake, ED.Mean.T, D.Amount),
    D.Unit.Amount = if_else(from_feed_intake & !is.na(Out.Unit.Amount), Out.Unit.Amount, D.Unit.Amount),
    D.Unit.Animals = if_else(from_feed_intake & !is.na(Out.Unit.Animals), Out.Unit.Animals, D.Unit.Animals),
    D.Unit.Time = if_else(from_feed_intake & !is.na(Out.Unit.Time), Out.Unit.Time, D.Unit.Time)
  ) %>%
  select(-ED.Mean.T, -Out.Unit.Amount, -Out.Unit.Animals, -Out.Unit.Time) %>%
  distinct()





# Step 8: Final conversion of unit to grams (g)
ingredients <- ingredients %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Amount == "kg" ~ D.Amount * 1000,
      TRUE ~ D.Amount
    ),
    D.Unit.Amount = case_when(
      D.Unit.Amount == "kg" ~ "g",
      TRUE ~ D.Unit.Amount
    )
  )


```


### 4.2) Feed intake at the diet level  

```{r}
# Step: Extract feed intake entries that refer to the entire diet
feed_intake_diet <- feed_intake %>%
  filter(ED.Intake.Item == A.Level.Name | ED.Intake.Item == "Entire Diet")

# Step: Merge and overwrite ingredients using diet-level feed intake
ingredients <- ingredients %>%
  left_join(
    feed_intake_diet %>%
      select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw, ED.Mean.T, Out.Unit.Amount, Out.Unit.Animals, Out.Unit.Time),
    by = c("B.Code", "A.Level.Name", "D.Type" = "ED.Intake.Item.Raw")
  ) %>%
  mutate(
    from_feed_intake2 = !is.na(ED.Mean.T),
    D.Amount = if_else(from_feed_intake2, ED.Mean.T, D.Amount),
    D.Unit.Amount = if_else(from_feed_intake2 & !is.na(Out.Unit.Amount), Out.Unit.Amount, D.Unit.Amount),
    D.Unit.Animals = if_else(from_feed_intake2 & !is.na(Out.Unit.Animals), Out.Unit.Animals, D.Unit.Animals),
    D.Unit.Time = if_else(from_feed_intake2 & !is.na(Out.Unit.Time), Out.Unit.Time, D.Unit.Time)
  ) %>%
  select(-ED.Mean.T, -Out.Unit.Amount, -Out.Unit.Animals, -Out.Unit.Time,-ED.Intake.Item) %>%
  distinct()




# Step 8: Final conversion of unit to grams (g)
ingredients <- ingredients %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Amount == "kg" ~ D.Amount * 1000,
      TRUE ~ D.Amount
    ),
    D.Unit.Amount = case_when(
      D.Unit.Amount == "kg" ~ "g",
      TRUE ~ D.Unit.Amount
    )
  )


```


now for ingredient if there is only one missing in the diet (not taking water into account) and that we have the total feed intake we can deduce it form the total feed intake by the other ingredients amounts 

TO DO 

## 5) Suspicious mismatch between ingredient diet names and feed intake diet names

Some feed intake entries are not matched to any ingredient data due to discrepancies in A.Level.Name between the feed intake and ingredient datasets for the same study codes. In this section, we build a table to flag studies where diet names do not align, helping to identify cases that may require manual review or correction.

```{r}
# Extract unique combinations
ingredient_levels <- ingredients %>%
  distinct(B.Code, A.Level.Name) %>%
  mutate(Source = "Ingredient")

feedintake_levels <- feed_intake %>%
  distinct(B.Code, A.Level.Name) %>%
  mutate(Source = "FeedIntake")

# Combine both
combined_levels <- bind_rows(ingredient_levels, feedintake_levels)

# Count presence in each source
comparison_table <- combined_levels %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    Sources = paste(sort(unique(Source)), collapse = " & "),
    .groups = "drop"
  ) %>%
  mutate(
    Match_Status = case_when(
      Sources == "FeedIntake & Ingredient" ~ "Match",
      Sources == "Ingredient" ~ "Only in Ingredient",
      Sources == "FeedIntake" ~ "Only in FeedIntake",
      TRUE ~ "Unknown"
    )
  )



# Step 1: Find B.Codes with both statuses

bcode_dual_mismatch <- comparison_table %>%
  filter(Match_Status %in% c("Only in Ingredient", "Only in FeedIntake")) %>%
  group_by(B.Code) %>%
  summarise(n_status = n_distinct(Match_Status), .groups = "drop") %>%
  filter(n_status == 2) %>%
  pull(B.Code)

# Step 2: Subset to those B.Codes and remove the matches
comparison_table_subset <- comparison_table %>%
  filter(B.Code %in% bcode_dual_mismatch) %>%
  filter(Match_Status != "Match")


#write.csv(comparison_table_subset, "Diet_name_match.csv", row.names = FALSE)


```


## 6) Subset for Dry data 

When feed intake data is missing and we rely on ingredient-level amounts, we need to ensure that the quantities used refer to the dry form of the ingredient. To do this, we subset the data to include only ingredients identified as dried—based on either the declared DC.Is.Dry value or our earlier flagging using ingredient names and moisture content predictions. This ensures consistency and accuracy in downstream calculations. 
```{r,message=FALSE,warning=FALSE}
#take out unspecified ingredients 

ingredients %>%
  group_by(DC.Is.Dry) %>%
  summarise(
    n_unique_ingredients = n_distinct(D.Item)
  ) %>%
  arrange(desc(n_unique_ingredients))

ingredients <- ingredients %>%
  left_join(
    dry_prediction %>%
      select(B.Code, D.Item, Dry_Prediction_Moisture, Dry_Prediction_DM),
    by = c("B.Code", "D.Item")
  )


ingredients <- ingredients %>%
  mutate(
    D.Amount = ifelse(
      (
        # Not marked as dry
        (DC.Is.Dry %in% c("No", "N", "Unspecified") | is.na(DC.Is.Dry)) &
        # Not from feed intake
        (is.na(from_feed_intake) | from_feed_intake == FALSE) &
        # Name does NOT suggest it's dry
        !str_detect(tolower(D.Item), "dry|dried") &
        # And not predicted to be dry from either source
        !(Dry_Prediction_Moisture == TRUE | Dry_Prediction_DM == TRUE)
      ),
      NA_real_,  # set to NA
      D.Amount    # keep original
    )
  )



```


## 7) Complete the nutrition and digestibility papers

Not all ingredients and diets are currently included in the digestibility and composition tables—only those for which data is available. We aim to add the missing ones to help complete the dataset and enable more comprehensive analysis.

```{r}
# --- Step 1: Get all unique (B.Code, D.Item, Source) from the Diet table ---
all_diet_items <- merged_metadata$Animals.Diet %>%
  distinct(B.Code, A.Level.Name, D.Item, Source) %>%
  mutate(
    is_entire_diet = if_else(is.na(D.Item), TRUE, FALSE),
    D.Item = coalesce(D.Item, A.Level.Name)
  ) %>%
  select(B.Code, D.Item, is_entire_diet, Source) %>%
  distinct()

# --- Step 2: Get unique variables in the nutrition and digestibility datasets ---
nutrition_variables <- unique(na.omit(merged_metadata$Animals.Diet.Comp$DC.Variable))
digestibility_variables <- unique(na.omit(merged_metadata$Animals.Diet.Digest$DD.Variable))

# --- Step 3: Expand to full grid of all D.Item × variables (by B.Code and Source) ---
# Nutrition (Comp)
comp_missing <- all_diet_items %>%
  crossing(DC.Variable = nutrition_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Comp %>% select(B.Code, D.Item, DC.Variable),
    by = c("B.Code", "D.Item", "DC.Variable")
  ) %>%
  mutate(
    DC.Value = NA_real_,
    DC.Unit = NA_character_
  )

# Digestibility (Digest)
digest_missing <- all_diet_items %>%
  crossing(DD.Variable = digestibility_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Digest %>% select(B.Code, D.Item, DD.Variable),
    by = c("B.Code", "D.Item", "DD.Variable")
  ) %>%
  mutate(
    DD.Value = NA_real_,
    DD.Unit = NA_character_
  )

# --- Step 4: Bind the missing values into the existing datasets ---
merged_metadata$Animals.Diet.Comp <- bind_rows(merged_metadata$Animals.Diet.Comp, comp_missing)
merged_metadata$Animals.Diet.Digest <- bind_rows(merged_metadata$Animals.Diet.Digest, digest_missing)

```


```{r}
# --- Step 5: Ensure each ingredient has a row for each nutrition variable ---
comp_completion <- merged_metadata$Animals.Diet.Comp %>%
  distinct(B.Code, D.Item, is_entire_diet, Source) %>%
  crossing(DC.Variable = nutrition_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Comp %>% select(B.Code, D.Item, DC.Variable),
    by = c("B.Code", "D.Item", "DC.Variable")
  ) %>%
  mutate(
    DC.Value = NA_real_,
    DC.Unit = NA_character_
  )

# --- Step 6: Ensure each ingredient has a row for each digestibility variable ---
digest_completion <- merged_metadata$Animals.Diet.Digest %>%
  distinct(B.Code, D.Item, is_entire_diet, Source) %>%
  crossing(DD.Variable = digestibility_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Digest %>% select(B.Code, D.Item, DD.Variable),
    by = c("B.Code", "D.Item", "DD.Variable")
  ) %>%
  mutate(
    DD.Value = NA_real_,
    DD.Unit = NA_character_
  )

# --- Step 7: Bind additional completion rows ---
merged_metadata$Animals.Diet.Comp <- bind_rows(merged_metadata$Animals.Diet.Comp, comp_completion)
merged_metadata$Animals.Diet.Digest <- bind_rows(merged_metadata$Animals.Diet.Digest, digest_completion)

```

## 8) Link ingredients to feedipedia 

Feedipedia is an online database that provides comprehensive information on feed ingredients used in livestock nutrition. It aggregates data on the nutritional composition, digestibility, and other relevant properties of various feeds.We will use it to infer data when info is missing from our data extraction. Links each ingredient to its corresponding entry in Feedipedia, enabling supplementation of missing nutritional and digestibility data such as GE, CP, EE, and Ash.

```{r,message=FALSE,warning=FALSE}

## download data from feedapedia (generated with the "feedapedia scrapping script available on github)

# Use the raw URL of the file on GitHub
url <- "https://github.com/ERAgriculture/ERL/raw/main/downloaded_data/feedipedia.parquet"

# Create a temporary file location with a .parquet extension
temp_file <- tempfile(fileext = ".parquet")

# Download the file in binary mode
download.file(url, destfile = temp_file, mode = "wb")

# Read the Parquet file
feedipedia <- read_parquet(temp_file)

## download AOM with the correspondance between feedipedia codes and diet items 

# Define the raw URL for the Excel file
url <- "https://raw.githubusercontent.com/peetmate/era_codes/main/era_master_sheet.xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, destfile = temp_file, mode = "wb")

# Read only the sheet called "AOM"
mastersheet <- read_excel(temp_file, sheet = "AOM") %>%
  filter(L5=="Feed Ingredient")

mastersheet$Feedipedia <- as.numeric(sub(".*node/", "", mastersheet$Feedipedia))


```

add nutrition and digestibility from feedipedia 

```{r,message=FALSE,warning=FALSE}
# --- A) Add Feedipedia node from mastersheet into Comp and Digest tables
add_feedipedia_node <- function(df, mastersheet) {
  df %>%
    left_join(
      mastersheet %>% select(Edge_Value, Feedipedia),
      by = c("D.Item" = "Edge_Value")
    ) %>%
    distinct()
}

merged_metadata$Animals.Diet.Comp   <- add_feedipedia_node(merged_metadata$Animals.Diet.Comp, mastersheet)
merged_metadata$Animals.Diet.Digest <- add_feedipedia_node(merged_metadata$Animals.Diet.Digest, mastersheet)

# --- B) Standardize Feedipedia variables and assign categories
digestibility_keywords <- c(
  "digestibility", "degradability", "energy digestibility", 
  "DE", "ME", "NE", "AMEn", "OM digestibility", "DM digestibility",
  "ruminants", "rabbit", "pig", "poultry", "horse", "salmonids", "hydrolysis"
)

feedipedia <- feedipedia %>%
  mutate(
    Variable = case_when(
      tolower(Variable) == "gross energy" ~ "GE",
      tolower(Variable) == "crude protein" ~ "CP",
      tolower(Variable) == "ether extract" ~ "EE",
      tolower(Variable) == "ash" ~ "Ash",
      tolower(Variable) == "Neutral detergent fibre
" ~ "NDF",
      TRUE ~ Variable
    ),
    category = case_when(
      Variable %in% c("GE", "CP", "EE", "Ash","NDF") ~ "nutrition",
      str_detect(Variable, regex(paste(digestibility_keywords, collapse = "|"), ignore_case = TRUE)) ~ "digestibility",
      TRUE ~ "nutrition"
    ),
    feedipedia_node = as.numeric(feedipedia_node)
  )

# --- C) Ensure source columns exist
if (!"nutrition_source" %in% names(merged_metadata$Animals.Diet.Comp)) {
  merged_metadata$Animals.Diet.Comp$nutrition_source <- NA_character_
}
if (!"digestibility_source" %in% names(merged_metadata$Animals.Diet.Digest)) {
  merged_metadata$Animals.Diet.Digest$digestibility_source <- NA_character_
}

# --- D) Nutrition variables: GE, CP, EE, Ash
nutrition_vars <- c("GE", "CP", "EE", "Ash","NDF")

feedipedia_nutrition <- feedipedia %>%
  filter(Variable %in% nutrition_vars) %>%
  group_by(feedipedia_node, Variable, Unit) %>%
  summarise(Value = mean(Avg, na.rm = TRUE), .groups = "drop") %>%
  rename(DC.Variable = Variable, DC.Value.feedipedia = Value)

# --- E) Join nutrition into Animals.Diet.Comp
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(Feedipedia = as.numeric(Feedipedia)) %>%
  left_join(feedipedia_nutrition, by = c("Feedipedia" = "feedipedia_node", "DC.Variable")) %>%
  mutate(
    value_missing = is.na(DC.Value),
    DC.Value = coalesce(DC.Value, DC.Value.feedipedia),
    DC.Unit  = if_else(value_missing & !is.na(DC.Value.feedipedia), Unit, DC.Unit),
    nutrition_source = if_else(value_missing & !is.na(DC.Value.feedipedia), "feedipedia", nutrition_source)
  ) %>%
  select(-DC.Value.feedipedia, -Unit, -value_missing)

# --- F) Digestibility: Energy digestibility, ruminants → DE
feedipedia_digestibility <- feedipedia %>%
  filter(str_detect(tolower(Variable), "energy digestibility")) %>%
  mutate(
    DD.Variable = "DE",  # Map all energy digestibility to "DE"
    feedipedia_node = as.numeric(feedipedia_node)
  ) %>%
  group_by(feedipedia_node, DD.Variable, Unit) %>%
  summarise(DD.Value.feedipedia = mean(Avg, na.rm = TRUE), .groups = "drop")

merged_metadata$Animals.Diet.Digest <- merged_metadata$Animals.Diet.Digest %>%
  mutate(Feedipedia = as.numeric(Feedipedia)) %>%
  left_join(feedipedia_digestibility, by = c("Feedipedia" = "feedipedia_node", "DD.Variable")) %>%
  mutate(
    value_missing = is.na(DD.Value),
    DD.Value = coalesce(DD.Value, DD.Value.feedipedia),
    DD.Unit  = if_else(value_missing & !is.na(DD.Value.feedipedia), Unit, DD.Unit),
    digestibility_source = if_else(value_missing & !is.na(DD.Value.feedipedia), "feedipedia", digestibility_source)
  ) %>%
  select(-DD.Value.feedipedia, -Unit, -value_missing)

```


## 9) Infer nutritional and digestibility data from raw data and feedipedia  


Now that we have both raw and Feedipedia data, we can use them to infer missing values. If some instances of an ingredient have nutritional or digestibility data while others do not, we can impute the missing values using the average of the available ones.

### 9.1) Nutritional composition inferrence

```{r}
# --- Standardize Units in Animals.Diet.Comp ---
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(
    DC.Unit = trimws(tolower(DC.Unit)),
    DC.Value = as.numeric(DC.Value),
    
    DC.Value = case_when(
      DC.Variable == "GE" ~ case_when(
        DC.Unit %in% c("mj/kg", "mj/kg dm") ~ DC.Value,
        DC.Unit == "kcal/kg" ~ DC.Value / 239,
        DC.Unit == "kcal/kg dm" ~ DC.Value / 239,
        DC.Unit == "kcal/g" ~ DC.Value * 1000 / 239,
        DC.Unit == "kcal/100g" ~ DC.Value * 10 / 239,
        DC.Unit == "cal/g" ~ DC.Value / 239,
        DC.Unit == "mj/g" ~ DC.Value * 1000,
        DC.Unit == "mcal/kg" ~ DC.Value * 4.184,
        DC.Unit %in% c("kj/g", "kj/100g", "kj/100g dm", "kj/100 g dm") ~ DC.Value / 100,
        DC.Unit == "kj/kg dm" ~ DC.Value / 1000,
        DC.Unit == "kj" ~ DC.Value / 1000,
        DC.Unit == "j/mg" ~ DC.Value / 1e6,
        DC.Unit == "mj/100g" ~ DC.Value * 10,
        TRUE ~ NA_real_
      ),
      # All other nutrients → g/kg
      TRUE ~ case_when(
        DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value * 10,
        DC.Unit %in% c("g", "g/kg", "g/kg dm") ~ DC.Value,
        DC.Unit %in% c("kg", "kg/day") ~ DC.Value * 1000,
        TRUE ~ NA_real_
      )
    ),
    DC.Unit = case_when(
      DC.Variable == "GE" ~ "MJ/kg",
      TRUE ~ "g/kg"
    )
  )

# --- Compute Mean Value Per Ingredient and Variable ---
ingredient_means <- merged_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet == FALSE, !is.na(DC.Value)) %>%
  group_by(D.Item, DC.Variable) %>%
  summarise(mean_value = mean(DC.Value, na.rm = TRUE), .groups = "drop")

merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  left_join(ingredient_means, by = c("D.Item", "DC.Variable")) %>%
  mutate(
    nutrition_source = case_when(
      nutrition_source == "feedipedia" ~ "feedipedia",  # preserve Feedipedia
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value) ~ "inferred_from_same_ingredient",  # imputed
      is.na(nutrition_source) & !is.na(DC.Value) ~ "raw data",  # only if there's a value
      TRUE ~ nutrition_source  # keep whatever is there
    ),
    DC.Value = if_else(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value),
      mean_value,
      DC.Value
    )
  ) %>%
  select(-mean_value)


```

### 9.2) Digestibility inferrence
```{r}

# --- B) Compute Mean Digestibility Per Ingredient and Variable ---
digest_means <- merged_metadata$Animals.Diet.Digest %>%
  filter(DD.Variable == "DE", is_entire_diet == FALSE, !is.na(DD.Value)) %>%
  group_by(D.Item, DD.Variable, DD.Unit) %>%
  summarise(mean_digest = mean(DD.Value, na.rm = TRUE), .groups = "drop")


# --- C) Fill Missing Digestibility Values with Ingredient Mean ---
merged_metadata$Animals.Diet.Digest <- merged_metadata$Animals.Diet.Digest %>%
  left_join(digest_means, by = c("D.Item", "DD.Variable")) %>%
  mutate(
    digestibility_source = case_when(
      digestibility_source == "feedipedia" ~ "feedipedia",
      is.na(DD.Value) & !is_entire_diet & !is.na(mean_digest) ~ "inferred_from_same_ingredient",
      is.na(digestibility_source) ~ "raw data",
      TRUE ~ digestibility_source
    ),
    DD.Value = if_else(
      is.na(DD.Value) & !is_entire_diet & !is.na(mean_digest),
      mean_digest,
      DD.Value
    )
  ) %>%
  select(-mean_digest)

```

since there is only very few digestibility in the raw data logic that we dont gain any data by doing this because feedipedia is infered for all the ingredients we have the info for. 

```{r,message=FALSE,warning=FALSE}
# ✅ Count how many papers had GE filled by Feedipedia, and how many of those are Cattle / Sheep / Goat

# Step 1: Filter for Feedipedia-filled GE entries
ge_filled_by_feedipedia <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", nutrition_source == "feedipedia") %>%
  distinct(B.Code)

# Step 2: Total number of studies
n_ge_feedipedia <- nrow(ge_filled_by_feedipedia)

# Step 3: Add species info and filter to ruminants
ge_species_feedipedia <- ge_filled_by_feedipedia %>%
  left_join(merged_metadata$`Prod.Out` %>% select(B.Code, P.Product), by = "B.Code") %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product)

# Step 4: Count per species
ge_species_counts <- ge_species_feedipedia %>%
  count(P.Product, name = "Papers_with_GE_from_Feedipedia")

# Display summary
cat("📌 Total papers with GE filled from Feedipedia:", n_ge_feedipedia, "\n")
cat("📌 Of those, species breakdown:\n")
print(ge_species_counts)

```


```{r,message=FALSE,warning=FALSE}

# Define species of interest
species_list <- c("Cattle", "Sheep", "Goat")

# Initialize result containers
species_summary <- list()
bcode_table <- list()

# Loop over each species
for (species in species_list) {
  
  # All B.Codes for this species
  species_bcodes <- merged_metadata$Prod.Out %>%
    filter(P.Product == species) %>%
    pull(B.Code) %>%
    unique()
  
  # Feed intake B.Codes
  feed_data <- merged_metadata$Data.Out %>%
    filter(B.Code %in% species_bcodes, !is.na(ED.Mean.T)) %>%
    select(B.Code, D.Item = ED.Intake.Item)
  
  feed_bcodes <- unique(feed_data$B.Code)
  
  # GE B.Codes
  ge_data <- merged_metadata$Animals.Diet.Comp %>%
    filter(B.Code %in% species_bcodes, DC.Variable == "GE", !is.na(DC.Value)) %>%
    select(B.Code, D.Item)
  
  ge_bcodes <- unique(ge_data$B.Code)
  
  # Matched B.Codes (join on both B.Code and D.Item)
  matched_bcodes <- inner_join(feed_data, ge_data, by = c("B.Code", "D.Item")) %>%
    pull(B.Code) %>%
    unique()
  
  # Store species-level summary
  species_summary[[species]] <- tibble(
    Species = species,
    Total_BCodes = length(species_bcodes),
    With_FEED = length(feed_bcodes),
    With_GE = length(ge_bcodes),
    With_BOTH = length(matched_bcodes)
  )
  
  # Store long-format bcode list
  bcode_table[[species]] <- tibble(
    B.Code = c(species_bcodes, feed_bcodes, ge_bcodes, matched_bcodes),
    Category = rep(c("All", "With_FEED", "With_GE", "With_BOTH"), 
                   times = c(length(species_bcodes), length(feed_bcodes), length(ge_bcodes), length(matched_bcodes))),
    Species = species
  ) %>% distinct()
}

# Bind results
species_summary_df <- bind_rows(species_summary)
bcode_table_df <- bind_rows(bcode_table)

# Show results
print(species_summary_df)
head(bcode_table_df)
```


## 10) Relative % of each ingredient in the diet

This table presents the relative percentage of each ingredient within a diet. To ensure accurate proportion calculations, we first filtered the dataset to include only diets with consistent units. Diets with incompatible or non-harmonizable units were excluded, as their proportions could not be reliably determined.

```{r,message=FALSE,warning=FALSE}
# We now have a single integrated dataset: 'ingredients'.
# 1) Filter to diets that have a single unit type per group (consistent units).
# 2) Compute total amounts and percentages.

inconsistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  filter(Unit_Count > 1) %>%
  ungroup()

# Step 1: Flag only consistent-unit diets
consistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  # Count only the non-NA units
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  # Keep groups with 0 or 1 distinct unit (ignoring NA)
  filter(Unit_Count <= 1) %>%
  ungroup()

#porportion of inconsistent diets 
proportion_inconsistent_diets<- length(unique(inconsistent_diets$A.Level.Name))/length(unique(ingredients$A.Level.Name))

# Step 2: Sum total amounts per (B.Code, A.Level.Name, D.Unit.Amount)
# Transform consistent_diets into a summary of totals, removing any NA in amount or unit
# Step 1: Count valid and missing D.Amount per diet
diet_completeness <- consistent_diets %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    n_total = n(),
    n_nonmissing = sum(!is.na(D.Amount)),
    .groups = "drop"
  ) %>%
  filter(n_total == n_nonmissing)  # only fully specified diets


# Step 2: Keep only complete diets when summing
total_amounts <- consistent_diets %>%
  semi_join(diet_completeness, by = c("B.Code", "A.Level.Name")) %>%
  group_by(B.Code, A.Level.Name, D.Unit.Amount) %>%
  summarise(Total_Amount = sum(D.Amount), .groups = "drop")

# Display the resulting table


#Create an interactive table for "total_amounts"
datatable(
  total_amounts,
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)



# Step 3: Calculate percentages
diet_percentages <- consistent_diets %>%
  left_join(total_amounts, by=c("B.Code","A.Level.Name","D.Unit.Amount")) %>%
  mutate(
    Percentage_of_Diet = ifelse(Total_Amount > 0,
                                (D.Amount / Total_Amount)*100,
                                NA_real_)
  ) %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    Percentage_of_Diet,
    from_basal
  )

# Table 2: Relative percentage of each ingredient in each diet


#Create an interactive table for "total_amounts"
datatable(
  diet_percentages,
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)



```

```{r,message=FALSE,warning=FALSE}
# Check if each diet sums close to 100%
validation <- diet_percentages %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(Total_Percentage = sum(Percentage_of_Diet, na.rm=TRUE)) %>%
  ungroup()

# Display diets and their total percentages
# datatable(
#   validation, 
#   options = list(
#     pageLength = 10,         # Show 10 rows per page by default
#     scrollX = TRUE,          # Enable horizontal scrolling
#     autoWidth = TRUE,        # Adjust column width automatically
#     searchHighlight = TRUE   # Highlight search terms
#   ),
#   caption = "Validation table"
# )


```

The lines that have 0 as validation are the ones that were feeding ad libitum for which we do not have a precise amount so it's impossible to calculate the relative proportion of the ingredients.

if we have total amounts and feed intake for ingredient we might be abe to deduce ad libitum feed intake data 



## 11) Emission calculations 

### 11.1 Estimation of Enteric Methane Emissions Using IPCC Tier 2

Enteric methane (CH₄) emissions were estimated using the **IPCC Tier 2** methodology, which incorporates animal-specific data such as **gross energy intake (GE)** and **methane conversion factors (Ym)**. This approach improves accuracy over Tier 1 by reflecting diet composition, feed intake, and species-specific differences.


### 11.2 Estimation of Daily GE Intake per Animal

For GE values reported at the **entire diet** level, daily gross energy intake per animal was calculated as:

\[
\text{GE}_{\text{daily}} = \text{GE}_{\text{diet}} \times \text{Diet Dry Amount}
\]

Where:

- $\text{GE}_{\text{diet}}$ is the energy content of the complete diet (MJ/kg)
- $\text{Diet Dry Amount}$ is the daily dry matter ingredient (kg/day/animal). We can also infer from feed intake if missing.Needs to be dry matter so we will use only data when DC.is.Dry=yes.

**To do that we will have ton convert all feed intake data into kg/day/animal.**

When GE values were reported at the **ingredient** level, and information on diet composition (ingredient proportions) was available, the diet-level GE was reconstructed as:

\[
\text{GE}_{\text{diet}} = \sum_{i=1}^{n} (\text{GE}_i \times p_i)
\]

Where:

- $\text{GE}_i$ is the GE value of ingredient $i$
- $p_i$ is the proportion of ingredient $i$ in the total diet

Then, daily intake was computed using the same formula above.

### 11.3. Methane Emissions Calculation (Tier 2)

Annual methane emissions were calculated using the IPCC Tier 2 equation:

\[
\text{CH}_4 = \frac{\text{GE}_{\text{daily}} \times \text{Ym} \times 365}{55.65}
\]

Where:

- $\text{CH}_4$ = annual enteric methane emissions (kg CH₄/animal/year)
- $\text{GE}_{\text{daily}}$ = daily gross energy intake per animal (MJ/day)
- $\text{Ym}$ = methane conversion factor:
  - **Cattle**: 0.065 (6.5%)
  - **Sheep and Goats**: 0.060 (6.0%)
- 365 = days per year
- 55.65 = energy content of methane (MJ/kg CH₄)


### 11.4 Data Requirements and Supporting Tables

**1. Table of Diet Composition Proportions**

This table defines the proportion of each ingredient in the overall diet:

```{r,message=FALSE,warning=FALSE}

# Create an interactive table for "total_amounts"
datatable(
  diet_percentages, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)
```


Used to compute reconstructed `GE_diet` from ingredient-level GE values:

\[
\text{GE}_{\text{diet}} = \sum_{i=1}^{n} (\text{GE}_i \times p_i)
\]

---

**2. Table of Ingredient Dry Amounts**

This table stores the amount of each ingredient fed per animal per day, already in or converted to dry matter:

```{r,message=FALSE,warning=FALSE}
# This 'ingredients' now contains all diets + basal integrated.
# Table 1: Final diet amounts with harmonized units (including basal)
datatable(
  ingredients, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Diet Ingredients"
)
```


Useful when GE is per ingredient and diet is not proportionally described.

---

**3. Table of Total Diet Dry Amounts**

This is the total daily dry matter offered or consumed per animal, either directly from intake or inferred from ingredient sums:

```{r,message=FALSE,warning=FALSE}
datatable(
  total_amounts, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)

```


\[
\text{GE}_{\text{daily}} = \text{GE}_{\text{diet}} \times \text{Diet Dry Amount}
\]





```{r,message=FALSE,warning=FALSE}

# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

```


## 12) Harmonize GE units 
```{r,message=FALSE,warning=FALSE}
# Inspect the units used for GE


Gross_energy <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE")

Gross_energy$DC.Value<-as.numeric(Gross_energy$DC.Value)

unique(Gross_energy$DC.Unit)

# Harmonize only GE values and overwrite DC.Value
Gross_energy <- Gross_energy %>%
  mutate(
    DC.Value = case_when(
      DC.Unit %in% c("MJ/kg", "MJ/kg DM") ~ as.numeric(DC.Value),
      DC.Unit == "kcal/kg" ~ as.numeric(DC.Value) / 239,
      DC.Unit == "kcal/kg DM" ~ as.numeric(DC.Value) / 239,
      DC.Unit == "kcal/g" ~ as.numeric(DC.Value) * 1000 / 239,
      DC.Unit == "kcal/100g" ~ as.numeric(DC.Value) * 10 / 239,
      DC.Unit == "cal/g" ~ as.numeric(DC.Value) / 239,  # new
      DC.Unit == "MJ/g" ~ as.numeric(DC.Value) * 1000,  # new
      DC.Unit == "Mcal/kg" ~ as.numeric(DC.Value) * 4.184,  # new
      DC.Unit %in% c("kJ/g", "kJ/100g", "kJ/100g DM","KJ/100 g DM") ~ as.numeric(DC.Value) / 100,  # already partly handled
      DC.Unit == "kJ/kg DM" ~ as.numeric(DC.Value) / 1000,
      DC.Unit == "KJ" ~ as.numeric(DC.Value) / 1000,
      DC.Unit == "J/mg" ~ as.numeric(DC.Value) / 1e6,
      DC.Unit == "MJ/100g" ~ as.numeric(DC.Value) * 10,
      TRUE ~ NA_real_
    ),
    DC.Unit = ifelse(!is.na(DC.Value), "MJ/kg", DC.Unit)
  )




unique(Gross_energy$DC.Unit)

Gross_energy <- Gross_energy %>%
  # STEP 1: Create GE_source from nutrition_source
  mutate(
    GE_source = case_when(
      nutrition_source == "feedipedia" ~ "feedipedia",
      nutrition_source == "inferred_from_same_ingredient" ~ "inferred_from_same_ingredient",
      nutrition_source == "raw data" ~ "raw data",
      TRUE ~ NA_character_
    )
  )%>%
  select(-"nutrition_source")



```
g/kg	❌ Not an energy unit	Likely a mistake, investigate or exclude
% / % DM / % as fed	❌ Not energy units	Same as above
Unspecified	❌ Ambiguous	Investigate or exclude

```{r,message=FALSE,warning=FALSE}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```


## 13) Estimate gross energy for ingredients from proximate nutritional composition

To estimate Gross Energy (GE) when direct measurements are not available, we use a predictive equation based on proximate nutritional components: **crude protein (CP)**, **ether extract (EE)**, and **ash**. This method provides an accurate approximation of the energy content of feed ingredients based on their macronutrient composition.

The equation used is adapted from **Weiss et al. (2019)** and is expressed as:

```
GE (Mcal/kg) = (0.056 × CP) + (0.094 × EE) + [0.042 × (100 − CP − EE − Ash)]
GE (MJ/kg) = GE (Mcal/kg) × 4.184
```


```{r}
# --- STEP 1: Prepare predictors (CP, EE, Ash) as % of DM ---
# Keep relevant metadata for later
metadata_cols <- merged_metadata$Animals.Diet.Comp %>%
  select(B.Code, D.Item, is_entire_diet, Feedipedia) %>%
  distinct()

ge_predictors <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable %in% c("CP", "EE", "Ash")) %>%
  mutate(
    DC.Unit = trimws(tolower(DC.Unit)),
    DC.Value = as.numeric(DC.Value),

    # Harmonize to % of DM
    DC.Value = case_when(
      DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value,
      DC.Unit %in% c("g", "g/kg", "g/kg dm")                         ~ DC.Value / 10,
      DC.Unit %in% c("kg", "kg/day")                                ~ DC.Value * 100,
      TRUE ~ NA_real_
    ),
    DC.Unit = "%"
  ) %>%
  group_by(B.Code, D.Item, DC.Variable) %>%
  summarise(DC.Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = DC.Value) %>%
  left_join(metadata_cols, by = c("B.Code", "D.Item"))

# --- STEP 2: Apply Weiss GE equation only where CP, EE, Ash are all present ---
ge_predicted <- ge_predictors %>%
  mutate(
    GE_Mcal_per_kg = ifelse(
      !is.na(CP) & !is.na(EE) & !is.na(Ash),
      (CP * 0.056) + (EE * 0.094) + ((100 - CP - EE - Ash) * 0.042),
      NA_real_
    ),
    GE_MJ_per_kg = GE_Mcal_per_kg * 4.184
  ) %>%
  select(B.Code, D.Item, GE_MJ_per_kg, is_entire_diet, Feedipedia)


# --- STEP 3: Merge and only impute where GE is missing ---
Gross_energy <- Gross_energy %>%
  
  # STEP 2: Join GE predictions (e.g. from Weiss)
  full_join(ge_predicted, by = c("B.Code", "D.Item","is_entire_diet","Feedipedia")) %>%
  
  # STEP 3: Only update GE_source if it is still missing
  mutate(
    GE_source = case_when(
      is.na(GE_source) & !is.na(GE_MJ_per_kg) & is.na(DC.Value) ~ "predicted_from_Weiss_GE",
      TRUE ~ GE_source
    ),
    DC.Value = if_else(is.na(DC.Value), GE_MJ_per_kg, DC.Value),
    DC.Unit = "MJ/kg",
    DC.Variable = "GE"
  ) %>%
  select(-GE_MJ_per_kg)




```


## 14) Infer gross energy from data 
As we did in 9) we are inferring again GE data after estimation by proximal nutritional composition. 

```{r}
ingredient_means <- Gross_energy %>%
  filter(is_entire_diet == FALSE, !is.na(DC.Value)) %>%
  group_by(D.Item, DC.Variable) %>%
  summarise(mean_value = mean(DC.Value, na.rm = TRUE), .groups = "drop")

# --- Fill Missing Values with Mean ---
Gross_energy <- Gross_energy %>%
  left_join(ingredient_means, by = c("D.Item", "DC.Variable")) %>%
  mutate(
    GE_source = case_when(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value) ~ "inferred_from_same_ingredient-Step2",
      TRUE ~ GE_source
    ),
    DC.Value = if_else(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value),
      mean_value,
      DC.Value
    )
  ) %>%
  select(-mean_value)
```


## 15) Estimate Digestible energy from procimate nutritional composition

```{r}
# --- STEP 4: Prepare predictors for Weiss DE equation ---
# --- STEP 1: Store relevant metadata before pivot ---
de_metadata_cols <- merged_metadata$Animals.Diet.Comp %>%
  select(B.Code, D.Item, is_entire_diet, Feedipedia) %>%
  distinct()

# --- STEP 2: Prepare DE predictors and harmonize units ---
de_predictors <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable %in% c("CP", "EE", "Ash", "NDF")) %>%
  mutate(
    DC.Unit = trimws(tolower(DC.Unit)),
    DC.Value = as.numeric(DC.Value),

    # Harmonize to g/kg DM
    DC.Value = case_when(
      DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value * 10,
      DC.Unit %in% c("g", "g/kg", "g/kg dm")                         ~ DC.Value,
      DC.Unit %in% c("kg", "kg/day")                                ~ DC.Value * 1000,
      TRUE ~ NA_real_
    ),
    DC.Unit = "g/kg"
  ) %>%
  group_by(B.Code, D.Item, DC.Variable) %>%
  summarise(DC.Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = DC.Value) %>%
  left_join(de_metadata_cols, by = c("B.Code", "D.Item"))

# --- STEP 3: Apply Weiss DE equation and convert to GE ---
de_predicted <- de_predictors %>%
  mutate(
    DE_Mcal_per_kg = ifelse(
      !is.na(CP) & !is.na(EE) & !is.na(Ash) & !is.na(NDF),
      1.044 + 0.0101 * CP + 0.0161 * EE - 0.0037 * Ash - 0.0023 * NDF,
      NA_real_
    ),
    DE_MJ_per_kg = DE_Mcal_per_kg * 4.184,
    GE_MJ_per_kg = ifelse(!is.na(DE_MJ_per_kg), DE_MJ_per_kg / 0.54, NA_real_)
  ) %>%
  select(B.Code, D.Item, GE_MJ_per_kg, is_entire_diet, Feedipedia)


# --- STEP 6: Fill remaining missing GE using DE-based predictions only if GE is still missing ---
Gross_energy <- full_join(Gross_energy, de_predicted, by = c("B.Code", "D.Item", "is_entire_diet", "Feedipedia")) %>%
  mutate(
    DC.Value = if_else(is.na(DC.Value) & !is.na(GE_MJ_per_kg), GE_MJ_per_kg, DC.Value),
    GE_source = case_when(
      is.na(GE_source) & is.na(DC.Value) & !is.na(GE_MJ_per_kg) ~ "predicted_from_Weiss_DE",
      TRUE ~ GE_source  # keep existing label (e.g., feedipedia, raw data, inferred, or missing)
    ),
    DC.Unit = "MJ/kg",
    DC.Variable = "GE"
  ) %>%
  select(-GE_MJ_per_kg)


```

```{r,message=FALSE,warning=FALSE}
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```

We dont get more papers seems logical because probably the same ingredient as the equation  for GE. 

## 16) Convert digestible energy into gross energy 
Converts digestible energy (DE) to gross energy (GE) using a default digestibility factor (54%) when DE is reported but GE is not.

```{r,message=FALSE,warning=FALSE}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
DE_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "DE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
DE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(DE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_DE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(DE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```

Most of the DE data is coming from feediepdia which is expressed as a % of DE so we can't convert that data back using the IPCC method. 

```{r,message=FALSE,warning=FALSE}

# # Define digestibility conversion factor (Song et al., 2025)
# default_DE_fraction <- 0.54  # 54%
# 
# # 1. Extract DE rows from merged_metadata
# de_data <- merged_metadata$Animals.Diet.Digest %>%
#   filter(DD.Variable == "DE", !is.na(DD.Value) & DD.Value != "") %>%
#   mutate(
#     DD.Value = as.numeric(DD.Value),
#     DD.Unit = trimws(tolower(DD.Unit)),
#     DD.Value = case_when(
#       DD.Unit %in% c("mj/kg", "mj/kg dm") ~ DD.Value,
#       DD.Unit == "mcal/kg" ~ DD.Value * 4.184,
#       DD.Unit == "kcal/kg" ~ DD.Value / 239,
#       DD.Unit == "kcal/kg dm" ~ DD.Value / 239,
#       DD.Unit == "kcal/100g" ~ (DD.Value * 10) / 239,
#       DD.Unit == "kj/g" ~ DD.Value / 1000,
#       DD.Unit == "kj/100g" ~ DD.Value / 100,
#       TRUE ~ NA_real_
#     ),
#     DD.Unit = "MJ/kg"
#   ) %>%
#   mutate(
#     GE_from_DE = DD.Value / default_DE_fraction
#   ) %>%
#   select(B.Code, D.Item, GE_from_DE)
# 
# # 2. Merge GE_from_DE into your existing Gross_energy dataset
# Gross_energy <- Gross_energy %>%
#   left_join(de_data, by = c("B.Code", "D.Item")) %>%
#   mutate(
#     # Flag only when GE was previously missing and DE is available
#     GE_source = case_when(
#       is.na(GE_source) & is.na(DC.Value) & !is.na(GE_from_DE) ~ "inferred_from_DE",
#       TRUE ~ GE_source
#     ),
#     # Fill in missing GE values with GE_from_DE
#     DC.Value = if_else(is.na(DC.Value), GE_from_DE, DC.Value),
#     DC.Unit = "MJ/kg",
#     DC.Variable = "GE"
#   ) %>%
#   select(-GE_from_DE)


```

```{r,message=FALSE,warning=FALSE}
# 1. Total studies per species
species_total <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product)

# 2. GE: Only B.Codes where DC.Variable == GE and DC.Value is NOT NA
ge_bcodes <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value)) %>%
  distinct(B.Code) %>%
  pull()

# 3. Feed: Only B.Codes where D.Amount is NOT NA
feed_amount_bcodes <- ingredients %>%
  filter(!is.na(D.Amount)) %>%
  distinct(B.Code) %>%
  pull()

# 4. B.Codes with BOTH GE value AND feed amount
both_bcodes <- intersect(ge_bcodes, feed_amount_bcodes)

# 5. Build final summary table by species
final_summary <- species_total %>%
  mutate(
    Has_GE = B.Code %in% ge_bcodes,
    Has_FEED_or_Amount = B.Code %in% feed_amount_bcodes,
    Has_BOTH = B.Code %in% both_bcodes
  ) %>%
  group_by(P.Product) %>%
  summarise(
    Total_Studies = n_distinct(B.Code),
    Studies_with_GE = sum(Has_GE),
    Studies_with_FEED_or_Amount = sum(Has_FEED_or_Amount),
    Studies_with_BOTH = sum(Has_BOTH)
  ) %>%
  arrange(P.Product)

# Show summary
knitr::kable(final_summary, caption = "Final Summary (with actual GE and Feed Amount values)")

# Optional: List the actual B.Codes with both
bcode_list_both <- species_total %>%
  filter(B.Code %in% both_bcodes) %>%
  arrange(P.Product, B.Code)

# Show the B.Codes table
knitr::kable(bcode_list_both, caption = "B.Codes with BOTH GE (DC.Value) and Feed Intake (D.Amount)")


```



## 17) Calculation of CH4 emissions


### 17.1) Reconstruct Gross Energy at the diet level from ingredient data 
Reconstructs the gross energy of the entire diet from the weighted average of GE values of individual ingredients, when diet-level GE is not reported.

```{r,message=FALSE,warning=FALSE}

# --- Calculate GE_diet (MJ/kg) from ingredient GE and amount ---

# 1. Join GE values to ingredient amounts
# --- Calculate GE_diet (MJ/kg) ONLY if all ingredients have GE ---

# Step 1: Create unique diet ID
ingredients_with_GE <- ingredients %>%
  #filter(!is.na(D.Amount)) %>%
  mutate(Diet_ID = paste(B.Code, A.Level.Name, sep = "__")) %>%
  left_join(
    Gross_energy %>%
      filter(!is_entire_diet) %>%
      select(B.Code, D.Item, DC.Value),
    by = c("B.Code", "D.Item")
  ) %>%
  mutate(
    D.Amount_kg = case_when(
      D.Unit.Amount == "g" ~ D.Amount / 1000,
      D.Unit.Amount == "kg" ~ D.Amount,
      TRUE ~ NA_real_
    )
  )

# Step 2: Count number of ingredients and number with GE per diet
ingredient_counts <- ingredients_with_GE %>%
  group_by(Diet_ID) %>%
  summarise(
    n_total = n(),
    n_with_GE = sum(!is.na(DC.Value)),
    .groups = "drop"
  ) %>%
  filter(n_total == n_with_GE)  # keep only complete GE diets

# Step 3: Join back to keep only those complete diets
GE_from_ingredients <- ingredients_with_GE %>%
  semi_join(ingredient_counts, by = "Diet_ID") %>%
  group_by(Diet_ID, B.Code, A.Level.Name) %>%
  summarise(
    GE_diet = sum(DC.Value * D.Amount_kg) / sum(D.Amount_kg),
    .groups = "drop"
  )


# --- Use GE_diet from ingredients when GE for entire diet is missing ---
GE_diet_reconstructed <- GE_from_ingredients %>%
  # Join intake amount (total) and units
  left_join(
    total_amounts %>% select(B.Code, A.Level.Name, Total_Amount, D.Unit.Amount),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  # Add species (P.Product)
  left_join(
    merged_metadata$Prod.Out %>% select(B.Code, P.Product),
    by = "B.Code"
  ) %>%
  # Only keep ruminants
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  mutate(
    Total_Amount_kg = case_when(
      D.Unit.Amount == "g" ~ Total_Amount / 1000,
      D.Unit.Amount == "kg" ~ Total_Amount,
      TRUE ~ NA_real_
    ),
    
    Ym = case_when(
      P.Product == "Cattle" ~ 0.065,
      P.Product %in% c("Sheep", "Goat") ~ 0.060,
      TRUE ~ NA_real_
    ),
    
    GE_daily = ifelse(!is.na(GE_diet) & !is.na(Total_Amount_kg),
                      GE_diet * Total_Amount_kg,
                      NA_real_),
    
    CH4_kg_year = ifelse(!is.na(GE_daily) & !is.na(Ym),
                         GE_daily * Ym * 365 / 55.65,
                         NA_real_),
    
    CH4_flag = case_when(
      is.na(GE_diet) ~ "missing_GE_diet",
      is.na(Total_Amount_kg) ~ "missing_amount",
      is.na(Ym) ~ "missing_Ym",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      TRUE ~ "GE_diet_based"
    ),
    
    source = "GE_diet_from_ingredients"
  ) %>%
  rename(D.Item = A.Level.Name) %>%
  select(B.Code, D.Item, P.Product, GE_daily, CH4_kg_year, Total_Amount = Total_Amount_kg,
         D.Unit.Amount, ED.Mean.T = Total_Amount_kg, CH4_flag, source)


```


### 17.2) Gross energy at the entire diet level 
```{r,message=FALSE,warning=FALSE}


# Fix join to preserve D.Unit.Amount for further use
Gross_energy_entire <- Gross_energy %>%
  filter(is_entire_diet == TRUE) %>%
  left_join(
    total_amounts %>% select(B.Code, A.Level.Name, Total_Amount, D.Unit.Amount),
    by = c("B.Code" = "B.Code", "D.Item" = "A.Level.Name")
  ) %>%
  mutate(
    # Convert Total_Amount and update unit if needed
    Total_Amount = case_when(
      D.Unit.Amount == "g" ~ Total_Amount / 1000,
      D.Unit.Amount == "kg" ~ Total_Amount,
      TRUE ~ NA_real_
    ),
    D.Unit.Amount = case_when(
      D.Unit.Amount == "g" ~ "kg",
      TRUE ~ D.Unit.Amount
    ),
    reason_missing = case_when(
      is.na(Total_Amount) & D.Unit.Amount %in% c("g/kg", "%", "% Diet") ~ "Cannot convert unit to kg",
      TRUE ~ NA_character_
    ),
    GE_daily = ifelse(
      !is.na(Total_Amount) & !is.na(DC.Value),
      DC.Value * Total_Amount,
      NA_real_
    )
  )




```

### 17.3) Merge GE data with feed intake 

```{r,message=FALSE,warning=FALSE}


  

# 2. Merge and compute GE and CH4
GE_all <- Gross_energy_entire %>%
  # Add species info
  left_join(merged_metadata$`Prod.Out` %>% select(B.Code, P.Product), by = "B.Code") %>%
  # Keep only ruminants
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  # Merge feed intake data
  left_join(feed_intake_diet %>% 
              select(B.Code, A.Level.Name, ED.Mean.T),
            by = c("B.Code", "D.Item" = "A.Level.Name")) %>%
  # Overwrite Total_Amount with ED.Mean.T when available
  mutate(
    Total_Amount = if_else(!is.na(ED.Mean.T), ED.Mean.T, Total_Amount),
    D.Unit.Amount = if_else(!is.na(ED.Mean.T), "kg", D.Unit.Amount),

    Ym = case_when(
      P.Product == "Cattle" ~ 0.065,
      P.Product %in% c("Sheep", "Goat") ~ 0.060,
      TRUE ~ NA_real_
    ),

    GE_daily = ifelse(!is.na(DC.Value) & !is.na(Total_Amount),
                      DC.Value * Total_Amount,
                      NA_real_),

    CH4_kg_year = ifelse(!is.na(GE_daily) & !is.na(Ym),
                         GE_daily * Ym * 365 / 55.65,
                         NA_real_),

    CH4_flag = case_when(
      is.na(DC.Value) ~ "missing_GE_value",
      is.na(Ym) ~ "missing_Ym",
      is.na(Total_Amount) ~ "missing_amount",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      !is.na(ED.Mean.T) ~ "amount_inferred from feed intake",
      TRUE ~ "ok"
    )
  )



```

```{r,message=FALSE,warning=FALSE}
###############################################################################
## SINGLE CHUNK: Sum Ingredient Intake & Merge into Entire-Diet GE + CH₄ Calc ##
###############################################################################

# 1) Sum ingredient-level feed intake -----------------------------------------


ingredient_intake_summed <- feed_intake_ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    sum_ingredient_intake = sum(ED.Mean.T, na.rm = TRUE),
    .groups = "drop"
  )

# 2) Merge this ingredient sum into the entire-diet GE table ------------------
#    Then calculate final intake, daily GE, and CH₄.
GE_all_reconstructed <- Gross_energy_entire %>%
  # Add species info and keep only ruminants
  left_join(
    merged_metadata$Prod.Out %>%
      select(B.Code, P.Product),
    by = "B.Code"
  ) %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  
  # Merge the ingredient-level summed intake (aliasing A.Level.Name to D.Item)
  left_join(
    ingredient_intake_summed,
    by = c("B.Code", "D.Item" = "A.Level.Name")
  ) %>%
  
  # Compute CH₄-related fields
  mutate(
    # Methane conversion factor by species
    Ym = case_when(
      P.Product == "Cattle" ~ 0.065,
      P.Product %in% c("Sheep", "Goat") ~ 0.060,
      TRUE ~ NA_real_
    ),

    # Final daily intake (kg/day): pick the best available, but avoid using DC.Value
    final_intake_kg_day = case_when(
      !is.na(Total_Amount) & D.Unit.Amount == "kg" ~ Total_Amount,
      !is.na(sum_ingredient_intake) ~ sum_ingredient_intake,
      TRUE ~ NA_real_
    ),
    
    # Track source of intake
    final_intake_source = case_when(
      !is.na(Total_Amount) & D.Unit.Amount == "kg" ~ "Total_Amount",
      !is.na(sum_ingredient_intake) ~ "Total feed intake",
      TRUE ~ "Missing"
    ),
    
    # Daily GE intake (MJ/day) = GE content × daily intake
    GE_daily = case_when(
      !is.na(final_intake_kg_day) & !is.na(DC.Value) ~ DC.Value * final_intake_kg_day,
      TRUE ~ NA_real_
    ),
    
    # Annual CH₄ emissions
    CH4_kg_year = ifelse(
      !is.na(GE_daily) & !is.na(Ym),
      GE_daily * Ym * 365 / 55.65,
      NA_real_
    ),
    
    # Flag potential issues
    CH4_flag = case_when(
      is.na(DC.Value) ~ "missing_GE_value",
      is.na(Ym) ~ "missing_Ym",
      is.na(final_intake_kg_day) ~ "missing_final_intake",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      TRUE ~ "ok"
    )
  )


```

### 17.4) Final calculation

```{r,message=FALSE,warning=FALSE}
#combine
# 1. Tag and select relevant columns from both tables
GE_all_clean <- GE_all %>%
  mutate(source = "GE_all") %>%
  select(
    B.Code, 
    D.Item, 
    P.Product, 
    GE_daily, 
    CH4_kg_year,
    Total_Amount,
    D.Unit.Amount,
    ED.Mean.T, 
    CH4_flag, 
    source
  )

GE_reconstructed_clean <- GE_all_reconstructed %>%
  mutate(source = "Reconstructed") %>%
  rename(ED.Mean.T = final_intake_kg_day) %>%
  select(
    B.Code, 
    D.Item, 
    P.Product, 
    GE_daily, 
    CH4_kg_year, 
    Total_Amount,
    D.Unit.Amount,
    ED.Mean.T, 
    CH4_flag, 
    source
  )

GE_combined <- bind_rows(GE_all_clean, GE_reconstructed_clean, GE_diet_reconstructed) %>%
  mutate(
    source_priority = case_when(
      source == "GE_all" ~ 1,
      source == "Reconstructed" ~ 2,
      source == "GE_diet_from_ingredients" ~ 3,
      TRUE ~ 4
    ),
    has_CH4 = !is.na(CH4_kg_year)
  ) %>%
  arrange(B.Code, D.Item, desc(has_CH4), source_priority) %>%
  distinct(B.Code, D.Item, .keep_all = TRUE) %>%
  select(-has_CH4, -source_priority)

GE_data <- GE_combined %>%
  filter(!is.na(GE_daily) & GE_daily != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

# Count unique B.Code with non-missing CH4 emissions
n_CH4_codes <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code) %>%
  nrow()

n_CH4_codes

n_CH4_codes_by_product <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "n_CH4_codes")

n_CH4_codes_by_product
```


```{r,message=FALSE,warning=FALSE}
# Step 1: Get list of B.Codes from GE_combined
relevant_BCodes <- GE_combined %>%
  distinct(B.Code)

# Step 2: Filter Animals.Diet.Comp for relevant B.Codes
diet_data_filtered <- merged_metadata$Animals.Diet %>%
  semi_join(relevant_BCodes, by = "B.Code")

# Step 3: Count unique diets and ingredients
total_diets <- diet_data_filtered %>%
  filter(!is.na(A.Level.Name)) %>%
  distinct(B.Code, A.Level.Name) %>%
  nrow()

total_ingredients <- diet_data_filtered %>%
  filter(!is.na(D.Item)) %>%
  distinct(B.Code, D.Item) %>%
  nrow()

# Step 4: Output
cat("Total unique diets:", total_diets, "\n")
cat("Total unique ingredients:", total_ingredients, "\n")

```



## 18) logical checks 
Performs basic outlier checks on feed intake data, identifying suspiciously high or low intake values based on predefined unit thresholds.

```{r,message=FALSE,warning=FALSE}

# Assuming your data is stored in a data frame named `feed_intake_data`

# Filter for units that imply kg/day/individual and check if ED.Mean.T > 50
# Identify suspicious feed intake values
suspicious_feed_intake <- feed_intake %>%
  filter(
    # Suspiciously high intake in kg/day/individual
    (
      Out.Unit.Amount %in% c("kg", "kg dm") & 
      ED.Mean.T > 50 &
      Out.Unit.Time == "day" &
      Out.Unit.Animals == "individual"
    ) |
    
    # Suspiciously low intake in g or g dm (regardless of time/animal units)
    (
      Out.Unit.Amount %in% c("g", "g dm") &
      ED.Mean.T >= 1 & ED.Mean.T <= 20
    )
  ) %>%
  select(B.Code, A.Level.Name, ED.Mean.T, Out.Unit.Amount, Out.Unit.Time, Out.Unit.Animals)

# View suspicious entries
print(suspicious_feed_intake)


```

## 19) Link emissions to productivity data 

here we will have the same issue as with feed intake data since the Alevel in the data tab dont always latch the ones in the diet tab 

```{r}
yield_data<-merged_metadata$Data.Out%>%
  filter(Out.Subind %in% c("Meat Yield","Milk Yield") )
```


