---
title: "Livestock diet items"
author:
  - name: "Lolita Muller"
    email: "m.lolita@cgiar.org"
    orcid: "0009-0002-8345-7219"
  - name: "Peter Steward"
    email: "p.steward@cgiar.org"
    orcid: "0000-0003-3985-4911"
  - name: "Namita Joshi"
    email: "n.joshi@cgiar.org"
    orcid: "0000-0002-1623-6910"
  - name: "Todd Rosenstock"
    email: "t.rosenstock@cgiar.org"
    orcid: "0000-0002-1958-9500"
date: "2025-02-19"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show

---

## Introduction


## 1) Set up 
This section installs and loads required packages, connects to the S3 bucket containing ERA data, and loads both the livestock and camel metadata into memory.

```{r,message=FALSE,warning=FALSE}
# Install and load 'pacman' package manager if not already installed

library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)
library(readxl)
library(tidyr)
library(arrow)
library(ggplot2)
library(data.table)
library(purrr)
library(janitor)
library(ggalt)
# Set up connection to S3 bucket
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("skinny_cow_2022", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
livestock_metadata <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))



```


```{r,message=FALSE,warning=FALSE}
# Set up connection to S3 bucket for camel data 
s3 <- s3fs::S3FileSystem$new(anonymous = TRUE)
era_s3 <- "s3://digital-atlas/era"

# List the files in the s3 bucket
files <- s3$dir_ls(file.path(era_s3,"data"))

# Identify the latest skinny_cow_2022-YYYY-MM-DD.RData file
files <- tail(grep(".RData", grep("courageous_camel_2024", files, value=TRUE), value=TRUE), 1)

# Download to local if not already present
save_path <- file.path(getwd(), basename(files))
if(!file.exists(save_path)){
  s3$file_download(files, save_path, overwrite = TRUE)
}

# Load the data
camel <- miceadds::load.Rdata2(file=basename(save_path), path=dirname(save_path))


```


```{r}
missing_unit_feed_intake<- camel$Data.Out %>%
  select(B.Code, T.Name, ED.Intake.Item,
         ED.Mean.T, ED.Error, Out.Unit, Out.Subind,Out.Code.Joined) %>%
  filter(Out.Subind == "Feed Intake") %>%
  filter(is.na(Out.Unit))
```


```{r,message=FALSE,warning=FALSE}

##### 0. column templates ------------------------------------------------------
diet_cols   <- c("B.Code","A.Level.Name","D.Item","D.Type","D.Amount",
                 "D.Unit.Amount","D.Unit.Time","D.Unit.Animals",
                 "DC.Is.Dry","D.Ad.lib")
mt_cols     <- c("B.Code","A.Level.Name","T.Animals","T.Name","T.Control")
comp_cols   <- c("B.Code","D.Item","is_entire_diet",
                 "DC.Variable","DC.Value","DC.Unit")
digest_cols <- c("B.Code","D.Item","is_entire_diet",
                 "DD.Variable","DD.Value","DD.Unit")
data_cols   <- c("B.Code","A.Level.Name","Out.Code.Joined",
                 "ED.Intake.Item","ED.Intake.Item.Raw",
                 "ED.Mean.T","ED.Error","Out.Subind","Out.Unit",
                 "Out.WG.Start","Out.WG.Unit","is_entire_diet","T.Animals")
prod_cols   <- c("B.Code","P.Product")

##### 1. livestock tables (unchanged) ----------------------------------------
diet_livestock   <- livestock_metadata$Animals.Diet        [ , ..diet_cols ][, Source:="livestock"]
mt_livestock     <- livestock_metadata$MT.Out              [ , ..mt_cols   ][, Source:="livestock"]
comp_livestock   <- livestock_metadata$Animals.Diet.Comp   [ , ..comp_cols ][, Source:="livestock"]
digest_livestock <- livestock_metadata$Animals.Diet.Digest [ , ..digest_cols][, Source:="livestock"]
data_livestock   <- livestock_metadata$Data.Out            [ , ..data_cols ][, Source:="livestock"]
prod_livestock   <- livestock_metadata$Prod.Out            [ , ..prod_cols ][, Source:="livestock"]

##### 2. camel – build one key that links everything --------------------------
## 2·1  T-level information from MT.Out  (diet names + control flag)
diet_key <- camel$MT.Out[ , .(B.Code, T.Name, A.Level.Name, T.Control,Herd.Level.Name,Herd.Sublevel.Name)]

## 2·2  Animal numbers & start-weights from Herd.Out
herd_info <- camel$Herd.Out[ ,
   .(B.Code,
    Herd.Level.Name,
    Herd.Sublevel.Name,
    Herd.Sex,
    Herd.Stage,
     T.Animals    = Herd.N,
     Out.WG.Start = Herd.Start.Weight,
     Out.WG.Unit  = Herd.Start.Weight.Unit)
]

##### 3. camel tables ----------------------------------------------------------
## 3·1  Animals.Diet
diet_camel <- camel$Animal.Diet[ , ..diet_cols ][ , Source:="camel"]

## 3·2  MT.Out  (add T.Animals via herd_info)
mt_camel <- merge(diet_key,
                  herd_info,
                  by = c("B.Code","Herd.Level.Name","Herd.Sublevel.Name"),
                  all.x = TRUE)[, Source := "camel"]
        # ..mt_cols  **includes T.Animals**


## 3·3  Comp / Digest
comp_camel <- camel$Animal.Diet.Comp[ ,
       .(B.Code, D.Item, is_entire_diet,
         DC.Variable = DN.Variable,
         DC.Value    = DN.Value,
         DC.Unit     = DN.Unit)][, Source:="camel"]

digest_camel <- camel$Animal.Diet.Digest[ ,
       .(B.Code, D.Item, is_entire_diet,
         DD.Variable, DD.Value, DD.Unit)][, Source:="camel"]

## 3·4  Data.Out   (join diet_key → A.Level.Name, then herd_info)
data_camel <- merge(camel$Data.Out,                # brings in T.Name*
                    diet_key,                      # adds A.Level.Name & control
                    by = c("B.Code","T.Name"),
                    all.x = TRUE, sort = FALSE)

data_camel <- merge(data_camel, herd_info, by = c("B.Code","Herd.Level.Name","Herd.Sublevel.Name"),
                    all.x = TRUE, sort = FALSE)

data_camel <- data_camel[ ,
   .(B.Code, A.Level.Name,
     ED.Intake.Item,
     ED.Intake.Item.Raw = ED.Intake.Item,
     ED.Mean.T, ED.Error, Out.Unit, Out.Subind,
     Out.WG.Start, Out.WG.Unit,
     T.Animals)][, Source:="camel"]

## 3·5  product table
prod_camel <- camel$Herd.Out[ ,
   .(B.Code,
     P.Product = V.Product,
     Herd.Stage)][, Source:="camel"]

##### 4. bind camel + livestock ----------------------------------------------
diet_all   <- rbindlist(list(diet_livestock,   diet_camel  ), fill = TRUE)
mt_all     <- rbindlist(list(mt_livestock,     mt_camel    ), fill = TRUE)
comp_all   <- rbindlist(list(comp_livestock,   comp_camel  ), fill = TRUE)
digest_all <- rbindlist(list(digest_livestock, digest_camel), fill = TRUE)
data_all   <- rbindlist(list(data_livestock,   data_camel  ), fill = TRUE)
prod_all   <- rbindlist(list(prod_livestock,   prod_camel  ), fill = TRUE)

merged_metadata <- list(
  Animals.Diet        = diet_all,
  MT.Out              = mt_all,
  Animals.Diet.Comp   = comp_all,
  Animals.Diet.Digest = digest_all,
  Data.Out            = data_all,
  Prod.Out            = prod_all
)


```

```{r,message=FALSE,warning=FALSE}
# Clean up intermediate objects
rm(
  diet_cols, mt_cols, comp_cols, digest_cols, data_cols, prod_cols,
  diet_livestock, mt_livestock, comp_livestock, digest_livestock, data_livestock, prod_livestock,
  diet_camel, mt_camel, comp_camel, digest_camel, data_camel, prod_camel, herd_weight,comp_all,data_all,diet_all,digest_all,mt_all,prod_all,livestock_metadata,camel,herd_info,diet_key)

# (Optional) Run gc() to free memory
gc()
```
## 1.1 Correct T.Control

When t control is yes for one of the diets then automatically enter No for the other diets of the same paper. Also for missing t control if the name of the diet contains control or Con or Zero then automatically deduce this is a control diet and others are treatment diets. 

```{r}

MT <- copy(merged_metadata$MT.Out)


MT[, T.Control := tolower(trimws(as.character(T.Control))) ]


MT[
  ## diet name suggests a control …
  grepl("\\bzero\\b|\\bcon\\b|control", A.Level.Name, ignore.case = TRUE) &

  ## … but flag is still empty / ambiguous
  (is.na(T.Control) | T.Control %in% c("0", "false","na")),

  T.Control := "yes"
]



MT[, T.Control := {
  has_ctrl <- any(T.Control == "yes", na.rm = TRUE)          # does study have a control?
  
  if (has_ctrl) {
    # 4·a  normalise obvious “no” variants
    flag <- fifelse(
      T.Control %in% c("false", "0", "n", "no"), "no", T.Control
    )
    # 4·b  ensure exactly one category: yes / no
    fifelse(flag == "yes", "yes", "no")
    
  } else {
    # keep whatever was there (even "0", "false", NA, …)
    T.Control
  }
}, by = B.Code]


MT[, T.Control := fifelse(
  (is.na(T.Control) | T.Control %in% c("", "na")) & any(T.Control == "yes", na.rm = TRUE),
  "no",
  T.Control
), by = B.Code]


merged_metadata$MT.Out <- MT

```

## 1.2) Correct A.Level.Name
Some feed intake entries are not matched to any ingredient data due to discrepancies in A.Level.Name between the feed intake and ingredient datasets for the same study codes. In this section, we build a table to flag studies where diet names do not align, helping to identify cases that may require manual review or correction.


```{r}
# Step 1: Prepare cleaned input
ingredient_levels <- merged_metadata$Animals.Diet %>%
  distinct(B.Code, A.Level.Name) %>%
  mutate(Source = "Ingredient")

feedintake_levels <- merged_metadata$Data.Out %>%
  filter(Out.Subind == "Feed Intake", !is.na(A.Level.Name)) %>%
  distinct(B.Code, A.Level.Name) %>%
  mutate(Source = "FeedIntake")

# Step 2: Create mismatch table
combined_levels <- bind_rows(ingredient_levels, feedintake_levels)

comparison_table <- combined_levels %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    Sources = paste(sort(unique(Source)), collapse = " & "),
    .groups = "drop"
  ) %>%
  mutate(
    Match_Status = case_when(
      Sources == "FeedIntake & Ingredient" ~ "Match",
      Sources == "Ingredient" ~ "Only in Ingredient",
      Sources == "FeedIntake" ~ "Only in FeedIntake",
      TRUE ~ "Unknown"
    )
  )

# Step 3: Identify B.Codes with exactly 1 unmatched in both
to_correct <- comparison_table %>%
  filter(Match_Status %in% c("Only in Ingredient", "Only in FeedIntake")) %>%
  group_by(B.Code) %>%
  summarise(
    n_ing = sum(Match_Status == "Only in Ingredient"),
    n_feed = sum(Match_Status == "Only in FeedIntake"),
    Ingredient_Name = A.Level.Name[Match_Status == "Only in Ingredient"][1],
    FeedIntake_Name = A.Level.Name[Match_Status == "Only in FeedIntake"][1],
    .groups = "drop"
  ) %>%
  filter(n_ing == 1, n_feed == 1)

# Step 4: Apply the corrections to Data.Out (rename unmatched FeedIntake names)
merged_metadata$Data.Out <- merged_metadata$Data.Out %>%
  left_join(to_correct, by = c("B.Code", "A.Level.Name" = "FeedIntake_Name")) %>%
  mutate(
    A.Level.Name = coalesce(Ingredient_Name, A.Level.Name)
  ) %>%
  select(-Ingredient_Name)


rm(ingredient_levels,feedintake_levels,to_correct,combined_levels,comparison_table)

# Get merged stats
merged_stats <- list(
  papers = length(unique(merged_metadata$Animals.Diet$B.Code)),
  diets = length(unique(merged_metadata$Animals.Diet$A.Level.Name)),
  ingredients = length(unique(merged_metadata$Animals.Diet$D.Item))
)


```


## 1.3) Correct ingredient units 

Some ingredients appear multiple times with amounts reported in different units. In this section, we prioritize entries that are already in preferred units to minimize the need for conversions.

```{r,message=FALSE,warning=FALSE}
## keep ingredient that appear twice with different units only once 

# Define your unit preferences
preferred_units <- c("g", "kg")
preferred_time <- "day"
preferred_animals <- "individual"

# Prioritize and deduplicate
merged_metadata$Animals.Diet <- merged_metadata$Animals.Diet %>%
  # Add a priority score to sort by best rows
  mutate(
    priority = case_when(
      D.Unit.Amount %in% preferred_units &
        D.Unit.Time == preferred_time &
        D.Unit.Animals == preferred_animals ~ 1,
      TRUE ~ 2  # lower priority
    )
  ) %>%
  group_by(B.Code, A.Level.Name, D.Item) %>%
  slice_min(order_by = priority, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)  # optional: drop helper column

rm(preferred_units,preferred_time,preferred_animals)


### correct extraction errors 

merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(
    DC.Value = ifelse(
      B.Code == "BO1006" &
      DC.Variable == "Ash" &
      D.Item == "Ration 2" & 
      as.numeric(DC.Value) == 628,
      6.28,
      DC.Value
    )
  )


```

## 1.5) Correct D.Type

Some ingredients have more than one type, for those we merge the ingredients with the masterhseet to get the correct D.Type 
```{r}
# Step 1: Load master sheet
url <- "https://raw.githubusercontent.com/Mlolita26/ERA_Docs/main/era_master_sheet(7).xlsx"
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, destfile = temp_file, mode = "wb")

# Step 2: Read Ani diet and AOM sheets
ani_diet <- read_excel(temp_file, sheet = "ani_diet")
aom <- read_excel(temp_file, sheet = "AOM") %>% filter(L5 == "Feed Ingredient")

# Step 3: Clean Ani diet — rename columns and trim whitespace
ani_diet_clean <- ani_diet %>%
  rename_with(~ str_replace_all(., "\\s+", "_")) %>%
  mutate(across(everything(), ~ str_trim(as.character(.))))

# Step 4: Create long-format lookup for D.Item corrections
ditem_lookup <- ani_diet_clean %>% 
  select(D.Item.Root.Comp.Proc_Major, D.Item, D.Item.Root.Other.Comp.Proc_All) %>% 
  pivot_longer(cols = c(D.Item, D.Item.Root.Other.Comp.Proc_All),
               names_to = "Source_Field", values_to = "Name") %>% 
  filter(!is.na(Name), !is.na(D.Item.Root.Comp.Proc_Major)) %>% 
  separate_rows(Name, sep = "[,;]\\s*") %>%   # ← changed line
  mutate(Name = str_to_lower(str_trim(Name)),
         D.Item.Root.Comp.Proc_Major = str_trim(D.Item.Root.Comp.Proc_Major)) %>% 
  group_by(Name) %>% 
  slice(1) %>% 
  ungroup()


# Step 5: Apply correction to Animals.Diet using the lookup
animals_diet_updated <- merged_metadata$Animals.Diet %>%
  mutate(D.Item = str_to_lower(str_trim(D.Item))) %>%
  left_join(ditem_lookup, by = c("D.Item" = "Name")) %>%
  mutate(D.Item = coalesce(D.Item.Root.Comp.Proc_Major, D.Item)) %>%
  select(-D.Item.Root.Comp.Proc_Major)

# Step 6: Clean and prepare AOM mapping
aom_clean <- aom %>%
  rename_with(~ str_replace_all(., "\\s+", "_")) %>%
  mutate(
    Corrected_Type = ifelse(
      str_to_lower(L6) %in% c("other ingredients", "forage plants") & !is.na(L7),
      L7,                       # use L7 when rule matches
      L6                        # otherwise keep L6
    ),
    Edge_Value = str_trim(Edge_Value),
    Synonym    = str_trim(Synonym)
  )

# Step 7: Build mapping table from Edge_Value + Synonym
direct_match <- aom_clean %>%
  select(D.Item = Edge_Value, Corrected_Type) %>%
  filter(!is.na(D.Item))

synonym_match <- aom_clean %>%
  filter(!is.na(Synonym)) %>%
  separate_rows(Synonym, sep = ";\\s*") %>%
  mutate(Synonym = str_trim(Synonym)) %>%
  select(D.Item = Synonym, Corrected_Type)

aom_mapping <- bind_rows(direct_match, synonym_match) %>%
  mutate(D.Item = str_to_lower(str_trim(D.Item))) %>%
  distinct(D.Item, Corrected_Type)

# Step 8: Join to assign Corrected_Type → D.Type
animals_diet_updated <- animals_diet_updated %>%
  mutate(D.Item = str_to_lower(str_trim(D.Item))) %>%
  left_join(aom_mapping, by = "D.Item") %>%
  mutate(D.Type = ifelse(!is.na(Corrected_Type), Corrected_Type, D.Type)) %>%
  select(-Corrected_Type)

# # Step 9: Save back to merged metadata
merged_metadata$Animals.Diet <- animals_diet_updated
# 
# # Step 10: Optional check — items with no match
# check <- animals_diet_updated %>%
#   filter(is.na(Corrected_Type) & D.Type != "entire diet")
# 

```

also when same species so same in mastersheet they should have the same caterogy


those need to be categforized because there is nothing for d item 

```{r}
to_be_categorized<-merged_metadata$Animals.Diet%>%
  filter(is.na(D.Item))%>%
  filter(D.Type!="Entire Diet")

```


## 1.6) Summary of the data 
```{r,message=FALSE,warning=FALSE}

# Define species of interest
species_list <- c("Cattle", "Sheep", "Goat")

# Initialize result containers
species_summary <- list()
bcode_table <- list()

# Loop over each species
for (species in species_list) {
  
  # All B.Codes for this species
  species_bcodes <- merged_metadata$Prod.Out %>%
    filter(P.Product == species) %>%
    pull(B.Code) %>%
    unique()
  
  # Feed intake B.Codes
  feed_data <- merged_metadata$Data.Out %>%
    filter(B.Code %in% species_bcodes, !is.na(ED.Mean.T)) %>%
    filter(Out.Subind=="Feed Intake")%>%
    select(B.Code, D.Item = ED.Intake.Item)
  
  feed_bcodes <- unique(feed_data$B.Code)
  
  # GE B.Codes
  ge_data <- merged_metadata$Animals.Diet.Comp %>%
    filter(B.Code %in% species_bcodes, DC.Variable == "GE", !is.na(DC.Value)) %>%
    select(B.Code, D.Item)
  
  ge_bcodes <- unique(ge_data$B.Code)
  
  # Matched B.Codes (join on both B.Code and D.Item)
  matched_bcodes <- inner_join(feed_data, ge_data, by = c("B.Code", "D.Item")) %>%
    pull(B.Code) %>%
    unique()
  
  # Store species-level summary
  species_summary[[species]] <- tibble(
    Species = species,
    Total_BCodes = length(species_bcodes),
    With_FEED = length(feed_bcodes),
    With_GE = length(ge_bcodes),
    With_BOTH = length(matched_bcodes)
  )
  
  # Store long-format bcode list
  bcode_table[[species]] <- tibble(
    B.Code = c(species_bcodes, feed_bcodes, ge_bcodes, matched_bcodes),
    Category = rep(c("All", "With_FEED", "With_GE", "With_BOTH"), 
                   times = c(length(species_bcodes), length(feed_bcodes), length(ge_bcodes), length(matched_bcodes))),
    Species = species
  ) %>% distinct()
}

# Bind results
species_summary_df <- bind_rows(species_summary)
bcode_table_df <- bind_rows(bcode_table)

```



## 2) Harmonize units of ingredient amounts 
This step standardizes ingredient amount units (e.g., from mg, kg, %, etc.) to a consistent base (usually grams per individual per day), ensuring reliable downstream calculations and comparisons.

```{r,message=FALSE,warning=FALSE}
# diet ingredients: breakdown of all ingredients within each diet
ingredients <- merged_metadata$Animals.Diet %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    D.Unit.Time,
    D.Unit.Animals,
    DC.Is.Dry,
    D.Ad.lib
  ) %>%
  mutate(D.Amount = as.numeric(D.Amount))

# Function to harmonize diet ingredient units
harmonize_units <- function(df) {
  df %>%
    mutate(
      # Record original units (optional, for debugging/traceability)
      Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"),

      # Harmonize weight and volume units to a consistent base
      D.Amount = case_when(
        D.Unit.Amount %in% c("kg", "mg", "g/100g", "l", "kg/t", "g/300l", 
                             "mg/kg", "kg/100kg", "kg/100kg body weight", 
                             "kg/kg metabolic weight", "g/kg metabolic weight (0.75)", 
                             "g/kg DMI", "g/L", "g/kg DM") ~ {
          case_when(
            D.Unit.Amount == "kg" ~ D.Amount * 1000,                      # kg -> g
            D.Unit.Amount == "mg" ~ D.Amount / 1000,                     # mg -> g
            D.Unit.Amount == "g/100g" ~ D.Amount * 10,                   # g/100g -> g/kg
            D.Unit.Amount == "l" ~ D.Amount * 1000,                      # l -> ml
            D.Unit.Amount == "kg/t" ~ D.Amount * 1000,                   # kg/t -> g/kg
            D.Unit.Amount == "g/300l" ~ D.Amount / 300,                  # g/300l -> g/L
            D.Unit.Amount == "mg/kg" ~ D.Amount / 1000,                  # mg/kg -> g/kg
            D.Unit.Amount %in% c("kg/100kg", "kg/100kg body weight") ~ D.Amount * 10,     # -> g/kg body weight
            D.Unit.Amount == "kg/kg metabolic weight" ~ D.Amount * 1000, # -> g/kg metabolic weight
            D.Unit.Amount %in% c("g/kg body weight (0.75)", "g/kg Body Weight (0.75)") ~ D.Amount, # unchanged, just normalize name
            D.Unit.Amount == "g/kg DMI" ~ D.Amount,
            D.Unit.Amount == "g/kg DM" ~ D.Amount,
            TRUE ~ D.Amount
          )
        },
        TRUE ~ D.Amount
      ),

      # Update D.Unit.Amount to reflect new base
      D.Unit.Amount = case_when(
        D.Unit.Amount == "kg"  ~ "g",
        D.Unit.Amount == "mg"  ~ "g",
        D.Unit.Amount == "g/100g" ~ "g/kg",
        D.Unit.Amount == "l"  ~ "ml",
        D.Unit.Amount == "kg/t" ~ "g/kg",
        D.Unit.Amount == "g/300l" ~ "g/L",
        D.Unit.Amount == "mg/kg" ~ "g/kg",
        D.Unit.Amount %in% c("kg/100kg", "kg/100kg body weight") ~ "g/kg body weight",
        D.Unit.Amount %in% c("g/kg body weight (0.75)", "g/kg Body Weight (0.75)") ~ "g/kg metabolic weight",
        D.Unit.Amount == "kg/kg metabolic weight" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg metabolic weight (0.75)" ~ "g/kg metabolic weight",
        D.Unit.Amount == "g/kg DMI" ~ "g/kg",
        D.Unit.Amount == "g/kg DM" ~ "g/kg",
        D.Unit.Amount == "g/L" ~ "g/L",
        D.Unit.Amount == "g/kg Body Weight" ~ "g/kg body weight",  # case fix
        TRUE ~ D.Unit.Amount
      ),

      # Harmonize time units to day
      D.Amount = case_when(
        D.Unit.Time == "week" ~ D.Amount / 7,
        D.Unit.Time == "month" ~ D.Amount / 30,
        D.Unit.Time == "2 weeks" ~ D.Amount / 14,
        D.Unit.Time == "4 days interval" ~ D.Amount / 4,
        D.Unit.Time == "2x/day" ~ D.Amount * 2,
        TRUE ~ D.Amount
      ),
      D.Unit.Time = ifelse(!is.na(D.Unit.Time), "day", D.Unit.Time)
    ) %>%
    # Set unit fields to NA if amount is NA
    mutate(
      D.Unit.Amount   = ifelse(is.na(D.Amount), NA_character_, D.Unit.Amount),
      D.Unit.Time     = ifelse(is.na(D.Amount), NA_character_, D.Unit.Time),
      D.Unit.Animals  = ifelse(is.na(D.Amount), NA_character_, D.Unit.Animals)
    )
}



# Apply the harmonization
ingredients <- harmonize_units(ingredients)

# Load replicate info to adjust for "all animals in replicate"
reps <- merged_metadata$MT.Out %>%
  mutate(T.Animals = as.numeric(T.Animals)) %>%
  distinct(B.Code, A.Level.Name, .keep_all = TRUE)


ingredients <- ingredients %>%
  left_join(reps %>% dplyr::select(B.Code, A.Level.Name, T.Animals), by=c("B.Code","A.Level.Name")) %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Animals == "all animals in replicate" & !is.na(T.Animals) ~ D.Amount / T.Animals,
      TRUE ~ D.Amount
    ),
    D.Unit.Animals = ifelse(D.Unit.Animals == "all animals in replicate", "individual", D.Unit.Animals)
  )

# Refresh 'Units' column after these adjustments (optional)
ingredients <- ingredients %>%
  mutate(Units = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals, sep=";"))

rm(harmonize_units)

```

### 2.1) Suspicious dry amounts

This step identifies ingredients where DC.Is.Dry is marked as NA or unspecified. We cross-check these with moisture values from the composition table to infer whether the ingredient is likely dry or wet. We also flag inconsistencies between the ingredient name (e.g., containing "dried") and the recorded dry status.

```{r}
# --- 1. Classify ingredients from Moisture or DM values ---
dry_classified <- bind_rows(
  # Moisture-based rules
  merged_metadata$Animals.Diet.Comp %>%
    filter(str_to_lower(DC.Variable) %in% c("moisture", "moist")) %>%
    mutate(
      Predictor = "Moisture",
      Value = as.numeric(DC.Value),
      Unit = str_to_lower(DC.Unit),
      Dry_Prediction = case_when(
        Unit %in% c("%", "% dm", "g/100g") & Value < 15  ~ TRUE,
        Unit %in% c("%", "% dm", "g/100g") & Value > 50  ~ FALSE,
        Unit == "g/kg" & Value < 150                     ~ TRUE,
        Unit == "g/kg" & Value > 500                     ~ FALSE,
        TRUE ~ NA
      )
    ),
  # DM-based rules
  merged_metadata$Animals.Diet.Comp %>%
    filter(str_to_lower(DC.Variable) == "dm") %>%
    mutate(
      Predictor = "DM",
      Value = as.numeric(DC.Value),
      Unit = str_to_lower(DC.Unit),
      Dry_Prediction = case_when(
        Unit %in% c("%", "% dm", "g/100g") & Value > 85  ~ TRUE,
        Unit %in% c("%", "% dm", "g/100g") & Value < 50  ~ FALSE,
        Unit == "g/kg" & Value > 850                    ~ TRUE,
        Unit == "g/kg" & Value < 500                    ~ FALSE,
        TRUE ~ NA
      )
    )
) %>%
  select(B.Code, D.Item, Predictor, Dry_Prediction) %>%
  distinct() %>%
  group_by(B.Code, D.Item, Predictor) %>%
  summarise(Dry_Prediction = first(na.omit(Dry_Prediction)), .groups = "drop") %>%
  pivot_wider(names_from = Predictor, values_from = Dry_Prediction, names_prefix = "Dry_Prediction_")

# --- 2. Detect "dry" or "dried" in name ---
dry_name_match <- ingredients %>%
  mutate(
    name_lower = str_to_lower(D.Item),
    Dry_by_name = str_detect(
      name_lower,
      "\\bdry\\b|\\bdried\\b|\\bmeal\\b|\\bground\\b|\\bmilled\\b|\\bflour\\b"
    ) & !str_detect(name_lower, "fish meal boiled")
  ) %>%
  select(B.Code, A.Level.Name,D.Item, Dry_by_name)


extra_dry <- ingredients %>%
  filter(
    D.Type %in% c("Non-ERA Feed", "Other Feed","Other Ingredients","Supplement"),
    !str_detect(str_to_lower(D.Item), "water|oil")
  ) %>%
  select(B.Code,A.Level.Name, D.Item) 

# Remove A.Level.Name from extra_dry before combining
extra_dry <- extra_dry %>%
  select(B.Code, D.Item) %>%
  distinct() %>%
  mutate(DC.Is.Dry = "Yes")


# --- 3. Combine predictions ---
# Combine dry predictions from DM/moisture, name-based detection, and extra rules
dry_prediction_final <- bind_rows(
  dry_classified %>%
    mutate(Prediction_Source = "Moisture_or_DM") %>%
    filter(Dry_Prediction_Moisture == TRUE | Dry_Prediction_DM == TRUE) %>%
    select(B.Code, D.Item) %>%
    mutate(DC.Is.Dry = "Yes"),
  
  dry_name_match %>%
    filter(Dry_by_name) %>%
    select(B.Code, D.Item) %>%
    mutate(DC.Is.Dry = "Yes", Prediction_Source = "Name_match"),
  
  extra_dry %>%
    select(B.Code, D.Item) %>%
    mutate(DC.Is.Dry = "Yes", Prediction_Source = "Extra_rule")
) %>%
  distinct(B.Code, D.Item, .keep_all = TRUE)


# --- 4. Apply predictions to ingredients table (update only predicted dry) ---
# 1. Join on D.Item
ingredients <- ingredients %>%
  left_join(dry_prediction_final, by = c("B.Code", "D.Item"), suffix = c("", ".from_DItem"))

# 2. Join also on A.Level.Name
ingredients <- ingredients %>%
  left_join(
    dry_prediction_final %>%
      rename(A.Level.Name = D.Item),
    by = c("B.Code", "A.Level.Name"),
    suffix = c("", ".from_AName")
  )

# 3. Apply logic: keep existing "Yes", but if original is "No" or NA, check if any of the predictions is "Yes"
ingredients <- ingredients %>%
  mutate(
    DC.Is.Dry = case_when(
      (is.na(DC.Is.Dry) | str_to_lower(DC.Is.Dry) %in% c("no","unspecified")) &
        (DC.Is.Dry.from_DItem == "Yes" | DC.Is.Dry.from_AName == "Yes") ~ "Yes",
      TRUE ~ DC.Is.Dry
    )
  ) %>%
  select(-DC.Is.Dry.from_DItem, -DC.Is.Dry.from_AName)


#take out water 
ingredients<-ingredients%>%
filter(D.Item != "Water" | is.na(D.Item))


rm(dry_classified,dry_name_match,dry_prediction_final,extra_dry)

```


### 2.2) Filter data with acceptable units

We filter the dataset to retain only ingredients with acceptable units. Uncommon or unexpected unit formats are excluded to ensure consistency and simplify downstream processing

```{r,message=FALSE,warning=FALSE}


# Define a list of acceptable units
acceptable_units <- c(
  "g/kg", "g", "%", "ml", "L", "kg", "g/L",
  "g/kg metabolic weight", "% Diet", 
  "g/kg body weight", "ml/kg", "% Body Mass", 
  "% Concentrate", "g/kg body weight"  # leave only one if repeated
)


# Filter rows to keep only acceptable units
ingredients <- ingredients %>%
  filter(D.Unit.Amount %in% acceptable_units | is.na(D.Unit.Amount)) %>%
  mutate(D.Item = ifelse(is.na(D.Item), "Unspecified", D.Item))

#tracking paper loss because of harmnozied units
#total_studies_harmonized<- length(unique(ingredients_harmonized$B.Code))
#total_ingredients_harmonized<- length(unique(ingredients_harmonized$D.Item))
#total_diets_harmonized<- length(unique(ingredients_harmonized$A.Level.Name))

rm(acceptable_units)
```

### 2.3) Convert body weight units by linking animals weight to data
```{r}

# Pre-filter Data.Out to keep only the first non-NA weight per group
weights_unique <- merged_metadata$Data.Out %>%
  filter(Out.Subind == "Weight Gain") %>%
  mutate(
    Out.WG.Unit = tolower(Out.WG.Unit),
    Out.WG.Start = case_when(
      Out.WG.Unit == "g" ~ Out.WG.Start / 1000,
      Out.WG.Unit == "mg" ~ Out.WG.Start / 1e6,
      Out.WG.Unit %in% c("kg", "unspecified", NA) ~ Out.WG.Start,
      TRUE ~ NA_real_
    ),
    Out.WG.Unit = ifelse(!is.na(Out.WG.Start), "kg", NA_character_)
  ) %>%
  filter(!is.na(Out.WG.Start)) %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    Out.WG.Start = mean(Out.WG.Start, na.rm = TRUE),
    Out.WG.Unit = "kg",
    .groups = "drop"
  )



# Step 3: Join body weight and keep only ingredients with weight
ingredients <- ingredients %>%
  left_join(
    weights_unique,
    by = c("B.Code", "A.Level.Name")) 


# Step 4: Convert D.Amount to grams based on body weight and update unit accordingly
ingredients <- ingredients %>%
  mutate(
    # Recalculate D.Amount using body weight or metabolic weight, result in kg
    D.Amount = case_when(
      D.Unit.Amount == "g/kg body weight" & !is.na(Out.WG.Start) ~ (D.Amount * Out.WG.Start) / 1000,
      D.Unit.Amount == "g/kg metabolic weight" & !is.na(Out.WG.Start) ~ (D.Amount * Out.WG.Start^0.75) / 1000,
      D.Unit.Amount == "% Body Mass" & !is.na(Out.WG.Start) ~ ((D.Amount / 100) * Out.WG.Start),
      TRUE ~ D.Amount  # unchanged if not convertible
    ),

    # Update unit to "kg" if recalculated
    D.Unit.Amount = case_when(
      D.Unit.Amount %in% c("g/kg body weight", "g/kg metabolic weight", "% Body Mass") &
        !is.na(Out.WG.Start) ~ "kg",
      TRUE ~ D.Unit.Amount
    )
  )


```

### 2.4) Estimate amounts using feed intake

For ingredients reported in relative units such as %, % Diet, or g/kg, we can estimate their absolute amounts when total diet feed intake is available. This allows us to convert proportions into standardized quantities.


```{r}
missing_unit_feed_intake<- merged_metadata$Data.Out %>%
  select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw,
         ED.Mean.T, ED.Error, Out.Unit, Out.Subind,is_entire_diet,Source,Out.Code.Joined) %>%
  filter(Out.Subind == "Feed Intake") %>%
  filter(is.na(Out.Unit))
```


#### 2.4.1) Harmonization of feed intake data

```{r}
feed_intake <- merged_metadata$Data.Out %>%
  select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw,
         ED.Mean.T, ED.Error, Out.Unit, Out.Subind,is_entire_diet,T.Animals) %>%
  filter(Out.Subind == "Feed Intake") %>%
  filter(!is.na(ED.Mean.T))%>%
  mutate(
    Out.Unit = tolower(trimws(Out.Unit)),

    # Extract unit components
    Out.Unit.Animals = str_extract(Out.Unit, "\\b(individual|replicate|animal replicate)\\b"),
    Out.Unit.Time    = str_extract(Out.Unit, "\\b(day|d|yr|year)\\b"),
    Out.Unit.Amount  = Out.Unit %>%
      str_remove_all("\\b(individual|replicate|animal replicate)\\b") %>%
      str_remove_all("\\b(day|d|yr|year)\\b") %>%
      str_replace_all("/+", "/") %>%
      str_trim() %>%
      str_remove("/$"))


# Step 3: Convert replicate-level intake to individual-level when possible
feed_intake <- feed_intake %>%
  mutate(
    # Ensure numeric type
    ED.Mean.T = as.numeric(ED.Mean.T),
    T.Animals = as.numeric(T.Animals),

    # Convert intake and update unit
    ED.Mean.T = case_when(
      Out.Unit.Animals == "replicate" & !is.na(T.Animals) & T.Animals > 0 ~ ED.Mean.T / T.Animals,
      TRUE ~ ED.Mean.T
    ),
    Out.Unit.Animals = case_when(
      Out.Unit.Animals == "replicate" & !is.na(T.Animals) & T.Animals > 0 ~ "individual",
      TRUE ~ Out.Unit.Animals
    )
  )



# Define preferred units
preferred_units <- c("kg dm", "g dm")
preferred_time <- "day"
preferred_animals <- "individual"

# Create filtered table with prioritization
feed_intake <- feed_intake %>%
  mutate(
    priority = case_when(
      Out.Unit.Amount %in% preferred_units &
        Out.Unit.Time == preferred_time &
        Out.Unit.Animals == preferred_animals &
        !is.na(ED.Error) ~ 1,
      Out.Unit.Amount %in% preferred_units &
        Out.Unit.Time == preferred_time &
        Out.Unit.Animals == preferred_animals ~ 2,
      TRUE ~ 3
    )
  ) %>%
  group_by(B.Code, A.Level.Name, ED.Intake.Item) %>%
  slice_min(order_by = priority, with_ties = FALSE) %>%
  ungroup() %>%
  select(-priority)  # optional



### keep unique units of ingredients 


# Step 3: Join body weight
feed_intake <- feed_intake %>%
  left_join(
    weights_unique %>%
      select(B.Code, A.Level.Name, Out.WG.Start, Out.WG.Unit),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  group_by(A.Level.Name) %>%
  mutate(
    Out.WG.Start = ifelse(is.na(Out.WG.Start),
                          first(na.omit(Out.WG.Start)),
                          Out.WG.Start)
  ) %>%
  ungroup() %>%
  distinct()

# Step 4: Convert ED.Mean.T to kg/day
feed_intake <- feed_intake %>%
  mutate(
    Out.Unit = trimws(tolower(Out.Unit)),
    Out.Unit.Amount = trimws(str_remove(Out.Unit.Amount, "/$")),
    ED.Mean.T = case_when(
      Out.Unit.Amount %in% c("g", "g dm") ~ ED.Mean.T / 1000,
      Out.Unit.Amount %in% c("kg", "kg dm") ~ ED.Mean.T,
      Out.Unit.Amount == "g dm/100 g" ~ ED.Mean.T * 10 / 1000,
      Out.Unit.Amount == "g/week" ~ ED.Mean.T / 7 / 1000,
      Out.Unit.Amount == "g dm/kg metabolic weight" ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount == "g dm/kg body weight" ~ (ED.Mean.T * Out.WG.Start) / 1000,
      Out.Unit.Amount == "mg" ~ ED.Mean.T / 1e6,
      Out.Unit.Amount %in% c("g/kg body weight", "g dm/kg body weight") ~ (ED.Mean.T * Out.WG.Start) / 1000,
      Out.Unit.Amount %in% c("g/kg body weight (0.75)", "g dm/kg body weight (0.75)") ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount %in% c("g/kg metabolic weight","g dm/kg metabolic weight") ~ (ED.Mean.T * Out.WG.Start^0.75) / 1000,
      Out.Unit.Amount == "kg dm/kg metabolic weight" ~ ED.Mean.T * Out.WG.Start^0.75,
      Out.Unit.Amount %in% c("% body weight", "dm % body weight") ~ (ED.Mean.T / 100) * Out.WG.Start,
      Out.Unit.Amount == "g/g body weight" ~ (ED.Mean.T * Out.WG.Start) / 1000,
      TRUE ~ NA_real_
    ),
    Out.Unit.Amount = case_when(
      Out.Unit.Amount %in% c("g", "g dm", "g dm/100 g", "g/week", "g/kg body weight",
                             "g dm/kg body weight", "g/kg body weight (0.75)",
                             "g dm/kg body weight (0.75)", "g/kg metabolic weight","g dm/kg metabolic weight","g/g body weight",
                             "kg dm/kg metabolic weight", "% body weight", 
                             "dm % body weight", "kg", "kg dm", "mg") ~ "kg",
      TRUE ~ Out.Unit.Amount
    )
  ) %>%
  select(-Out.Unit)

feed_intake <- feed_intake %>%
  group_by(B.Code, A.Level.Name, ED.Intake.Item.Raw) %>%
  arrange(
    desc(!is.na(ED.Mean.T)), 
    desc(Out.Unit.Amount == "kg"),
    desc(Out.Unit.Animals == "individual"),
    desc(Out.Unit.Time == "day")
  ) %>%
  slice(1) %>%
  ungroup()
```


#### 2.4.2) Unit error check on feed intake
Performs basic outlier checks on feed intake data, identifying suspiciously high or low intake values based on predefined unit thresholds.

```{r,message=FALSE,warning=FALSE}

# Assuming your data is stored in a data frame named `feed_intake_data`

# Filter for units that imply kg/day/individual and check if ED.Mean.T > 50
# Identify suspicious feed intake values
## ---- detect_suspicious_feed_intake -----------------------------------------
# Thresholds
high_kg <- 50      # > 50 kg d-1 animal-1  ⟹ unrealistically high
low_kg  <- 0.1     # < 0.1 kg d-1 animal-1 ⟹ unrealistically low
low_g_min <- 1     # 1–20 g range unlikely for daily intake
low_g_max <- 20

suspicious_feed_intake <- feed_intake %>% 
  mutate(
    flag = case_when(
      Out.Unit.Amount %in% c("kg", "kg dm") &
      Out.Unit.Time   == "day"              &
      Out.Unit.Animals == "individual" & ED.Mean.T >  high_kg ~ "too_high_kg",

      Out.Unit.Amount %in% c("kg", "kg dm") &
      Out.Unit.Time   == "day"              &
      Out.Unit.Animals == "individual" & ED.Mean.T <  low_kg  ~ "too_low_kg",

      Out.Unit.Amount %in% c("g",  "g dm")  &
      ED.Mean.T >= low_g_min & ED.Mean.T <= low_g_max          ~ "implausibly_low_g",

      TRUE ~ NA_character_
    )
  ) %>% 
  filter(!is.na(flag)) %>% 
  select(B.Code, A.Level.Name, ED.Mean.T,
         Out.Unit.Amount, Out.Unit.Time, Out.Unit.Animals, flag)

print(suspicious_feed_intake)


low_intake_codes <- feed_intake %>% 
  filter(
    # (a) < 0.1 kg / day / individual
    (Out.Unit.Amount %in% c("kg", "kg dm") &
     Out.Unit.Time   == "day" &
     Out.Unit.Animals == "individual" &
     ED.Mean.T < 0.1) |
    
    # (b) 1–20 g reported in gram units
    (Out.Unit.Amount %in% c("g", "g dm") &
     ED.Mean.T >= 1 & ED.Mean.T <= 20)
  ) %>% 
  distinct(B.Code) %>%           # keep each code only once
  pull()                         # return as a plain character vector

low_intake_codes


```

Some papers were identified with errors. This section corrects those error if they are still present before the next update of the data. 


```{r}

# Step: Convert suspiciously high kg values that are likely mislabelled g
feed_intake <- feed_intake %>%
  mutate(
    Out.Unit.Amount = case_when(
      Out.Unit.Amount == "kg" & Out.Unit.Animals == "individual" & Out.Unit.Time == "day" & ED.Mean.T > 50 ~ "g",  # correct mislabeling
      TRUE ~ Out.Unit.Amount
    )
  )

```


#### 2.4.3) Use of feed intake for ingredient amounts

If feed intake data is there for the entire diet and that the ingredients in ingredients dataset are in % or % of diet or g/kg we can convert 

```{r}
# Step 5: Estimate ingredient absolute amounts for diets with "Entire Diet" intake
ingredients <- ingredients %>%
  # Join only feed intake values where ED.Intake.Item is "Entire Diet"
  left_join(
    feed_intake %>%
      filter(ED.Intake.Item == "Entire Diet") %>%
      select(B.Code, A.Level.Name, Intake_kg_day = ED.Mean.T),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    # Estimate absolute amount (in grams)
    D.Amount = case_when(
      D.Unit.Amount == "%" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "% Diet" & !is.na(Intake_kg_day) ~ (D.Amount / 100) * Intake_kg_day * 1000,
      D.Unit.Amount == "g/kg" & !is.na(Intake_kg_day) ~ D.Amount * Intake_kg_day,
      TRUE ~ D.Amount
    ),
    
    # Update unit only if we did a conversion
    D.Unit.Amount = case_when(
      D.Unit.Amount %in% c("%", "% Diet", "g/kg") & !is.na(Intake_kg_day) ~ "g",
      TRUE ~ D.Unit.Amount
    )
  ) %>%
  select(-Intake_kg_day)


```
## 3) Subset data

## 3.1) Take out basal diets 

We excluded the basal diets because, in most of the livestock trials in our database, the basal ration is offered ad libitum and therefore lacks two pieces of information that are essential for a fair, productivity-based comparison: (i) actual feed-intake measurements and (ii) corresponding enteric-methane estimates. Without intake data we cannot convert emissions—or, in many cases, even estimate them—on a per-kilogram-of-product basis, and retaining these records would artificially inflate the number of “treatments” while contributing no usable information to the efficiency analyses. By focusing on the control diet (with measured intake) and the formulated treatment diets, we keep a coherent set of observations in which every data point can be expressed as both absolute emissions and emission intensity, allowing robust, like-for-like comparisons across studies.

```{r,message=FALSE,warning=FALSE}

ingredients<-ingredients%>%
  filter(A.Level.Name!="Base")

```


## 3.1 Take out grazing experiments 

```{r}

species_by_bcode <- merged_metadata$Prod.Out %>%
  select(B.Code, P.Product) %>%
  distinct()

# 2. Papers per species before filtering
papers_before <- ingredients %>%
  select(B.Code) %>%
  distinct() %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_before")

# Step 1: Identify A.Level.Name (diet) that include Grazing
grazing_diets <- ingredients %>%
  filter(D.Type == "Grazing") %>%
  distinct(B.Code, A.Level.Name)

length(unique(grazing_diets$B.Code))

# Step 2: Count how many non-grazing diets exist per B.Code
nongrazing_counts <- ingredients %>%
  filter(!(D.Type == "Grazing")) %>%
  distinct(B.Code, A.Level.Name) %>%
  group_by(B.Code) %>%
  summarise(n_nongrazing = n(), .groups = "drop")

length(unique(nongrazing_counts$B.Code))


# Step 3: Keep only diets that are not marked as Grazing AND are in B.Codes with at least one non-grazing diet
ingredients <- ingredients %>%
  anti_join(grazing_diets, by = c("B.Code", "A.Level.Name")) %>%  # remove all diets that were grazing
  semi_join(nongrazing_counts, by = "B.Code")  # keep only papers with at least one non-grazing diet


species_by_bcode <- merged_metadata$Prod.Out %>%
  select(B.Code, P.Product) %>%
  distinct()

# 2. Papers per species before filtering
papers_after <- ingredients %>%
  select(B.Code) %>%
  distinct() %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_after")
```

## 4) Replace feed intake data where available

We prioritize feed intake data over the amounts reported in the methods section, as feed intake is the key variable needed for emission calculations. When feed intake data is unavailable, we fall back on the harmonized ingredient amounts extracted earlier.

### 4.1) Feed intake data at the diet level 

```{r,message=FALSE,warning=FALSE}
# Step 5: Remove feed intake entries that are actually full-diet labels
feed_intake_ingredients <- feed_intake %>%
  filter(str_trim(ED.Intake.Item) != str_trim(A.Level.Name) & 
         str_trim(ED.Intake.Item) != "Entire Diet") %>%
  filter(!is.na(ED.Mean.T))


# Step 6: Merge with ingredients and override with ED.Mean.T when available
# FIXED VERSION
# Merge with ingredients and override with ED.Mean.T when available
ingredients <- ingredients %>%
  left_join(
    feed_intake_ingredients %>%
      select(B.Code, A.Level.Name, ED.Intake.Item, ED.Mean.T, Out.Unit.Amount, Out.Unit.Animals, Out.Unit.Time),
    by = c("B.Code", "A.Level.Name", "D.Item" = "ED.Intake.Item")
  ) %>%
  mutate(
    from_feed_intake = !is.na(ED.Mean.T),
    D.Amount = if_else(from_feed_intake, ED.Mean.T, D.Amount),
    D.Unit.Amount = if_else(from_feed_intake & !is.na(Out.Unit.Amount), Out.Unit.Amount, D.Unit.Amount),
    D.Unit.Animals = if_else(from_feed_intake & !is.na(Out.Unit.Animals), Out.Unit.Animals, D.Unit.Animals),
    D.Unit.Time = if_else(from_feed_intake & !is.na(Out.Unit.Time), Out.Unit.Time, D.Unit.Time)
  ) %>%
  select(-ED.Mean.T, -Out.Unit.Amount, -Out.Unit.Animals, -Out.Unit.Time) %>%
  distinct()


# Step 8: Final conversion of unit to grams (g)
ingredients <- ingredients %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Amount == "kg" ~ D.Amount * 1000,
      TRUE ~ D.Amount
    ),
    D.Unit.Amount = case_when(
      D.Unit.Amount == "kg" ~ "g",
      TRUE ~ D.Unit.Amount
    )
  )


```


### 4.2) Feed intake at the diet level  

TO do : See BO1070 we want the full diet to be included from feed intake which is not the case now see feed intake data the goal would be to add rows in ingredient to add info at the entire diet level have a close look to make it the same format as entire diets already in ingredients 

```{r}
# Step: Extract feed intake entries that refer to the entire diet
feed_intake_diet <- feed_intake %>%
  filter(ED.Intake.Item == A.Level.Name | ED.Intake.Item == "Entire Diet")

# Step: Merge and overwrite ingredients using diet-level feed intake
ingredients <- ingredients %>%
  left_join(
    feed_intake_diet %>%
      select(B.Code, A.Level.Name, ED.Intake.Item, ED.Intake.Item.Raw, ED.Mean.T, Out.Unit.Amount, Out.Unit.Animals, Out.Unit.Time),
    by = c("B.Code", "A.Level.Name", "D.Type" = "ED.Intake.Item.Raw")
  ) %>%
  mutate(
    from_feed_intake2 = !is.na(ED.Mean.T),
    D.Amount = if_else(from_feed_intake2, ED.Mean.T, D.Amount),
    D.Unit.Amount = if_else(from_feed_intake2 & !is.na(Out.Unit.Amount), Out.Unit.Amount, D.Unit.Amount),
    D.Unit.Animals = if_else(from_feed_intake2 & !is.na(Out.Unit.Animals), Out.Unit.Animals, D.Unit.Animals),
    D.Unit.Time = if_else(from_feed_intake2 & !is.na(Out.Unit.Time), Out.Unit.Time, D.Unit.Time)
  ) %>%
  select(-ED.Mean.T, -Out.Unit.Amount, -Out.Unit.Animals, -Out.Unit.Time,-ED.Intake.Item) %>%
  distinct()


# Step 8: Final conversion of unit to grams (g)
ingredients <- ingredients %>%
  mutate(
    D.Amount = case_when(
      D.Unit.Amount == "kg" ~ D.Amount * 1000,
      TRUE ~ D.Amount
    ),
    D.Unit.Amount = case_when(
      D.Unit.Amount == "kg" ~ "g",
      TRUE ~ D.Unit.Amount
    )
  )


```



## 5) Subset for Dry data 

When feed intake data is missing and we rely on ingredient-level amounts, we need to ensure that the quantities used refer to the dry form of the ingredient. To do this, we subset the data to include only ingredients identified as dried—based on either the declared DC.Is.Dry value or our earlier flagging using ingredient names and moisture content predictions. This ensures consistency and accuracy in downstream calculations. 
```{r,message=FALSE,warning=FALSE}
#take out unspecified ingredients

species_by_bcode <- merged_metadata$Prod.Out %>%
  select(B.Code, P.Product) %>%
  distinct()

# 2. Papers per species before filtering
papers_before <- ingredients %>%
  select(B.Code) %>%
  distinct() %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_before")
length(unique(ingredients$B.Code))


# Step 1: Identify A.Level.Name groups with at least one ingredient NOT dry
non_dry_diets <- ingredients %>%
  filter(is.na(DC.Is.Dry) | str_to_lower(DC.Is.Dry) %in% c("no", "n", "unspecified")) %>%
  distinct(B.Code, A.Level.Name)

# Step 2: Filter out these diets entirely from ingredients
ingredients <- ingredients %>%
  anti_join(non_dry_diets, by = c("B.Code", "A.Level.Name"))

length(unique(ingredients$A.Level.Name))

rm(non_dry_diets)

length(unique(ingredients$B.Code))
species_by_bcode <- merged_metadata$Prod.Out %>%
  select(B.Code, P.Product) %>%
  distinct()

# 2. Papers per species before filtering
papers_after <- ingredients %>%
  select(B.Code) %>%
  distinct() %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_after")

```

## 6) Complete the nutrition and digestibility papers

Not all ingredients and diets are currently included in the digestibility and composition tables—only those for which data is available. We aim to add the missing ones to help complete the dataset and enable more comprehensive analysis.

```{r}
# --- Step 1: Get all unique (B.Code, D.Item, Source) from the Diet table ---
all_diet_items <- merged_metadata$Animals.Diet %>%
  distinct(B.Code, A.Level.Name, D.Item, Source) %>%
  mutate(
    is_entire_diet = if_else(is.na(D.Item), TRUE, FALSE),
    D.Item = coalesce(D.Item, A.Level.Name)
  ) %>%
  select(B.Code, D.Item, is_entire_diet, Source) %>%
  distinct()

# --- Step 2: Get unique variables in the nutrition and digestibility datasets ---
nutrition_variables <- unique(na.omit(merged_metadata$Animals.Diet.Comp$DC.Variable))
digestibility_variables <- unique(na.omit(merged_metadata$Animals.Diet.Digest$DD.Variable))

# --- Step 3: Expand to full grid of all D.Item × variables (by B.Code and Source) ---
# Nutrition (Comp)
comp_missing <- all_diet_items %>%
  crossing(DC.Variable = nutrition_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Comp %>% select(B.Code, D.Item, DC.Variable),
    by = c("B.Code", "D.Item", "DC.Variable")
  ) %>%
  mutate(
    DC.Value = NA_real_,
    DC.Unit = NA_character_
  )

# Digestibility (Digest)
digest_missing <- all_diet_items %>%
  crossing(DD.Variable = digestibility_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Digest %>% select(B.Code, D.Item, DD.Variable),
    by = c("B.Code", "D.Item", "DD.Variable")
  ) %>%
  mutate(
    DD.Value = NA_real_,
    DD.Unit = NA_character_
  )

# --- Step 4: Bind the missing values into the existing datasets ---
merged_metadata$Animals.Diet.Comp <- bind_rows(merged_metadata$Animals.Diet.Comp, comp_missing)
merged_metadata$Animals.Diet.Digest <- bind_rows(merged_metadata$Animals.Diet.Digest, digest_missing)

```


```{r}
# --- Step 5: Ensure each ingredient has a row for each nutrition variable ---
comp_completion <- merged_metadata$Animals.Diet.Comp %>%
  distinct(B.Code, D.Item, is_entire_diet, Source) %>%
  crossing(DC.Variable = nutrition_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Comp %>% select(B.Code, D.Item, DC.Variable),
    by = c("B.Code", "D.Item", "DC.Variable")
  ) %>%
  mutate(
    DC.Value = NA_real_,
    DC.Unit = NA_character_
  )

# --- Step 6: Ensure each ingredient has a row for each digestibility variable ---
digest_completion <- merged_metadata$Animals.Diet.Digest %>%
  distinct(B.Code, D.Item, is_entire_diet, Source) %>%
  crossing(DD.Variable = digestibility_variables) %>%
  anti_join(
    merged_metadata$Animals.Diet.Digest %>% select(B.Code, D.Item, DD.Variable),
    by = c("B.Code", "D.Item", "DD.Variable")
  ) %>%
  mutate(
    DD.Value = NA_real_,
    DD.Unit = NA_character_
  )

# --- Step 7: Bind additional completion rows ---
merged_metadata$Animals.Diet.Comp <- bind_rows(merged_metadata$Animals.Diet.Comp, comp_completion)
merged_metadata$Animals.Diet.Digest <- bind_rows(merged_metadata$Animals.Diet.Digest, digest_completion)

rm(comp_completion,comp_missing,digest_completion,digest_missing)

```


```{r}
# Step 1: Get species for each B.Code
species_by_bcode <- merged_metadata$Prod.Out %>%
  distinct(B.Code, P.Product)

# Step 2: Filter Comp table for GE values
ge_papers <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", !is.na(DC.Value)) %>%
  distinct(B.Code) %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_with_GE")

# View result
ge_papers

```

## 7) Link ingredients to feedipedia 

Feedipedia is an online database that provides comprehensive information on feed ingredients used in livestock nutrition. It aggregates data on the nutritional composition, digestibility, and other relevant properties of various feeds.We will use it to infer data when info is missing from our data extraction. Links each ingredient to its corresponding entry in Feedipedia, enabling supplementation of missing nutritional and digestibility data such as GE, CP, EE, and Ash.

```{r,message=FALSE,warning=FALSE}

## download data from feedapedia (generated with the "feedapedia scrapping script available on github)

# Use the raw URL of the file on GitHub
url <- "https://github.com/ERAgriculture/ERL/raw/main/downloaded_data/feedipedia.parquet"

# Create a temporary file location with a .parquet extension
temp_file <- tempfile(fileext = ".parquet")

# Download the file in binary mode
download.file(url, destfile = temp_file, mode = "wb")

# Read the Parquet file
feedipedia <- read_parquet(temp_file)

## download AOM with the correspondance between feedipedia codes and diet items 

# Define the raw URL for the Excel file
url <- "https://raw.githubusercontent.com/Mlolita26/ERA_Docs/main/era_master_sheet(5).xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, destfile = temp_file, mode = "wb")

# Read only the sheet called "AOM"
mastersheet <- read_excel(temp_file, sheet = "AOM") %>%
  filter(L5=="Feed Ingredient")

mastersheet$Feedipedia <- as.numeric(sub(".*node/", "", mastersheet$Feedipedia))


```

## 7.1) add nutrition and digestibility from feedipedia 

```{r,message=FALSE,warning=FALSE}
# --- A) Add Feedipedia node from mastersheet into Comp and Digest tables
add_feedipedia_node <- function(df, mastersheet) {
  df %>%
    left_join(
      mastersheet %>% select(Edge_Value, Feedipedia),
      by = c("D.Item" = "Edge_Value")
    ) %>%
    distinct()
}

merged_metadata$Animals.Diet.Comp   <- add_feedipedia_node(merged_metadata$Animals.Diet.Comp, mastersheet)
merged_metadata$Animals.Diet.Digest <- add_feedipedia_node(merged_metadata$Animals.Diet.Digest, mastersheet)

# --- B) Standardize Feedipedia variables and assign categories
digestibility_keywords <- c(
  "digestibility", "degradability", "energy digestibility", 
  "DE", "ME", "NE", "AMEn", "OM digestibility", "DM digestibility",
  "ruminants", "rabbit", "pig", "poultry", "horse", "salmonids", "hydrolysis"
)

feedipedia <- feedipedia %>%
  mutate(
    Variable = case_when(
      tolower(Variable) == "gross energy" ~ "GE",
      tolower(Variable) == "crude protein" ~ "CP",
      tolower(Variable) == "ether extract" ~ "EE",
      tolower(Variable) == "ash" ~ "Ash",
      tolower(Variable) == "Neutral detergent fibre
" ~ "NDF",
      TRUE ~ Variable
    ),
    category = case_when(
      Variable %in% c("GE", "CP", "EE", "Ash","NDF") ~ "nutrition",
      str_detect(Variable, regex(paste(digestibility_keywords, collapse = "|"), ignore_case = TRUE)) ~ "digestibility",
      TRUE ~ "nutrition"
    ),
    feedipedia_node = as.numeric(feedipedia_node)
  )

# --- C) Ensure source columns exist
if (!"nutrition_source" %in% names(merged_metadata$Animals.Diet.Comp)) {
  merged_metadata$Animals.Diet.Comp$nutrition_source <- NA_character_
}
if (!"digestibility_source" %in% names(merged_metadata$Animals.Diet.Digest)) {
  merged_metadata$Animals.Diet.Digest$digestibility_source <- NA_character_
}

# --- D) Nutrition variables: GE, CP, EE, Ash
nutrition_vars <- c("GE", "CP", "EE", "Ash","NDF")

feedipedia_nutrition <- feedipedia %>%
  filter(Variable %in% nutrition_vars) %>%
  group_by(feedipedia_node, Variable, Unit) %>%
  summarise(Value = mean(Avg, na.rm = TRUE), .groups = "drop") %>%
  rename(DC.Variable = Variable, DC.Value.feedipedia = Value)

# --- E) Join nutrition into Animals.Diet.Comp
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(Feedipedia = as.numeric(Feedipedia)) %>%
  left_join(feedipedia_nutrition, by = c("Feedipedia" = "feedipedia_node", "DC.Variable")) %>%
  mutate(
    value_missing = is.na(DC.Value),
    DC.Value = coalesce(DC.Value, DC.Value.feedipedia),
    DC.Unit  = if_else(value_missing & !is.na(DC.Value.feedipedia), Unit, DC.Unit),
    nutrition_source = if_else(value_missing & !is.na(DC.Value.feedipedia), "feedipedia", nutrition_source)
  ) %>%
  select(-DC.Value.feedipedia, -Unit, -value_missing)

# --- F) Digestibility: Energy digestibility, ruminants → DE
feedipedia_digestibility <- feedipedia %>%
  filter(str_detect(tolower(Variable), "energy digestibility")) %>%
  mutate(
    DD.Variable = "DE",  # Map all energy digestibility to "DE"
    feedipedia_node = as.numeric(feedipedia_node)
  ) %>%
  group_by(feedipedia_node, DD.Variable, Unit) %>%
  summarise(DD.Value.feedipedia = mean(Avg, na.rm = TRUE), .groups = "drop")

merged_metadata$Animals.Diet.Digest <- merged_metadata$Animals.Diet.Digest %>%
  mutate(Feedipedia = as.numeric(Feedipedia)) %>%
  left_join(feedipedia_digestibility, by = c("Feedipedia" = "feedipedia_node", "DD.Variable")) %>%
  mutate(
    value_missing = is.na(DD.Value),
    DD.Value = coalesce(DD.Value, DD.Value.feedipedia),
    DD.Unit  = if_else(value_missing & !is.na(DD.Value.feedipedia), Unit, DD.Unit),
    digestibility_source = if_else(value_missing & !is.na(DD.Value.feedipedia), "feedipedia", digestibility_source)
  ) %>%
  select(-DD.Value.feedipedia, -Unit, -value_missing)

```


```{r}
# Step 1: Get species for each B.Code
species_by_bcode <- merged_metadata$Prod.Out %>%
  distinct(B.Code, P.Product)

# Step 2: Filter Comp table for GE values
ge_papers <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", !is.na(DC.Value)) %>%
  distinct(B.Code) %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_with_GE")

# View result
ge_papers

```


## 8) Infer nutritional and digestibility data from raw data and feedipedia  


Now that we have both raw and Feedipedia data, we can use them to infer missing values. If some instances of an ingredient have nutritional or digestibility data while others do not, we can impute the missing values using the average of the available ones.

### 8.1) Nutritional composition inferrence

```{r}
# --- Standardize Units in Animals.Diet.Comp ---
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>% 
  mutate(
    # 1) normalise text --------------------------------------------------------
    DC.Unit     = tolower(trimws(DC.Unit)),
    DC.Value    = as.numeric(DC.Value)
  ) %>% 
  
  # 2) convert values ----------------------------------------------------------
  mutate(
    DC.Value = case_when(
      # --- gross energy → MJ kg-1 -------------------------------------------
      DC.Variable == "GE" ~ case_when(
        DC.Unit %in% c("mj/kg", "mj/kg dm")                       ~ DC.Value,
        DC.Unit %in% c("kcal/kg", "kcal/kg dm")                   ~ DC.Value * 0.004184,
        DC.Unit ==  "kcal/g"                                      ~ DC.Value * 4.184,
        DC.Unit ==  "kcal/100g"                                   ~ DC.Value * 0.04184,   # fixed
        DC.Unit ==  "cal/g"                                       ~ DC.Value * 0.004184,  # fixed
        DC.Unit ==  "mcal/kg"                                     ~ DC.Value * 4.184,
        DC.Unit %in% c("kj/kg", "kj/kg dm")                       ~ DC.Value / 1000,
        DC.Unit ==  "kj/g"                                        ~ DC.Value,             # already MJ/kg
        DC.Unit %in% c("kj/100g", "kj/100g dm", "kj/100 g dm")    ~ DC.Value * 0.01,
        DC.Unit ==  "mj/100g"                                     ~ DC.Value * 10,
        DC.Unit ==  "j/mg"                                        ~ DC.Value,             # fixed
        TRUE                                                      ~ NA_real_
      ),
      
      # --- everything else → g kg-1 -----------------------------------------
      TRUE ~ case_when(
        DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value * 10,
        DC.Unit %in% c("g", "g/kg", "g/kg dm")                         ~ DC.Value,
        DC.Unit %in% c("kg", "kg/day")                                 ~ DC.Value * 1000,
        TRUE                                                           ~ NA_real_
      )
    ),
    
    # 3) harmonised unit labels ----------------------------------------------
    DC.Unit = if_else(DC.Variable == "GE" & !is.na(DC.Value), "MJ/kg", "g/kg")
  )



# --- Compute Mean Value Per Ingredient and Variable ---
ingredient_means <- merged_metadata$Animals.Diet.Comp %>%
  filter(is_entire_diet == FALSE, !is.na(DC.Value)) %>%
  group_by(D.Item, DC.Variable) %>%
  summarise(mean_value = mean(DC.Value, na.rm = TRUE), .groups = "drop")

merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  left_join(ingredient_means, by = c("D.Item", "DC.Variable")) %>%
  mutate(
    nutrition_source = case_when(
      nutrition_source == "feedipedia" ~ "feedipedia",  # preserve Feedipedia
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value) ~ "inferred_from_same_ingredient",  # imputed
      is.na(nutrition_source) & !is.na(DC.Value) ~ "raw data",  # only if there's a value
      TRUE ~ nutrition_source  # keep whatever is there
    ),
    DC.Value = if_else(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value),
      mean_value,
      DC.Value
    )
  ) %>%
  select(-mean_value)


```

### 8.2) Digestibility inferrence
```{r}

# --- B) Compute Mean Digestibility Per Ingredient and Variable ---
digest_means <- merged_metadata$Animals.Diet.Digest %>%
  filter(DD.Variable == "DE", is_entire_diet == FALSE, !is.na(DD.Value)) %>%
  group_by(D.Item, DD.Variable, DD.Unit) %>%
  summarise(mean_digest = mean(DD.Value, na.rm = TRUE), .groups = "drop")


# --- C) Fill Missing Digestibility Values with Ingredient Mean ---
merged_metadata$Animals.Diet.Digest <- merged_metadata$Animals.Diet.Digest %>%
  left_join(digest_means, by = c("D.Item", "DD.Variable")) %>%
  mutate(
    digestibility_source = case_when(
      digestibility_source == "feedipedia" ~ "feedipedia",
      is.na(DD.Value) & !is_entire_diet & !is.na(mean_digest) ~ "inferred_from_same_ingredient",
      is.na(digestibility_source) ~ "raw data",
      TRUE ~ digestibility_source
    ),
    DD.Value = if_else(
      is.na(DD.Value) & !is_entire_diet & !is.na(mean_digest),
      mean_digest,
      DD.Value
    )
  ) %>%
  select(-mean_digest)

```

since there is only very few digestibility in the raw data logic that we dont gain any data by doing this because feedipedia is infered for all the ingredients we have the info for. 

```{r,message=FALSE,warning=FALSE}
# ✅ Count how many papers had GE filled by Feedipedia, and how many of those are Cattle / Sheep / Goat

# Step 1: Filter for Feedipedia-filled GE entries
ge_filled_by_feedipedia <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", nutrition_source == "feedipedia") %>%
  distinct(B.Code)

# Step 2: Total number of studies
n_ge_feedipedia <- nrow(ge_filled_by_feedipedia)

# Step 3: Add species info and filter to ruminants
ge_species_feedipedia <- ge_filled_by_feedipedia %>%
  left_join(merged_metadata$`Prod.Out` %>% select(B.Code, P.Product), by = "B.Code") %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product)

# Step 4: Count per species
ge_species_counts <- ge_species_feedipedia %>%
  count(P.Product, name = "Papers_with_GE_from_Feedipedia")

# Display summary
cat("📌 Total papers with GE filled from Feedipedia:", n_ge_feedipedia, "\n")
cat("📌 Of those, species breakdown:\n")
print(ge_species_counts)

```
```{r}
# Step 1: Get species for each B.Code
species_by_bcode <- merged_metadata$Prod.Out %>%
  distinct(B.Code, P.Product)

# Step 2: Filter Comp table for GE values
ge_papers <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", !is.na(DC.Value)) %>%
  distinct(B.Code) %>%
  left_join(species_by_bcode, by = "B.Code") %>%
  count(P.Product, name = "papers_with_GE")

# View result
ge_papers

```

```{r,message=FALSE,warning=FALSE}

# Define species of interest
species_list <- c("Cattle", "Sheep", "Goat")

# Initialize result containers
species_summary <- list()
bcode_table <- list()

# Loop over each species
for (species in species_list) {
  
  # All B.Codes for this species
  species_bcodes <- merged_metadata$Prod.Out %>%
    filter(P.Product == species) %>%
    pull(B.Code) %>%
    unique()
  
  # Feed intake B.Codes
  feed_data <- merged_metadata$Data.Out %>%
    filter(B.Code %in% species_bcodes, !is.na(ED.Mean.T)) %>%
    select(B.Code, D.Item = ED.Intake.Item)
  
  feed_bcodes <- unique(feed_data$B.Code)
  
  # GE B.Codes
  ge_data <- merged_metadata$Animals.Diet.Comp %>%
    filter(B.Code %in% species_bcodes, DC.Variable == "GE", !is.na(DC.Value)) %>%
    select(B.Code, D.Item)
  
  ge_bcodes <- unique(ge_data$B.Code)
  
  # Matched B.Codes (join on both B.Code and D.Item)
  matched_bcodes <- inner_join(feed_data, ge_data, by = c("B.Code", "D.Item")) %>%
    pull(B.Code) %>%
    unique()
  
  # Store species-level summary
  species_summary[[species]] <- tibble(
    Species = species,
    Total_BCodes = length(species_bcodes),
    With_FEED = length(feed_bcodes),
    With_GE = length(ge_bcodes),
    With_BOTH = length(matched_bcodes)
  )
  
  # Store long-format bcode list
  bcode_table[[species]] <- tibble(
    B.Code = c(species_bcodes, feed_bcodes, ge_bcodes, matched_bcodes),
    Category = rep(c("All", "With_FEED", "With_GE", "With_BOTH"), 
                   times = c(length(species_bcodes), length(feed_bcodes), length(ge_bcodes), length(matched_bcodes))),
    Species = species
  ) %>% distinct()
}

# Bind results
species_summary_df <- bind_rows(species_summary)
bcode_table_df <- bind_rows(bcode_table)

# Show results
print(species_summary_df)
head(bcode_table_df)
```

## Link data to ILRI database

```{r}
## Link data to ILRI database

# Define the raw URL for the Excel file
url <- "https://raw.githubusercontent.com/Mlolita26/ERA_Docs/main/era_master_sheet(5).xlsx"

# Download the file to a temporary location
temp_file <- tempfile(fileext = ".xlsx")
download.file(url, destfile = temp_file, mode = "wb")

# ---- Load ILRI nutrition data ----
ILRI <- read_excel(temp_file, sheet = "ssa_feedsdb", col_types = "text") %>%
  janitor::clean_names()
names(ILRI) <- gsub("\\.\\.\\.[0-9]+", "", names(ILRI))

# Reshape ILRI data and standardize units
ilri_long <- ILRI %>%
  select(ilri_feedcode = ilri_feedcode_6, cp, om, ndf, adf, adl, dm, me, n_el) %>%
  pivot_longer(
    cols = -ilri_feedcode,
    names_to = "DC.Variable",
    values_to = "ILRI_Value"
  ) %>%
  mutate(
    DC.Variable = toupper(DC.Variable),
    ILRI_Value = case_when(
      DC.Variable %in% c("ME", "NEL", "NEM", "NEG") ~ as.numeric(ILRI_Value),
      TRUE ~ as.numeric(ILRI_Value) * 10  # convert % DM to g/kg DM
    ),
    ILRI_Unit = case_when(
      DC.Variable %in% c("ME", "NEL", "NEM", "NEG") ~ "MJ/kg DM",
      TRUE ~ "g/kg DM"
    )
  )

# ---- Load and clean AOM mapping sheet ----
aom_map <- read_excel(temp_file, sheet = "AOM", col_types = "text") %>%
  janitor::clean_names() %>%
  filter(l5 == "Feed Ingredient") %>%
  select(edge_value, ilri_code) %>%
  mutate(ilri_code = as.character(ilri_code))

# ---- Merge ILRI feed codes into diet composition ----
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  left_join(aom_map, by = c("D.Item" = "edge_value"))

```

join nutritional values
```{r}

# ---- Join nutritional values from ILRI ----
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  left_join(ilri_long, by = c("ilri_code" = "ilri_feedcode", "DC.Variable"))


```
fill missing values from ILRI database

```{r}
merged_metadata$Animals.Diet.Comp <- merged_metadata$Animals.Diet.Comp %>%
  mutate(
    DC.Value = if_else(is.na(DC.Value) & !is.na(ILRI_Value), ILRI_Value, DC.Value),
    DC.Unit = if_else(is.na(DC.Unit) & !is.na(ILRI_Value), ILRI_Unit, DC.Unit),
    nutrition_source = if_else(is.na(nutrition_source) & !is.na(ILRI_Value), "ilri", nutrition_source)
  ) %>%
  select(-ILRI_Value, -ILRI_Unit)

```



## 9) Tagg diet practices 

## 9.1) Tagg ingredient with control or trt 

```{r}
##— build the control key from your merged MT.Out
ctrl_key <- merged_metadata$MT.Out[, .(
  B.Code,
  A.Level.Name,
  T.Name,
  T.Control
)]

##— example join back onto your ingredient table
# (assuming your ingredients DT has B.Code + A.Level.Name)
ingredients <- merge(
  x = ingredients, 
  y = ctrl_key, 
  by = c("B.Code","A.Level.Name"), 
  all.x = TRUE, 
  sort = FALSE
)

```

In this section we are going to compare control treatments and treatment and define whether ingredients were substituted or added and what type of ingredient was changed. 

```{r}

classify_practice_changes <- function(ing_tbl) {
  ing_tbl <- ing_tbl %>%
    mutate(T.Control = tolower(trimws(T.Control)))
  
  ing_tbl %>%
    group_by(B.Code) %>%
    group_split() %>%
    map_dfr(function(study) {
      # control items
      ctrl_items <- study %>% 
        filter(T.Control == "yes") %>% 
        pull(D.Item) %>% unique()
      if (length(ctrl_items)==0) return(NULL)
      
      # all non-control diet names
      trt_diets <- study %>% 
        filter(T.Control != "yes") %>% 
        pull(A.Level.Name) %>% unique() %>% sort()
      
      # helper to build one comparison block
      cmp_block <- function(base_items, comp_items,
                      base_name, comp_name, cmp_flag) {

  added   <- setdiff(comp_items, base_items)
  removed <- setdiff(base_items, comp_items)

  practice_type <- case_when(
    length(added)   > 0 & length(removed) > 0 ~ "substitution",
    length(added)   > 0                       ~ "addition",
    TRUE                                       ~ NA_character_
  )

  ## ---------- NEW: label “other_trt:<name>” instead of plain “other_trt” ----
  cmp_label <- ifelse(
    cmp_flag == "other_trt",
    paste0("other_trt:", base_name),      # e.g.  other_trt:Diet_A
    "control"
  )

  bind_rows(
    tibble(
      A.Level.Name  = comp_name,
      compare_to    = cmp_label,          # <-- changed
      other_diet    = ifelse(cmp_flag=="other_trt", base_name, NA_character_),
      change_type   = "added",
      practice_type = practice_type,
      D.Item        = added
    ),
    tibble(
      A.Level.Name  = comp_name,
      compare_to    = cmp_label,          # <-- changed
      other_diet    = ifelse(cmp_flag=="other_trt", base_name, NA_character_),
      change_type   = "removed",
      practice_type = practice_type,
      D.Item        = removed
    )
  ) %>% 
    filter(!is.na(D.Item))
}

      
      # 1) each treatment vs control
      vs_ctrl <- map_dfr(
        trt_diets,
        ~{
          comp_items <- study %>% filter(A.Level.Name==.x) %>% pull(D.Item) %>% unique()
          cmp_block(ctrl_items, comp_items,
                    base_name = "control", comp_name = .x,
                    cmp_flag  = "control")
        }
      )
      
      # 2) pairwise treatment vs treatment
      vs_trt <- if (length(trt_diets)>=2) {
        combn(trt_diets, 2, simplify=FALSE) %>%
          map_dfr(function(pair) {
            d1 <- pair[1]; d2 <- pair[2]
            items1 <- study %>% filter(A.Level.Name==d1)%>%pull(D.Item)%>%unique()
            items2 <- study %>% filter(A.Level.Name==d2)%>%pull(D.Item)%>%unique()
            bind_rows(
              cmp_block(items2, items1, base_name=d2, comp_name=d1, cmp_flag="other_trt"),
              cmp_block(items1, items2, base_name=d1, comp_name=d2, cmp_flag="other_trt")
            )
          })
      } else {
        tibble(
          A.Level.Name=character(),
          compare_to=character(),
          other_diet=character(),
          change_type=character(),
          practice_type=character(),
          D.Item=character()
        )
      }
      
      # combine, attach feed‐type, and finalize
      # combine, attach feed‐type, and finalize
bind_rows(vs_ctrl, vs_trt) %>%
  mutate(B.Code = unique(study$B.Code)) %>%  # <-- add B.Code BEFORE the join
  left_join(
  ing_tbl %>% select(B.Code, D.Item, D.Type) %>% distinct(),
  by = c("B.Code", "D.Item")
) %>%
  select(B.Code, A.Level.Name, compare_to, other_diet,
         change_type, practice_type, D.Item, D.Type)
})
}


```


## 10) Relative % of each ingredient in the diet

This table presents the relative percentage of each ingredient within a diet. To ensure accurate proportion calculations, we first filtered the dataset to include only diets with consistent units. Diets with incompatible or non-harmonizable units were excluded, as their proportions could not be reliably determined.

```{r,message=FALSE,warning=FALSE}
# We now have a single integrated dataset: 'ingredients'.
# 1) Filter to diets that have a single unit type per group (consistent units).
# 2) Compute total amounts and percentages.

inconsistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  filter(Unit_Count > 1) %>%
  ungroup()

# Step 1: Flag only consistent-unit diets
consistent_diets <- ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  # Count only the non-NA units
  mutate(Unit_Count = n_distinct(na.omit(D.Unit.Amount))) %>%
  # Keep groups with 0 or 1 distinct unit (ignoring NA)
  filter(Unit_Count <= 1) %>%
  ungroup()

#porportion of inconsistent diets 
proportion_inconsistent_diets<- length(unique(inconsistent_diets$A.Level.Name))/length(unique(ingredients$A.Level.Name))

# Step 2: Sum total amounts per (B.Code, A.Level.Name, D.Unit.Amount)
# Transform consistent_diets into a summary of totals, removing any NA in amount or unit
# Step 1: Count valid and missing D.Amount per diet
diet_completeness <- consistent_diets %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    n_total = n(),
    n_nonmissing = sum(!is.na(D.Amount)),
    .groups = "drop"
  ) %>%
  filter(n_total == n_nonmissing)  # only fully specified diets


# Step 2: Keep only complete diets when summing
total_amounts <- consistent_diets %>%
  semi_join(diet_completeness, by = c("B.Code", "A.Level.Name")) %>%
  group_by(B.Code, A.Level.Name, D.Unit.Amount) %>%
  summarise(Total_Amount = sum(D.Amount), .groups = "drop")

# Step 1: Standardize feed intake units to grams (g)
feed_intake_diet_g <- feed_intake_diet %>%
  mutate(
    Intake_Amount_g = case_when(
      Out.Unit.Amount == "kg" ~ ED.Mean.T * 1000,
      Out.Unit.Amount == "g" ~ ED.Mean.T,
      TRUE ~ NA_real_
    ),
    D.Unit.Amount = "g"
  ) %>%
  select(B.Code, A.Level.Name, Intake_Amount_g, D.Unit.Amount)

# Step 2: Join with total_amounts — right_join ensures all feed intake rows are retained
total_amounts<- full_join(
  total_amounts %>% rename(Total_Amount_orig = Total_Amount),
  feed_intake_diet_g,
  by = c("B.Code", "A.Level.Name", "D.Unit.Amount")
) %>%
  mutate(
    # Keep original total amount if available, otherwise use feed intake
    Total_Amount = coalesce(Total_Amount_orig, Intake_Amount_g)
  ) %>%
  select(B.Code, A.Level.Name, D.Unit.Amount, Total_Amount)


# Display the resulting table


#Create an interactive table for "total_amounts"
datatable(
  total_amounts,
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)



# Step 3: Calculate percentages
diet_percentages <- consistent_diets %>%
  left_join(total_amounts, by=c("B.Code","A.Level.Name","D.Unit.Amount")) %>%
  mutate(
    Percentage_of_Diet = ifelse(Total_Amount > 0,
                                (D.Amount / Total_Amount)*100,
                                NA_real_)
  ) %>%
  dplyr::select(
    B.Code,
    A.Level.Name,
    D.Item,
    D.Type,
    D.Amount,
    D.Unit.Amount,
    Percentage_of_Diet,
  )

# Table 2: Relative percentage of each ingredient in each diet


#Create an interactive table for "total_amounts"
datatable(
  diet_percentages,
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)

rm(inconsistent_diets,consistent_diets,diet_completeness)

```


```{r,message=FALSE,warning=FALSE}
# Check if each diet sums close to 100%
validation <- diet_percentages %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(Total_Percentage = sum(Percentage_of_Diet, na.rm=TRUE)) %>%
  ungroup()

# Display diets and their total percentages
# datatable(
#   validation, 
#   options = list(
#     pageLength = 10,         # Show 10 rows per page by default
#     scrollX = TRUE,          # Enable horizontal scrolling
#     autoWidth = TRUE,        # Adjust column width automatically
#     searchHighlight = TRUE   # Highlight search terms
#   ),
#   caption = "Validation table"
# )


```

The lines that have 0 as validation are the ones that were feeding ad libitum for which we do not have a precise amount so it's impossible to calculate the relative proportion of the ingredients.


```{r}
# ──────────────────────────────────────────────────────────────────────────────
# 0 · Pre-processing helpers ----------------------------------------------------
# Convert D.Amount to numeric and drop strict duplicates that would otherwise
# blow up the joins further down the pipeline.
# ──────────────────────────────────────────────────────────────────────────────

ingredients <- ingredients %>%
  mutate(D.Amount = as.numeric(D.Amount)) %>%
  distinct(
    B.Code, A.Level.Name, D.Item, D.Amount,
    D.Unit.Amount, D.Unit.Time, D.Unit.Animals,
    .keep_all = TRUE
  )

# Optional sanity check ---------------------------------------------------------

dup_left <- ingredients %>%
  count(B.Code, A.Level.Name, D.Item, D.Amount,
        D.Unit.Amount, D.Unit.Time, D.Unit.Animals) %>%
  filter(n > 1)

if (nrow(dup_left) == 0) {
  message("✅ all identical-amount duplicates removed.")
} else {
  message("⚠️ still some duplicates—investigate dup_left.")
}

# ──────────────────────────────────────────────────────────────────────────────
# quantify_practice_changes() ---------------------------------------------------
#   Computes how one diet changes relative to another within each study.
#   * Works diet-by-diet, study-by-study (keyed by B.Code)
#   * Collapses duplicate ingredient rows *within* each diet before joining
#   * Avoids Cartesian explosions without requiring allow.cartesian = TRUE
# ──────────────────────────────────────────────────────────────────────────────

quantify_practice_changes <- function(ingredients, min_pct = 0.5) {
  ing <- as.data.table(copy(ingredients))  # never modify caller’s object

  #── 0 · basic housekeeping ---------------------------------------------------
  ing[, D.Amount := as.numeric(D.Amount)]
  ing[, T.Control := tolower(trimws(T.Control))]

  # kg → g so we have only one mass basis to compare
  ing[D.Unit.Amount == "kg", `:=`(D.Amount = D.Amount * 1000,
                                    D.Unit.Amount = "g")]

  # % of diet DM per ingredient (computed diet-by-diet)
  ing[, total_g := sum(D.Amount, na.rm = TRUE), by = .(B.Code, A.Level.Name)]
  ing[, pct_diet := (D.Amount / total_g) * 100]

  #── 1 · work study-by-study --------------------------------------------------
  result <- ing[, {
    dt <- .SD                         # only rows for this B.Code (paper)

    # Identify control diets BEFORE collapsing, because we need T.Control ----
    ctrl  <- unique(dt[T.Control == "yes", A.Level.Name])
    diets <- unique(dt$A.Level.Name)

    # Collapse duplicate ingredient rows *within* each diet -------------------
    dt <- dt[, .(
      D.Type          = first(D.Type),          # keep the category
      pct_diet        = sum(pct_diet),
      D.Unit.Amount   = first(D.Unit.Amount),
      D.Unit.Time     = first(D.Unit.Time),
      D.Unit.Animals  = first(D.Unit.Animals)
    ), by = .(A.Level.Name, D.Item)]

    # Build comparison pairs --------------------------------------------------
    pairs <- list()
    if (length(ctrl)) {
      trt <- setdiff(diets, ctrl)
      if (length(trt)) {
        for (c_diet in ctrl) {
          for (t_diet in trt) {
            pairs[[length(pairs) + 1]] <- c(c_diet, t_diet)
          }
        }
      }
    }
    if (length(diets) > 1) {
      pairs <- c(pairs, combn(diets, 2, simplify = FALSE))
    }

    # If no pairs, emit empty DT ---------------------------------------------
    if (length(pairs) == 0) {
      NULL
    } else {
      rbindlist(lapply(pairs, function(p) {
        diet_A <- p[1]; diet_B <- p[2]
        dA <- dt[A.Level.Name == diet_A]
        dB <- dt[A.Level.Name == diet_B]

        grid <- merge(
          dA[, .(D.Item, D.Type, pct_A = pct_diet,
                   unit_A = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals))],
          dB[, .(D.Item, D.Type, pct_B = pct_diet,
                   unit_B = paste(D.Unit.Amount, D.Unit.Time, D.Unit.Animals))],
          by = c("D.Item", "D.Type"), all = TRUE, allow.cartesian = FALSE
        )

        # Unit compatibility check -------------------------------------------
        grid[, unit_ok := (is.na(unit_A) | is.na(unit_B) | unit_A == unit_B)]
        unit_mismatch <- grid[unit_ok == FALSE, D.Item]
        grid <- grid[unit_ok == TRUE]

        # Fill NAs, compute deltas, apply threshold ---------------------------
        grid[is.na(pct_A), pct_A := 0]
        grid[is.na(pct_B), pct_B := 0]
        grid[, delta := pct_B - pct_A]
        grid <- grid[abs(delta) >= min_pct]

        if (nrow(grid) == 0) return(NULL)

        # Aggregate change metrics (overall) ----------------------------------
        total_added    <- grid[pct_A == 0 & delta  > 0,  sum(delta)]
        total_removed  <- grid[pct_B == 0 & delta  < 0, -sum(delta)]
        total_increase <- grid[pct_A > 0 & pct_B > 0 & delta > 0,  sum(delta)]
        total_decrease <- grid[pct_A > 0 & pct_B > 0 & delta < 0, -sum(delta)]
        n_changed      <- nrow(grid)

        # Aggregate *by D.Type* ----------------------------------------------
        fmt <- function(dt) {
          if (nrow(dt) == 0) return(NA_character_)
          paste(paste0(dt$D.Type, ":", round(dt$V1,2)), collapse = "; ")
        }

        add_by_type  <- fmt(grid[pct_A == 0 & delta  > 0,  .(V1 = sum(delta)), by = D.Type])
        rem_by_type  <- fmt(grid[pct_B == 0 & delta  < 0,  .(V1 = -sum(delta)), by = D.Type])
        inc_by_type  <- fmt(grid[pct_A > 0 & pct_B > 0 & delta > 0,  .(V1 = sum(delta)), by = D.Type])
        dec_by_type  <- fmt(grid[pct_A > 0 & pct_B > 0 & delta < 0,  .(V1 = -sum(delta)), by = D.Type])

        # Practice label ------------------------------------------------------
        practice_type <- fifelse(total_added   > 0 & total_removed > 0, "substitution",
                          fifelse(total_added   > 0 & total_removed == 0, "addition",
                          fifelse(total_added   == 0 & total_removed > 0, "removal",
                                 "dosage_change")))

        compare_to <- if (diet_A %in% ctrl) "control" else paste0("other_trt:", diet_A)

        data.table(
          diet_A               = diet_A,
          diet_B               = diet_B,
          compare_to           = compare_to,
          practice_type        = practice_type,
          total_pct_added      = round(total_added,   2),
          total_pct_removed    = round(total_removed, 2),
          total_pct_increase   = round(total_increase,2),
          total_pct_decrease   = round(total_decrease,2),
          added_by_type        = add_by_type,
          removed_by_type      = rem_by_type,
          increase_by_type     = inc_by_type,
          decrease_by_type     = dec_by_type,
          n_ingredients_changed= n_changed,
          unit_mismatch        = paste(unit_mismatch, collapse = "; ")
        )
      }), fill = TRUE)
    }
  }, by = .(B.Code)]

  # final column order ---------------------------------------------------------
  setcolorder(result,
              c("B.Code","diet_A","diet_B","compare_to","practice_type",
                "total_pct_added","total_pct_removed",
                "total_pct_increase","total_pct_decrease",
                "added_by_type","removed_by_type",
                "increase_by_type","decrease_by_type",
                "n_ingredients_changed","unit_mismatch"))

  as.data.frame(result)
}


changes <- quantify_practice_changes(ingredients, min_pct = 1)  # 1 % threshold


changes<- changes %>% 
  mutate(pair_id = map2_chr(diet_A, diet_B, ~paste(sort(c(.x, .y)), collapse = "|"))) %>% 
  distinct(B.Code, pair_id, .keep_all = TRUE) %>%        # keep the first orientation
  select(-pair_id)


```


```{r}

# ── 1 · Helper: explode one “…_by_type” column into tidy rows ─────────────
explode <- function(.data, col, kind) {                         # { 1
  .data %>% 
    select(B.Code, diet_A, diet_B, all_of(col)) %>% 
    mutate(pair_id = paste(diet_A, "→", diet_B)) %>%            # tag each comparison
    mutate(across(all_of(col), ~ ifelse(is.na(.x) | .x == "", NA, .x))) %>% 
    separate_rows(all_of(col), sep = ";\\s*") %>%               # "A:10; B:20" → rows
    filter(!is.na(.data[[col]])) %>% 
    separate(.data[[col]], into = c("D.Type", "pct"),           # "A:10" → A | 10
             sep = ":", convert = TRUE, fill = "right") %>% 
    mutate(
      # ─ collapse categories ────────────────────────────────────────────
      D.Type = case_when(
        str_detect(D.Type, regex("supplement",  ignore_case = TRUE)) ~ "Supplement",
        str_detect(D.Type, regex("concentrate", ignore_case = TRUE)) ~ "Concentrate",
        TRUE                                                        ~ D.Type
      ),
      change_kind = kind
    ) %>% 
    filter(D.Type != "Entire Diet")                               # drop whole-diet rows
}                                                                 # } 1

# ── 2 · Build a long table of ingredient-type moves ────────────────────
type_changes <- bind_rows(                                       # ( 2
  explode(changes, "added_by_type",    "added"),
  explode(changes, "removed_by_type",  "removed"),
  explode(changes, "increase_by_type", "increase"),
  explode(changes, "decrease_by_type", "decrease")
)                                                                 # ) 2

# ── 3 · Count how many comparisons show each Practice × Type combo ────
summary <- type_changes %>% 
  group_by(D.Type, change_kind) %>% 
  summarise(n_comparisons = n_distinct(pair_id), .groups = "drop")

# ── 4 · Plot: bar height = number of comparisons ───────────────────────
ggplot(summary,
       aes(D.Type, n_comparisons, fill = change_kind)) +
  geom_col(position = "dodge") +
  labs(title = "How often each practice affected each ingredient type",
       x     = "Ingredient type",
       y     = "# of diet-pair comparisons",
       fill  = "Change kind") +
  theme_bw() +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        legend.position = "top")


```
```{r}

changes %>% 
  mutate(abs_delta = case_when(
            practice_type == "addition"      ~  total_pct_added,
            practice_type == "removal"       ~  total_pct_removed,
            practice_type == "substitution"  ~  pmax(total_pct_added,
                                                    total_pct_removed),
            practice_type == "dosage_change" ~  pmax(total_pct_increase,
                                                    total_pct_decrease)
          )) %>% 
  ggplot(aes(practice_type, abs_delta, fill = practice_type)) +
  geom_violin(trim = FALSE, alpha = .4, colour = NA) +
  geom_boxplot(width = .15, outlier.shape = NA) +
  stat_summary(fun = median, geom = "point", size = 2, colour = "black") +
  labs(x = NULL, y = "Magnitude (%-points of diet DM)",
       title = "Distribution of practice intensities") +
  theme_bw() +
  theme(legend.position = "none")

```


## 11) Emission calculations 

### 11.1 Estimation of Enteric Methane Emissions Using IPCC Tier 2

Enteric methane (CH₄) emissions were estimated using the **IPCC Tier 2** methodology, which incorporates animal-specific data such as **gross energy intake (GE)** and **methane conversion factors (Ym)**. This approach improves accuracy over Tier 1 by reflecting diet composition, feed intake, and species-specific differences.


### 11.2 Estimation of Daily GE Intake per Animal

For GE values reported at the **entire diet** level, daily gross energy intake per animal was calculated as:

\[
\text{GE}_{\text{daily}} = \text{GE}_{\text{diet}} \times \text{Diet Dry Amount}
\]

Where:

- $\text{GE}_{\text{diet}}$ is the energy content of the complete diet (MJ/kg)
- $\text{Diet Dry Amount}$ is the daily dry matter ingredient (kg/day/animal). We can also infer from feed intake if missing.Needs to be dry matter so we will use only data when DC.is.Dry=yes.

**To do that we will have ton convert all feed intake data into kg/day/animal.**

When GE values were reported at the **ingredient** level, and information on diet composition (ingredient proportions) was available, the diet-level GE was reconstructed as:

\[
\text{GE}_{\text{diet}} = \sum_{i=1}^{n} (\text{GE}_i \times p_i)
\]

Where:

- $\text{GE}_i$ is the GE value of ingredient $i$
- $p_i$ is the proportion of ingredient $i$ in the total diet

Then, daily intake was computed using the same formula above.

### 11.3. Methane Emissions Calculation (Tier 2)

Annual methane emissions were calculated using the IPCC Tier 2 equation:

\[
\text{CH}_4 = \frac{\text{GE}_{\text{daily}} \times \text{Ym} \times 365}{55.65}
\]

Where:

- $\text{CH}_4$ = annual enteric methane emissions (kg CH₄/animal/year)
- $\text{GE}_{\text{daily}}$ = daily gross energy intake per animal (MJ/day)
- $\text{Ym}$ = methane conversion factor:
  - **Cattle**: 0.065 (6.5%)
  - **Sheep and Goats**: 0.060 (6.0%)
- 365 = days per year
- 55.65 = energy content of methane (MJ/kg CH₄)


### 11.4 Data Requirements and Supporting Tables

**1. Table of Diet Composition Proportions**

This table defines the proportion of each ingredient in the overall diet:

```{r,message=FALSE,warning=FALSE}

# Create an interactive table for "total_amounts"
datatable(
  diet_percentages, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Relative % of Ingredients"
)
```


Used to compute reconstructed `GE_diet` from ingredient-level GE values:

\[
\text{GE}_{\text{diet}} = \sum_{i=1}^{n} (\text{GE}_i \times p_i)
\]

---

**2. Table of Ingredient Dry Amounts**

This table stores the amount of each ingredient fed per animal per day, already in or converted to dry matter:

```{r,message=FALSE,warning=FALSE}
# This 'ingredients' now contains all diets + basal integrated.
# Table 1: Final diet amounts with harmonized units (including basal)
datatable(
  ingredients, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Diet Ingredients"
)
```


Useful when GE is per ingredient and diet is not proportionally described.

---

**3. Table of Total Diet Dry Amounts**

This is the total daily dry matter offered or consumed per animal, either directly from intake or inferred from ingredient sums:

```{r,message=FALSE,warning=FALSE}
datatable(
  total_amounts, 
  options = list(
    pageLength = 10,         # Show 10 rows per page by default
    scrollX = TRUE,          # Enable horizontal scrolling
    autoWidth = TRUE,        # Adjust column width automatically
    searchHighlight = TRUE   # Highlight search terms
  ),
  caption = "Interactive Table: Total Amounts of Diet Ingredients"
)

```


\[
\text{GE}_{\text{daily}} = \text{GE}_{\text{diet}} \times \text{Diet Dry Amount}
\]



```{r,message=FALSE,warning=FALSE}

# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

```


## 12) Gross energy data 

## 12.1) Gross energy harmonization

```{r}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE") %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

```


```{r,message=FALSE,warning=FALSE}
# Inspect the units used for GE


Gross_energy <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "GE")

Gross_energy$DC.Value<-as.numeric(Gross_energy$DC.Value)

unique(Gross_energy$DC.Unit)

# Harmonize only GE values and overwrite DC.Value
## ---- harmonise_gross_energy_to_MJkg -----------------------------------------
Gross_energy <- Gross_energy %>% 
  mutate(
    ## 1) lower-case + trim once ----------------------------------------------
    unit_raw    = tolower(trimws(DC.Unit))
  ) %>% 
  
  ## 2) convert GE values to MJ/kg --------------------------------------------
  mutate(
    DC.Value = case_when(
      DC.Variable == "GE" ~ case_when(
        unit_raw %in% c("mj/kg", "mj/kg dm")                       ~ DC.Value,
        unit_raw %in% c("kcal/kg", "kcal/kg dm")                   ~ DC.Value * 0.004184,
        unit_raw ==  "kcal/g"                                      ~ DC.Value * 4.184,
        unit_raw ==  "kcal/100g"                                   ~ DC.Value * 0.04184,  # 1 kcal = 0.004184 MJ, ×10 (100 g→kg)
        unit_raw ==  "cal/g"                                       ~ DC.Value * 0.004184, # 1 cal g⁻¹ → 0.004184 MJ kg⁻¹
        unit_raw ==  "mcal/kg"                                     ~ DC.Value * 4.184,
        unit_raw %in% c("kj/kg", "kj/kg dm")                       ~ DC.Value / 1000,
        unit_raw ==  "kj/g"                                        ~ DC.Value,            # already MJ kg⁻¹ (kJ g⁻¹ ×1000)
        unit_raw %in% c("kj/100g", "kj/100g dm", "kj/100 g dm")    ~ DC.Value * 0.01,
        unit_raw ==  "mj/100g"                                     ~ DC.Value * 10,
        unit_raw ==  "j/mg"                                        ~ DC.Value,            # J mg⁻¹ → MJ kg⁻¹ (10⁻⁶ / 10⁻⁶)
        TRUE                                                       ~ NA_real_
      ),
      TRUE ~ DC.Value  # non-GE rows stay intact
    ),
    
    ## 3) update unit label ----------------------------------------------------
    DC.Unit = if_else(DC.Variable == "GE" & !is.na(DC.Value), "MJ/kg", unit_raw)
  ) %>% 
  
  ## 4) tidy up ---------------------------------------------------------------
  select(-unit_raw)




unique(Gross_energy$DC.Unit)

Gross_energy <- Gross_energy %>%
  # STEP 1: Create GE_source from nutrition_source
  mutate(
    GE_source = case_when(
      nutrition_source == "feedipedia" ~ "feedipedia",
      nutrition_source == "inferred_from_same_ingredient" ~ "inferred_from_same_ingredient",
      nutrition_source == "raw data" ~ "raw data",
      TRUE ~ NA_character_
    )
  )%>%
  select(-"nutrition_source")



```
g/kg	❌ Not an energy unit	Likely a mistake, investigate or exclude
% / % DM / % as fed	❌ Not energy units	Same as above
Unspecified	❌ Ambiguous	Investigate or exclude

```{r,message=FALSE,warning=FALSE}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```


## 12.2 Gross energy data check 

```{r}

Gross_energy_check <- Gross_energy %>%
  filter(!is.na(DC.Value))%>%
  filter(GE_source=="raw data")


```



## 13) Estimate gross energy for ingredients from proximate nutritional composition

To estimate Gross Energy (GE) when direct measurements are not available, we use a predictive equation based on proximate nutritional components: **crude protein (CP)**, **ether extract (EE)**, and **ash**. This method provides an accurate approximation of the energy content of feed ingredients based on their macronutrient composition.

The equation used is adapted from **Weiss et al. (2019)** and is expressed as:

```
GE (Mcal/kg) = (0.056 × CP) + (0.094 × EE) + [0.042 × (100 − CP − EE − Ash)]
GE (MJ/kg) = GE (Mcal/kg) × 4.184
```


```{r}
# --- STEP 1: Prepare predictors (CP, EE, Ash) as % of DM ---
# Keep relevant metadata for later
metadata_cols <- merged_metadata$Animals.Diet.Comp %>%
  select(B.Code, D.Item, is_entire_diet, Feedipedia) %>%
  distinct()

ge_predictors <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable %in% c("CP", "EE", "Ash")) %>%
  mutate(
    DC.Unit = trimws(tolower(DC.Unit)),
    DC.Value = as.numeric(DC.Value),

    # Harmonize to % of DM
    DC.Value = case_when(
      DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value,
      DC.Unit %in% c("g", "g/kg", "g/kg dm")                         ~ DC.Value / 10,
      DC.Unit %in% c("kg", "kg/day")                                ~ DC.Value * 100,
      TRUE ~ NA_real_
    ),
    DC.Unit = "%"
  ) %>%
  group_by(B.Code, D.Item, DC.Variable) %>%
  summarise(DC.Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = DC.Value) %>%
  left_join(metadata_cols, by = c("B.Code", "D.Item"))

# --- STEP 2: Apply Weiss GE equation only where CP, EE, Ash are all present ---
ge_predicted <- ge_predictors %>%
  mutate(
    GE_Mcal_per_kg = ifelse(
      !is.na(CP) & !is.na(EE) & !is.na(Ash),
      (CP * 0.056) + (EE * 0.094) + ((100 - CP - EE - Ash) * 0.042),
      NA_real_
    ),
    GE_MJ_per_kg = GE_Mcal_per_kg * 4.184
  ) %>%
  select(B.Code, D.Item, GE_MJ_per_kg, is_entire_diet, Feedipedia)


# --- STEP 3: Merge and only impute where GE is missing ---
Gross_energy <- Gross_energy %>%
  
  # STEP 2: Join GE predictions (e.g. from Weiss)
  full_join(ge_predicted, by = c("B.Code", "D.Item","is_entire_diet","Feedipedia")) %>%
  
  # STEP 3: Only update GE_source if it is still missing
  mutate(
    GE_source = case_when(
      is.na(GE_source) & !is.na(GE_MJ_per_kg) & is.na(DC.Value) ~ "predicted_from_Weiss_GE",
      TRUE ~ GE_source
    ),
    DC.Value = if_else(is.na(DC.Value), GE_MJ_per_kg, DC.Value),
    DC.Unit = "MJ/kg",
    DC.Variable = "GE"
  ) %>%
  select(-GE_MJ_per_kg)




```


```{r}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```


## 14) Infer gross energy from data 
As we did in 9) we are inferring again GE data after estimation by proximal nutritional composition. 

```{r}
ingredient_means <- Gross_energy %>%
  filter(is_entire_diet == FALSE, !is.na(DC.Value)) %>%
  group_by(D.Item, DC.Variable) %>%
  summarise(mean_value = mean(DC.Value, na.rm = TRUE), .groups = "drop")

# --- Fill Missing Values with Mean ---
Gross_energy <- Gross_energy %>%
  left_join(ingredient_means, by = c("D.Item", "DC.Variable")) %>%
  mutate(
    GE_source = case_when(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value) ~ "inferred_from_same_ingredient-Step2",
      TRUE ~ GE_source
    ),
    DC.Value = if_else(
      is.na(DC.Value) & !is_entire_diet & !is.na(mean_value),
      mean_value,
      DC.Value
    )
  ) %>%
  select(-mean_value)
```


```{r}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```


## 15) Estimate Digestible energy from proximate nutritional composition

```{r}
# --- STEP 4: Prepare predictors for Weiss DE equation ---
# --- STEP 1: Store relevant metadata before pivot ---
de_metadata_cols <- merged_metadata$Animals.Diet.Comp %>%
  select(B.Code, D.Item, is_entire_diet, Feedipedia) %>%
  distinct()

# --- STEP 2: Prepare DE predictors and harmonize units ---
de_predictors <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable %in% c("CP", "EE", "Ash", "NDF")) %>%
  mutate(
    DC.Unit = trimws(tolower(DC.Unit)),
    DC.Value = as.numeric(DC.Value),

    # Harmonize to g/kg DM
    DC.Value = case_when(
      DC.Unit %in% c("%", "% dm", "% as fed", "g/100g", "g/100g dm") ~ DC.Value * 10,
      DC.Unit %in% c("g", "g/kg", "g/kg dm")                         ~ DC.Value,
      DC.Unit %in% c("kg", "kg/day")                                ~ DC.Value * 1000,
      TRUE ~ NA_real_
    ),
    DC.Unit = "g/kg"
  ) %>%
  group_by(B.Code, D.Item, DC.Variable) %>%
  summarise(DC.Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = DC.Variable, values_from = DC.Value) %>%
  left_join(de_metadata_cols, by = c("B.Code", "D.Item"))

# --- STEP 3: Apply Weiss DE equation and convert to GE ---
de_predicted <- de_predictors %>%
  mutate(
    DE_Mcal_per_kg = ifelse(
      !is.na(CP) & !is.na(EE) & !is.na(Ash) & !is.na(NDF),
      1.044 + 0.0101 * CP + 0.0161 * EE - 0.0037 * Ash - 0.0023 * NDF,
      NA_real_
    ),
    DE_MJ_per_kg = DE_Mcal_per_kg * 4.184,
    GE_MJ_per_kg = ifelse(!is.na(DE_MJ_per_kg), DE_MJ_per_kg / 0.54, NA_real_)
  ) %>%
  select(B.Code, D.Item, GE_MJ_per_kg, is_entire_diet, Feedipedia)


# --- STEP 6: Fill remaining missing GE using DE-based predictions only if GE is still missing ---
Gross_energy <- full_join(Gross_energy, de_predicted, by = c("B.Code", "D.Item", "is_entire_diet", "Feedipedia")) %>%
  mutate(
    DC.Value = if_else(is.na(DC.Value) & !is.na(GE_MJ_per_kg), GE_MJ_per_kg, DC.Value),
    GE_source = case_when(
      is.na(GE_source) & is.na(DC.Value) & !is.na(GE_MJ_per_kg) ~ "predicted_from_Weiss_DE",
      TRUE ~ GE_source  # keep existing label (e.g., feedipedia, raw data, inferred, or missing)
    ),
    DC.Unit = "MJ/kg",
    DC.Variable = "GE"
  ) %>%
  select(-GE_MJ_per_kg)


```

```{r,message=FALSE,warning=FALSE}
GE_data <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```

We dont get more papers seems logical because probably the same ingredient as the equation  for GE. 

## 16) Convert digestible energy into gross energy 
Converts digestible energy (DE) to gross energy (GE) using a default digestibility factor (54%) when DE is reported but GE is not.

```{r,message=FALSE,warning=FALSE}
# 1. Total papers per species
species_counts <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Total_Studies")

# 2. All papers with GE (non-NA)
DE_data <- merged_metadata$Animals.Diet.Comp %>%
  filter(DC.Variable == "DE", !is.na(DC.Value) & DC.Value != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
DE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(DE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_DE")

# 4. Combine the two
summary_table <- species_counts %>%
  full_join(DE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

# Display table
kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")
```

Most of the DE data is coming from feediepdia which is expressed as a % of DE so we can't convert that data back using the IPCC method. 

```{r,message=FALSE,warning=FALSE}

# # Define digestibility conversion factor (Song et al., 2025)
# default_DE_fraction <- 0.54  # 54%
# 
# # 1. Extract DE rows from merged_metadata
# de_data <- merged_metadata$Animals.Diet.Digest %>%
#   filter(DD.Variable == "DE", !is.na(DD.Value) & DD.Value != "") %>%
#   mutate(
#     DD.Value = as.numeric(DD.Value),
#     DD.Unit = trimws(tolower(DD.Unit)),
#     DD.Value = case_when(
#       DD.Unit %in% c("mj/kg", "mj/kg dm") ~ DD.Value,
#       DD.Unit == "mcal/kg" ~ DD.Value * 4.184,
#       DD.Unit == "kcal/kg" ~ DD.Value / 239,
#       DD.Unit == "kcal/kg dm" ~ DD.Value / 239,
#       DD.Unit == "kcal/100g" ~ (DD.Value * 10) / 239,
#       DD.Unit == "kj/g" ~ DD.Value / 1000,
#       DD.Unit == "kj/100g" ~ DD.Value / 100,
#       TRUE ~ NA_real_
#     ),
#     DD.Unit = "MJ/kg"
#   ) %>%
#   mutate(
#     GE_from_DE = DD.Value / default_DE_fraction
#   ) %>%
#   select(B.Code, D.Item, GE_from_DE)
# 
# # 2. Merge GE_from_DE into your existing Gross_energy dataset
# Gross_energy <- Gross_energy %>%
#   left_join(de_data, by = c("B.Code", "D.Item")) %>%
#   mutate(
#     # Flag only when GE was previously missing and DE is available
#     GE_source = case_when(
#       is.na(GE_source) & is.na(DC.Value) & !is.na(GE_from_DE) ~ "inferred_from_DE",
#       TRUE ~ GE_source
#     ),
#     # Fill in missing GE values with GE_from_DE
#     DC.Value = if_else(is.na(DC.Value), GE_from_DE, DC.Value),
#     DC.Unit = "MJ/kg",
#     DC.Variable = "GE"
#   ) %>%
#   select(-GE_from_DE)


```

```{r,message=FALSE,warning=FALSE}
# 1. Total studies per species
species_total <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  distinct(B.Code, P.Product)

# 2. GE: Only B.Codes where DC.Variable == GE and DC.Value is NOT NA
ge_bcodes <- Gross_energy %>%
  filter(DC.Variable == "GE", !is.na(DC.Value)) %>%
  distinct(B.Code) %>%
  pull()

# 3. Feed: Only B.Codes where D.Amount is NOT NA
feed_amount_bcodes <- ingredients %>%
  filter(!is.na(D.Amount)) %>%
  distinct(B.Code) %>%
  pull()

# 4. B.Codes with BOTH GE value AND feed amount
both_bcodes <- intersect(ge_bcodes, feed_amount_bcodes)

# 5. Build final summary table by species
final_summary <- species_total %>%
  mutate(
    Has_GE = B.Code %in% ge_bcodes,
    Has_FEED_or_Amount = B.Code %in% feed_amount_bcodes,
    Has_BOTH = B.Code %in% both_bcodes
  ) %>%
  group_by(P.Product) %>%
  summarise(
    Total_Studies = n_distinct(B.Code),
    Studies_with_GE = sum(Has_GE),
    Studies_with_FEED_or_Amount = sum(Has_FEED_or_Amount),
    Studies_with_BOTH = sum(Has_BOTH)
  ) %>%
  arrange(P.Product)

# Show summary
knitr::kable(final_summary, caption = "Final Summary (with actual GE and Feed Amount values)")

# Optional: List the actual B.Codes with both
bcode_list_both <- species_total %>%
  filter(B.Code %in% both_bcodes) %>%
  arrange(P.Product, B.Code)

# Show the B.Codes table
knitr::kable(bcode_list_both, caption = "B.Codes with BOTH GE (DC.Value) and Feed Intake (D.Amount)")


```



## 17) Calculation of CH4 emissions

To estimate enteric methane emissions, a methane conversion factor (Ym) is applied to the gross energy (GE) intake of livestock. Ym represents the proportion of the animal’s gross energy intake that is converted to methane and varies based on the species, productivity level, and diet composition. The IPCC (2019) provides updated Ym values based on feed digestibility and forage content. Below is the logic applied:


```{r}
library(knitr)
kable(
  data.frame(
    Category = c("Dairy Cattle (High)", "Dairy Cattle (Medium)", "Dairy Cattle (Low)", "Dairy Cattle (Unknown)",
                 "Meat Cattle (>75% forage)", "Meat Cattle (15–75% forage)", "Meat Cattle (<15% forage)", "Meat Cattle (Unknown)",
                 "Sheep", "Lambs", "Goats"),
    Ym_Percent = c(5.7, 6.3, 6.5, 6.0, 7.0, 6.3, 4.0, 6.5, 6.7, 4.5, 5.5),
    Condition = c("> 8500 kg milk/year", "5000–8500 kg milk/year", "< 5000 kg milk/year", "Unknown milk yield",
                  "Forage > 75%", "Forage 15–75%", "Forage < 15%", "Unknown forage %", 
                  "Default for Sheep", "If 'lamb' in Herd.Stage", "Default for Goats")
  ),
  caption = "Assigned Ym Values Based on IPCC 2019 Guidelines and Available Data"
)

```


### 17.1) Ym estimations 

```{r}
# Step 1: Identify purpose of each B.Code from the production output table
# Dairy = Milk Yield, Meat = Meat Yield

#harmonize yield data units 

yield_data<-merged_metadata$Data.Out%>%
  filter(Out.Subind %in% c("Meat Yield","Milk Yield"))

yield_data <- yield_data %>%
  mutate(
    Out.Unit = tolower(trimws(Out.Unit)),  # Normalize casing and whitespace
    Out.Unit = case_when(
      # Common direct mappings
      Out.Unit %in% c("kg", "kg/d", "kg/day", "kg/day/individual", "kg/individual/day", 
                      "kg/d/individual", "kg/individual") ~ "kg/individual/day",
      
      Out.Unit %in% c("g", "g/d", "g/day", "g/day/individual", "g/d/individual", 
                      "g/individual/day", "g/individual") ~ "g/individual/day",
      
      Out.Unit %in% c("mg/d/individual") ~ "mg/individual/day",
      
      Out.Unit %in% c("l/d", "l/day", "l/d/individual", "l/individual/day", "L/individual/day",
                      "l/individual") ~ "l/individual/day",
      
      Out.Unit %in% c("ml/d/individual", "ml/individual", "ml/individual/day") ~ "ml/individual/day",
      
      
      # Composite or unclear units
      Out.Unit == "g/kg/individual" ~ "g/kg/individual",

      
      Out.Unit == "kg/replicate/experiment" ~ "kg/replicate/experiment",
      
      TRUE ~ Out.Unit  # fallback
    )
  )

yield_data <- yield_data %>%
  mutate(
    Out.Unit = tolower(trimws(Out.Unit)),

    # Convert to standard units
    ED.Mean.T = case_when(
      Out.Unit == "g/individual/day" ~ ED.Mean.T / 1000,         # g → kg
      Out.Unit == "mg/individual/day" ~ ED.Mean.T / 1e6,         # mg → kg
      Out.Unit == "ml/individual/day" ~ ED.Mean.T / 1000,        # ml → l
      TRUE ~ ED.Mean.T                                            # leave others
    ),

    # Normalize units
    Out.Unit = case_when(
      Out.Unit %in% c("kg/individual/day", "g/individual/day", "mg/individual/day") ~ "kg/individual/day",
      Out.Unit %in% c("l/individual/day", "ml/individual/day") ~ "l/individual/day",
      TRUE ~ NA_character_  # mark invalid units
    )
  ) %>%
  # Keep only acceptable units
  filter(Out.Unit %in% c("kg/individual/day", "l/individual/day"))

milk_yield <- yield_data %>%
  filter(Out.Subind == "Milk Yield", Out.Unit == "kg/individual/day") %>%
  group_by(B.Code) %>%
  summarise(
    milk_kg_year = mean(ED.Mean.T, na.rm = TRUE) * 365,
    .groups = "drop"
  )


# Step 1: Get purpose (Milk vs. Meat) per B.Code, prioritizing Milk if both exist
purpose_by_code <- merged_metadata$Data.Out %>%
  filter(Out.Subind %in% c("Milk Yield", "Meat Yield")) %>%
  mutate(
    Purpose = case_when(
      Out.Subind == "Milk Yield" ~ "Dairy",
      Out.Subind == "Meat Yield" ~ "Meat",
      TRUE ~ NA_character_
    )
  ) %>%
  group_by(B.Code) %>%
  # Prioritize Dairy if both exist
  summarise(Purpose = if ("Dairy" %in% Purpose) "Dairy" else first(Purpose), .groups = "drop")



ingredients_cleaned <- ingredients %>%
  mutate(
    D.Amount = ifelse(D.Unit.Amount == "g", D.Amount / 1000, D.Amount),
    D.Unit.Amount = ifelse(D.Unit.Amount == "g", "kg", D.Unit.Amount)
  )

# Step 3: Keep only diets where all units are now consistent (all in kg)
diet_valid_units <- ingredients_cleaned %>%
  group_by(B.Code, A.Level.Name) %>%
  filter(n_distinct(D.Unit.Amount, na.rm = TRUE) == 1) %>%
  ungroup()


# Step 2: Tag forage types (you can expand this list as needed)
forage_types <- c("Forage Trees", "Crop Byproduct", 
                  "Herbaceous Fodders", "Agroforestry Fodders")

forage_proportions <- diet_valid_units %>%
  filter(D.Type != "Entire Diet") %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    any_na = any(is.na(D.Amount)),  # Check for NA in D.Amount
    total_diet = ifelse(any_na, NA_real_, sum(D.Amount, na.rm = TRUE)),
    total_forage = ifelse(any_na, NA_real_, sum(D.Amount[D.Type %in% forage_types], na.rm = TRUE)),
    n_ingredients = n(),
    is_forage = all(D.Type[1] %in% forage_types),
    D.Type.first = D.Type[1],
    D.Unit.Amount = first(D.Unit.Amount),
    forage_prop = case_when(
      n_ingredients == 1 & D.Type.first %in% forage_types ~ 1,
      n_ingredients == 1 & !(D.Type.first %in% forage_types) ~ 0,
      total_diet > 0 ~ total_forage / total_diet,
      TRUE ~ NA_real_
    ),
    .groups = "drop"
  )



```


### 17.2) Reconstruct Gross Energy at the diet level from ingredient data 
Reconstructs the gross energy of the entire diet from the weighted average of GE values of individual ingredients, when diet-level GE is not reported.


```{r,message=FALSE,warning=FALSE}

# --- Calculate GE_diet (MJ/kg) from ingredient GE and amount ---

# 1. Join GE values to ingredient amounts
# --- Calculate GE_diet (MJ/kg) ONLY if all ingredients have GE ---

# 1. Join GE values to ingredient amounts
ingredients_with_GE <- ingredients %>%
  mutate(Diet_ID = paste(B.Code, A.Level.Name, sep = "__")) %>%
  left_join(
    Gross_energy %>%
      filter(!is_entire_diet) %>%
      select(B.Code, D.Item, DC.Value),
    by = c("B.Code", "D.Item")
  ) %>%
  mutate(
    D.Amount_kg = case_when(
      D.Unit.Amount == "g" ~ D.Amount / 1000,
      D.Unit.Amount == "kg" ~ D.Amount,
      TRUE ~ NA_real_
    )
  )

# 2. Count ingredients with/without GE
ingredient_counts <- ingredients_with_GE %>%
  group_by(Diet_ID) %>%
  summarise(
    n_total = n(),
    n_with_GE = sum(!is.na(DC.Value)),
    .groups = "drop"
  ) %>%
  filter(n_total == n_with_GE)

# 3. Reconstruct GE_diet (MJ/kg)
GE_from_ingredients <- ingredients_with_GE %>%
  semi_join(ingredient_counts, by = "Diet_ID") %>%
  group_by(Diet_ID, B.Code, A.Level.Name) %>%
  summarise(
    GE_diet = sum(DC.Value * D.Amount_kg) / sum(D.Amount_kg),
    .groups = "drop"
  )

# 4. Use GE_diet when full diet-level GE is missing
# Join forage proportions into GE_diet_reconstructed
GE_diet_reconstructed <- GE_from_ingredients %>%
  left_join(total_amounts %>%
              select(B.Code, A.Level.Name, Total_Amount, D.Unit.Amount),
            by = c("B.Code", "A.Level.Name")) %>%
  left_join(merged_metadata$Prod.Out %>% select(B.Code, P.Product, Herd.Stage), by = "B.Code") %>%
  left_join(purpose_by_code, by = "B.Code") %>%
  left_join(forage_proportions %>% select(B.Code, A.Level.Name, forage_prop), 
            by = c("B.Code", "A.Level.Name")) %>%
  left_join(milk_yield, by = "B.Code") %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  mutate(
    Total_Amount_kg = case_when(
      D.Unit.Amount == "g" ~ Total_Amount / 1000,
      D.Unit.Amount == "kg" ~ Total_Amount,
      TRUE ~ NA_real_
    ),
    D.Unit.Amount = "kg",  # <- FORCE ALL TO BE KG
    
    Ym = case_when(
      P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year > 8500 ~ 0.057,
      P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year >= 5000 ~ 0.063,
      P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year < 5000 ~ 0.065,
      P.Product == "Cattle" & Purpose == "Dairy" & is.na(milk_kg_year)  ~ 0.06,
      P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.75 ~ 0.07,
      P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.15 ~ 0.063,
      P.Product == "Cattle" & Purpose == "Meat" & forage_prop >= 0 ~ 0.04,
      P.Product == "Cattle" & Purpose == "Meat" & is.na(forage_prop) ~ 0.065,
      P.Product == "Sheep" & grepl("lamb", Herd.Stage, ignore.case = TRUE) ~ 0.045,
      P.Product == "Sheep" ~ 0.067,
      P.Product == "Goat" ~ 0.055,
      TRUE ~ NA_real_
    ),
    
    GE_daily = ifelse(!is.na(GE_diet) & !is.na(Total_Amount_kg),
                      GE_diet * Total_Amount_kg,
                      NA_real_),
    
    CH4_kg_year = ifelse(!is.na(GE_daily) & !is.na(Ym),
                         GE_daily * Ym * 365 / 55.65,
                         NA_real_),
    
    CH4_flag = case_when(
      is.na(GE_diet) ~ "missing_GE_diet",
      is.na(Total_Amount_kg) ~ "missing_amount",
      is.na(Ym) ~ "missing_Ym",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      TRUE ~ "GE_diet_based"
    ),
    
    source = "GE_diet_from_ingredients"
  ) %>%
  rename(D.Item = A.Level.Name) %>%
  select(B.Code, D.Item, P.Product, Herd.Stage, GE_daily, CH4_kg_year, 
         Total_Amount = Total_Amount_kg, D.Unit.Amount, ED.Mean.T = Total_Amount_kg, 
         CH4_flag, Ym, source)


```


### 17.3) Gross energy at the entire diet level 
```{r,message=FALSE,warning=FALSE}

Gross_energy_entire <- Gross_energy %>%
  filter(is_entire_diet == TRUE) %>%
  
  # Add feed amount and unit
  left_join(
    total_amounts %>% select(B.Code, A.Level.Name, Total_Amount, D.Unit.Amount),
    by = c("B.Code" = "B.Code", "D.Item" = "A.Level.Name")
  ) %>%

  # Add product info and herd stage
  left_join(
    merged_metadata$Prod.Out %>% select(B.Code, P.Product, Herd.Stage),
    by = "B.Code"
  ) %>%

  # Add dairy vs meat purpose
  left_join(purpose_by_code, by = "B.Code") %>%

  mutate(
    # Convert Total_Amount to kg/day
    Total_Amount = case_when(
      D.Unit.Amount == "g" ~ Total_Amount / 1000,
      D.Unit.Amount == "kg" ~ Total_Amount,
      TRUE ~ NA_real_
    ),

    # Fix unit
    D.Unit.Amount = case_when(
      D.Unit.Amount == "g" ~ "kg",
      TRUE ~ D.Unit.Amount
    ),

    # Diagnostic reason for failure
    reason_missing = case_when(
      is.na(Total_Amount) & D.Unit.Amount %in% c("g/kg", "%", "% Diet") ~ "Cannot convert unit to kg",
      TRUE ~ NA_character_
    ),

    # Gross energy intake
    GE_daily = ifelse(
      !is.na(Total_Amount) & !is.na(DC.Value),
      DC.Value * Total_Amount,
      NA_real_
    )
  )




```

### 17.4) Merge GE data with feed intake 

```{r,message=FALSE,warning=FALSE}
  # Make sure 'purpose_by_code' is unique per B.Code
purpose_by_code_unique <- purpose_by_code %>%
  distinct(B.Code, Purpose)

# Then join safely before using 'Purpose'
GE_all <- Gross_energy_entire %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  left_join(forage_proportions %>% select(B.Code, A.Level.Name, forage_prop), 
            by = c("B.Code",  "D.Item" = "A.Level.Name")) %>%
  left_join(milk_yield, by = "B.Code") %>%
  left_join(
    feed_intake_diet %>% 
      select(B.Code, A.Level.Name, ED.Mean.T),
    by = c("B.Code", "D.Item" = "A.Level.Name")
  ) %>%
  mutate(
    Total_Amount = if_else(!is.na(ED.Mean.T), ED.Mean.T, Total_Amount),
    D.Unit.Amount = if_else(!is.na(ED.Mean.T), "kg", D.Unit.Amount),
Ym = case_when(
  # --- Dairy cattle based on productivity ---
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year > 8500 ~ 0.057,
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year >= 5000 ~ 0.063,
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year < 5000 ~ 0.065,
  P.Product == "Cattle" & Purpose == "Dairy" & is.na(milk_kg_year)  ~ 0.06,
  
  # --- Meat cattle based on forage proportions ---
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.75 ~ 0.07,
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.15 ~ 0.063,
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop >= 0 ~ 0.04,
  P.Product == "Cattle" & Purpose == "Meat" & is.na(forage_prop) ~ 0.065,

  # --- Sheep & Lambs ---
  P.Product == "Sheep" & grepl("lamb", Herd.Stage, ignore.case = TRUE) ~ 0.045,
  P.Product == "Sheep" ~ 0.067,

  # --- Goats ---
  P.Product == "Goat" ~ 0.055,
  
  TRUE ~ NA_real_
),
    
    GE_daily = ifelse(!is.na(DC.Value) & !is.na(Total_Amount),
                      DC.Value * Total_Amount,
                      NA_real_),
    
    CH4_kg_year = ifelse(!is.na(GE_daily) & !is.na(Ym),
                         GE_daily * Ym * 365 / 55.65,
                         NA_real_),
    
    CH4_flag = case_when(
      is.na(DC.Value) ~ "missing_GE_value",
      is.na(Ym) ~ "missing_Ym",
      is.na(Total_Amount) ~ "missing_amount",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      !is.na(ED.Mean.T) ~ "amount_inferred from feed intake",
      TRUE ~ "ok"
    )
  )




```



```{r,message=FALSE,warning=FALSE}
###############################################################################
## SINGLE CHUNK: Sum Ingredient Intake & Merge into Entire-Diet GE + CH₄ Calc ##
###############################################################################

# 1) Sum ingredient-level feed intake -----------------------------------------


ingredient_intake_summed <- feed_intake_ingredients %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    sum_ingredient_intake = sum(ED.Mean.T, na.rm = TRUE),
    .groups = "drop"
  )

# 2) Merge this ingredient sum into the entire-diet GE table ------------------
#    Then calculate final intake, daily GE, and CH₄.
GE_all_reconstructed <- Gross_energy_entire %>%
    left_join(forage_proportions %>% select(B.Code, A.Level.Name, forage_prop), 
            by = c("B.Code",  "D.Item" = "A.Level.Name")) %>%
  left_join(milk_yield, by = "B.Code") %>%
  # Add species info and keep only ruminants
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  
  # Merge the ingredient-level summed intake (aliasing A.Level.Name to D.Item)
  left_join(
    ingredient_intake_summed,
    by = c("B.Code", "D.Item" = "A.Level.Name")
  ) %>%

  # Compute CH₄-related fields
  mutate(
    # Methane conversion factor by species
Ym = case_when(
  # --- Dairy cattle based on productivity ---
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year > 8500 ~ 0.057,
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year >= 5000 ~ 0.063,
  P.Product == "Cattle" & Purpose == "Dairy" & milk_kg_year < 5000 ~ 0.065,
  P.Product == "Cattle" & Purpose == "Dairy" & is.na(milk_kg_year)  ~ 0.06,
  
  # --- Meat cattle based on forage proportions ---
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.75 ~ 0.07,
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop > 0.15 ~ 0.063,
  P.Product == "Cattle" & Purpose == "Meat" & forage_prop >= 0 ~ 0.04,
  P.Product == "Cattle" & Purpose == "Meat" & is.na(forage_prop) ~ 0.065,

  # --- Sheep & Lambs ---
  P.Product == "Sheep" & grepl("lamb", Herd.Stage, ignore.case = TRUE) ~ 0.045,
  P.Product == "Sheep" ~ 0.067,

  # --- Goats ---
  P.Product == "Goat" ~ 0.055,
  
  TRUE ~ NA_real_
),

    # Final daily intake (kg/day): pick the best available, but avoid using DC.Value
    final_intake_kg_day = case_when(
      !is.na(Total_Amount) & D.Unit.Amount == "kg" ~ Total_Amount,
      !is.na(sum_ingredient_intake) ~ sum_ingredient_intake,
      TRUE ~ NA_real_
    ),
    
    # Track source of intake
    final_intake_source = case_when(
      !is.na(Total_Amount) & D.Unit.Amount == "kg" ~ "Total_Amount",
      !is.na(sum_ingredient_intake) ~ "Total feed intake",
      TRUE ~ "Missing"
    ),
    
    # Daily GE intake (MJ/day) = GE content × daily intake
    GE_daily = case_when(
      !is.na(final_intake_kg_day) & !is.na(DC.Value) ~ DC.Value * final_intake_kg_day,
      TRUE ~ NA_real_
    ),
    
    # Annual CH₄ emissions
    CH4_kg_year = ifelse(
      !is.na(GE_daily) & !is.na(Ym),
      GE_daily * Ym * 365 / 55.65,
      NA_real_
    ),
    
    # Flag potential issues
    CH4_flag = case_when(
      is.na(DC.Value) ~ "missing_GE_value",
      is.na(Ym) ~ "missing_Ym",
      is.na(final_intake_kg_day) ~ "missing_final_intake",
      is.na(GE_daily) ~ "GE_not_calculated",
      is.na(CH4_kg_year) ~ "CH4_not_calculated",
      TRUE ~ "ok"
    )
  )


```



### 17.5) Final calculation

```{r,message=FALSE,warning=FALSE}
#combine
# 1. Tag and select relevant columns from both tables
GE_all_clean <- GE_all %>%
  mutate(source = "GE_all") %>%
  select(
    B.Code, 
    D.Item, 
    P.Product,
    Herd.Stage,
    GE_daily, 
    CH4_kg_year,
    Total_Amount,
    D.Unit.Amount,
    ED.Mean.T, 
    CH4_flag,
    Ym,
    source
  )

GE_reconstructed_clean <- GE_all_reconstructed %>%
  mutate(source = "Reconstructed") %>%
  rename(ED.Mean.T = final_intake_kg_day) %>%
  select(
    B.Code, 
    D.Item, 
    P.Product, 
    Herd.Stage,
    GE_daily, 
    CH4_kg_year, 
    Total_Amount,
    D.Unit.Amount,
    ED.Mean.T, 
    CH4_flag, 
    Ym,
    source
  )

GE_combined <- bind_rows(GE_all_clean, GE_reconstructed_clean, GE_diet_reconstructed) %>%
  mutate(
    source_priority = case_when(
      source == "GE_all" ~ 1,
      source == "Reconstructed" ~ 2,
      source == "GE_diet_from_ingredients" ~ 3,
      TRUE ~ 4
    ),
    has_CH4 = !is.na(CH4_kg_year)
  ) %>%
  arrange(B.Code, D.Item, desc(has_CH4), source_priority) %>%
  distinct(B.Code, D.Item, .keep_all = TRUE) %>%
  select(-has_CH4, -source_priority)


# Define non-emitting or pre-ruminant stages
non_emitting_stages <- c(
   "Kid", "Kid Male", "Calf Female", "Weaned Calf Male",
  "Young Male", "Young Female", 
   "Intact Kid Male"
)

# Filter them out
GE_combined <- GE_combined %>%
  filter(!(Herd.Stage %in% non_emitting_stages))

GE_combined <- GE_combined %>%
  left_join(ctrl_key, by = c("B.Code", "D.Item" = "A.Level.Name"))



GE_data <- GE_combined %>%
  filter(!is.na(GE_daily) & GE_daily != "") %>%
  distinct(B.Code)

# 3. Papers with GE and species (intersection)
GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

# Count unique B.Code with non-missing CH4 emissions
n_CH4_codes <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code) %>%
  nrow()

n_CH4_codes

n_CH4_codes_by_product <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "n_CH4_codes")

n_CH4_codes_by_product
```


```{r,message=FALSE,warning=FALSE}
# Step 1: Get list of B.Codes from GE_combined
relevant_BCodes <- GE_combined %>%
  distinct(B.Code)

# Step 2: Filter Animals.Diet.Comp for relevant B.Codes
diet_data_filtered <- merged_metadata$Animals.Diet %>%
  semi_join(relevant_BCodes, by = "B.Code")

# Step 3: Count unique diets and ingredients
total_diets <- diet_data_filtered %>%
  filter(!is.na(A.Level.Name)) %>%
  distinct(B.Code, A.Level.Name) %>%
  nrow()

total_ingredients <- diet_data_filtered %>%
  filter(!is.na(D.Item)) %>%
  distinct(B.Code, D.Item) %>%
  nrow()

# Step 4: Output
cat("Total unique diets:", total_diets, "\n")
cat("Total unique ingredients:", total_ingredients, "\n")

```

total number of codes that have ge and have both treatment and control 

```{r}
# Identify B.Codes that contain both control and treatment diets
b_codes_with_both_ctrl_and_trt <- GE_combined %>%
  filter(T.Control %in% c("yes", "no")) %>%   # Only if T.Control is properly labeled
  group_by(B.Code) %>%
  summarise(n_control_status = n_distinct(T.Control), .groups = "drop") %>%
  filter(n_control_status > 1)   # Keep only those with both Yes and No

# Count of such B.Codes
n_both_ctrl_and_trt <- nrow(b_codes_with_both_ctrl_and_trt)
cat("Number of B.Codes with both control and treatment diets:", n_both_ctrl_and_trt, "\n")

```

papers that have no tagg for t control 

```{r}
check_control<-ingredients%>%
  filter(is.na(T.Control))%>%
  select(B.Code,A.Level.Name,D.Item,T.Control)
```

filter ge combined with only papers with both control and treatments 

```{r}
GE_combined <- GE_combined %>%
  semi_join(b_codes_with_both_ctrl_and_trt, by = "B.Code")

GE_species <- merged_metadata$`Prod.Out` %>%
  filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
  semi_join(GE_data, by = "B.Code") %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "Studies_with_GE")

summary_table <- species_counts %>%
  full_join(GE_species, by = "P.Product") %>%
  mutate(across(everything(), ~replace_na(., 0)))

kable(summary_table, caption = "Summary of Data Availability for Enteric CH₄ Emissions (Tier 2)")

# Count unique B.Code with non-missing CH4 emissions
n_CH4_codes <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code) %>%
  nrow()

n_CH4_codes

n_CH4_codes_by_product <- GE_combined %>%
  filter(!is.na(CH4_kg_year)) %>%
  distinct(B.Code, P.Product) %>%
  count(P.Product, name = "n_CH4_codes")

n_CH4_codes_by_product

relevant_BCodes <- GE_combined %>%
  distinct(B.Code)

# Step 2: Filter Animals.Diet.Comp for relevant B.Codes
diet_data_filtered <- merged_metadata$Animals.Diet %>%
  semi_join(relevant_BCodes, by = "B.Code")

# Step 3: Count unique diets and ingredients
total_diets <- diet_data_filtered %>%
  filter(!is.na(A.Level.Name)) %>%
  distinct(B.Code, A.Level.Name) %>%
  nrow()

total_ingredients <- diet_data_filtered %>%
  filter(!is.na(D.Item)) %>%
  distinct(B.Code, D.Item) %>%
  nrow()

# Step 4: Output
cat("Total unique diets:", total_diets, "\n")
cat("Total unique ingredients:", total_ingredients, "\n")
```


## 18) Link emissions to productivity data 

here we will have the same issue as with feed intake data since the Alevel in the data tab dont always latch the ones in the diet tab 


### 18.1) Merge yield and emission data 

```{r}

# Step 1: Make sure the key columns match (D.Item and A.Level.Name)
GE_combined_for_merge <- GE_combined %>%
  rename(A.Level.Name = D.Item)%>%
  select(-"ED.Mean.T")

# Step 2: Merge yield and emission data
yield_with_emissions <- yield_data %>%
  left_join(
    GE_combined_for_merge,
    by = c("B.Code", "A.Level.Name")
  )


```

### 18.2) Plot the data
```{r}

# Filter meat yield data and drop missing values
yield_meat <- yield_with_emissions %>%
  filter(Out.Subind == "Meat Yield") %>%
  filter(!is.na(ED.Mean.T) & !is.na(CH4_kg_year) & !is.na(P.Product))

# Get axis limits
x_range <- range(yield_meat$ED.Mean.T, na.rm = TRUE)
y_range <- range(yield_meat$CH4_kg_year, na.rm = TRUE)

# Plot
ggplot(yield_meat, aes(x = ED.Mean.T, y = CH4_kg_year, color = P.Product)) +
  geom_point(alpha = 0.8, size = 2.5) +
  scale_x_log10(
    limits = x_range,
    breaks = scales::log_breaks(n = 5),
    labels = scales::comma_format()
  ) +
  scale_y_continuous(
    limits = y_range,
    labels = scales::comma_format()
  ) +
  labs(
    title = "CH4 Emissions vs. Meat Yield",
    x = "Meat Yield per Individual (kg)",
    y = "Annual CH4 Emissions (kg)",
    color = "Livestock Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")



```

```{r}
# Filter meat yield data and drop missing values
yield_milk <- yield_with_emissions %>%
  filter(Out.Subind == "Milk Yield") %>%
  filter(!is.na(ED.Mean.T) & !is.na(CH4_kg_year) & !is.na(P.Product))

# Get axis limits
x_range <- range(yield_milk$ED.Mean.T, na.rm = TRUE)
y_range <- range(yield_milk$CH4_kg_year, na.rm = TRUE)

# Plot
ggplot(yield_milk, aes(x = ED.Mean.T, y = CH4_kg_year, color = P.Product)) +
  geom_point(alpha = 0.8, size = 2.5) +
  scale_x_log10(
    limits = x_range,
    breaks = scales::log_breaks(n = 5),
    labels = scales::comma_format()
  ) +
  scale_y_continuous(
    limits = y_range,
    labels = scales::comma_format()
  ) +
  labs(
    title = "CH4 Emissions vs. Meat Yield",
    x = "Milk Yield per Individual (kg)",
    y = "Annual CH4 Emissions (kg)",
    color = "Livestock Species"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

```


```{r}
# Filter and clean data
emissions_by_species <- yield_with_emissions %>%
  filter(!is.na(CH4_kg_year), !is.na(P.Product))

# Boxplot with transparent points
ggplot(emissions_by_species, aes(x = P.Product, y = CH4_kg_year, fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +  # lighter fill and no outliers
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 2) +  # transparent points
  labs(
    title = "Annual CH4 Emissions by Livestock Species",
    x = "Species",
    y = "Annual CH4 Emissions (kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")  # remove redundant legend

```


```{r}
# Filter and clean data for Goat and Sheep only
emissions_by_species <- yield_with_emissions %>%
  filter(!is.na(CH4_kg_year), P.Product %in% c("Goat", "Sheep"))

# Boxplot with transparent points for Goat and Sheep
ggplot(emissions_by_species, aes(x = P.Product, y = CH4_kg_year, fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.2, alpha = 0.5, color = "black", size = 2) +
  labs(
    title = "Annual CH4 Emissions: Goat vs. Sheep",
    x = "Species",
    y = "Annual CH4 Emissions (kg)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```



```{r}
## 1) Live‑weight table --------------------------------------------------------
bw_tbl <- weights_unique %>% 
  select(B.Code, LiveWeight_kg = Out.WG.Start) %>% 
  filter(!is.na(LiveWeight_kg))

## 2) Merge with CH₄ and keep P.Product already present -----------------------
ch4_vs_bw <- GE_combined %>%                       # contains P.Product
  filter(!is.na(CH4_kg_year)) %>% 
  left_join(bw_tbl, by = "B.Code") %>% 
  filter(!is.na(LiveWeight_kg) & LiveWeight_kg > 0)

## 3) Plot ---------------------------------------------------------------------
##########################################################
## CH4 vs. live‑weight with 95 % confidence intervals   ##
##########################################################

ggplot(ch4_vs_bw,
       aes(x = LiveWeight_kg, y = CH4_kg_year, colour = P.Product)) +
  geom_point(alpha = 0.65, size = 2.3) +
  geom_smooth(method  = "lm",
              formula = y ~ log10(x),  # linear in log‑weight
              se      = TRUE,          # << confidence ribbon
              linetype = "dashed",
              linewidth = 1) +
  scale_x_log10(breaks = scales::log_breaks(), labels = scales::comma) +
  labs(
    title  = "Annual CH4 emissions vs. live weight",
    x      = "Body weight (kg, log scale)",
    y      = "CH₄  (kg animal⁻¹ year⁻¹)",
    colour = "Species"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")



```
```{r}
wg<-merged_metadata$Data.Out%>%filter(Out.Subind=="Weight Gain")
###############################################################################
##  STANDARDISE WEIGHT-GAIN TO kg animal⁻¹ day⁻¹  (no totals, no wk/mo)      ##
###############################################################################
wg_clean <- wg %>%
  ## 1. normalise text ---------------------------------------------------------
  mutate(
    unit_raw = tolower(trimws(Out.Unit)),
    value    = as.numeric(ED.Mean.T)                 # be sure it's numeric
  ) %>%
  
  ## 2. discard relative indices (% IBW, % SGR) -------------------------------
  filter(
    !str_detect(unit_raw, "% ibw") &
    !str_detect(unit_raw, "% sgr")
  ) %>%
  
  ## 3. keep only entries that already contain “/d” or “/day” ------------------
  filter(str_detect(unit_raw, "/d") | str_detect(unit_raw, "/day") |
         str_detect(unit_raw, "g/kg metabolic weight")) %>%
  
  ## 4. parse amount unit ------------------------------------------------------
  mutate(
    amount_unit = case_when(
      str_starts(unit_raw, "mg") ~ "mg",
      str_starts(unit_raw, "g")  ~ "g",
      str_starts(unit_raw, "kg") ~ "kg",
      TRUE                       ~ NA_character_
    ),
    gain_kg = case_when(
      amount_unit == "mg" ~ value / 1e6,
      amount_unit == "g"  ~ value / 1e3,
      amount_unit == "kg" ~ value,
      TRUE                ~ NA_real_
    )
  ) %>%
  
  ## 5. handle “g/kg metabolic weight” -----------------------------------------
  left_join(
    weights_unique %>% 
      select(B.Code, A.Level.Name, BW_start_kg = Out.WG.Start),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  mutate(
    ADG_kg = case_when(
      str_detect(unit_raw, "g/kg metabolic weight") &
        !is.na(value) & !is.na(BW_start_kg)
          ~ (value / 1000) * (BW_start_kg ^ 0.75),   # g → kg × BW^0.75
      TRUE ~ gain_kg                                # already kg/d
    )
  ) %>%
  
  ## 6. final screen -----------------------------------------------------------
  filter(!is.na(ADG_kg) & ADG_kg > 0) %>%
  select(B.Code, A.Level.Name, ADG_kg)

```

```{r}
############################################################
##  EMISSION INTENSITY  (kg CH₄  kg-¹ live-weight gain)   ##
############################################################


## 1. merge CH₄ and ADG ------------------------------------
ei_df <- GE_combined %>%                    # CH₄ table
  select(B.Code, A.Level.Name = D.Item,
         CH4_kg_year, P.Product) %>%
  inner_join(
    wg_clean,                               # weight-gain table
    by = c("B.Code", "A.Level.Name")
  ) %>%

  ## 2. compute emission intensity -------------------------
  mutate(
    gain_kg_year = ADG_kg * 365,                           # kg yr-¹
    EI_CH4_kg = CH4_kg_year / gain_kg_year
  ) %>%

  ## 3. basic QC  ------------------------------------------
  filter(
    !is.na(EI_CH4_kg) & EI_CH4_kg > 0,          # remove missing / negative
    EI_CH4_kg < 1                               # drop extreme outliers (>1 kg CH₄ kg-¹ LW-gain)
  )

length(unique(ei_df$B.Code))

# A) box-and-jitter by species -----------------------------------------------
ggplot(ei_df, aes(x = P.Product, y = EI_CH4_kg, fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.35) +
  geom_jitter(width = 0.15, alpha = 0.65, size = 2) +
  scale_y_continuous("Emission intensity (kg CH₄ kg⁻¹ LW-gain)",
                     labels = scales::comma_format()) +
  labs(title = "Enteric-methane intensity of growth",
       x = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# B) relationship between growth rate and EI ---------------------------------
ggplot(ei_df, aes(x = ADG_kg, y = EI_CH4_kg, colour = P.Product)) +
  geom_point(alpha = 0.7, size = 2.3) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") +
  scale_x_log10("Average daily gain (kg animal⁻¹ day⁻¹)",
                breaks = scales::log_breaks(),
                labels = scales::comma_format()) +
  scale_y_continuous("Emission intensity (kg CH₄ kg⁻¹ LW-gain)",
                     labels = scales::comma_format()) +
  labs(title = "Faster-growing animals generally emit less CH₄ per kg gain",
       colour = "Species") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

```

```{r,warning=FALSE}
library(dplyr); library(ggplot2); library(scales)
# 0.  Make sure your weight-gain table is available
#     (this is the object you built in the earlier step)
adg_clean <- wg_clean %>%                 # <- rename for clarity
  select(B.Code, A.Level.Name, ADG_kg)    # keep only the essentials
# 1.  Bring EI together with the absolute-CH₄ table you already built
ei_tbl <- adg_clean %>%                           # has A.Level.Name
  inner_join(
    GE_combined, 
    by = c("B.Code"       = "B.Code",             # same name
           "A.Level.Name" = "D.Item")             # different names
  ) %>% 
  mutate(
    EI = CH4_kg_year / (ADG_kg * 365)             # kg CH₄ kg⁻¹ LW-gain
  )


# 2.  Median split lines (or choose any threshold you like)
vline <- median(ei_tbl$CH4_kg_year,   na.rm = TRUE)
hline <- median(ei_tbl$EI,            na.rm = TRUE)

x_lim <- range(ei_tbl$ADG_kg,  na.rm = TRUE)   # e.g. 0.002 – 0.8
y_lim <- range(ei_tbl$EI,      na.rm = TRUE)   # e.g. 0.005 – 1.2
ch4_lim <- range(ei_tbl$CH4_kg_year, na.rm = TRUE)


# 3.  Plot ----------------------------------------------------------------
ggplot(ei_tbl,
       aes(x = ADG_kg,
           y = EI,
           colour = CH4_kg_year,
           size   = CH4_kg_year)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, colour = "black") +
  scale_x_log10(limits = x_lim,
                breaks  = log_breaks(),
                labels  = comma_format()) +
  scale_y_log10(limits = y_lim,
                breaks  = c(.005, .01, .03, .1, .3, 1),
                labels  = comma_format()) +
  scale_colour_viridis_c(option = "plasma",
                         trans   = "log10",
                         limits  = ch4_lim,
                         name    = "Annual CH₄ (kg)") +
  scale_size(range = c(2, 6),
             trans  = "log10",
             limits = ch4_lim,
             guide  = "none") +
  facet_wrap(~ P.Product) +                # same axes in every facet
  labs(title   = "Faster-growing animals emit less CH4 per kg gain",
       subtitle = "Axes are identical across cattle, goat and sheep panels",
       x = "Average daily gain (kg animal⁻¹ day⁻¹, log scale)",
       y = "Emission intensity (kg CH4 kg⁻¹ live-weight gain)",
       caption = "Point colour & size scale with absolute annual CH4 per animal") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")

```

```{r}
##############################################
##  1.  PREPARE DAILY YIELD (kg animal-1 d-1)
##############################################

## density of milk for L → kg conversion
MILK_DENSITY <- 1.03   # kg L-1

yield_d_clean <- yield_data %>%                      # ← your harmonised table
  filter(Out.Subind %in% c("Milk Yield", "Meat Yield")) %>%
  mutate(
    Yield_kg_day = case_when(
      # ───────────── Milk ─────────────────────────────────────────────
      Out.Subind == "Milk Yield" & Out.Unit == "l/individual/day"  ~ ED.Mean.T * MILK_DENSITY,
      Out.Subind == "Milk Yield" & Out.Unit == "kg/individual/day" ~ ED.Mean.T,
      
      # ───────────── Meat (already kg day-1) ─────────────────────────
      Out.Subind == "Meat Yield" & Out.Unit == "kg/individual/day" ~ ED.Mean.T,
      
      # ───────────── Everything else (drop) ─────────────────────────
      TRUE ~ NA_real_
    )
  ) %>%
  filter(!is.na(Yield_kg_day) & Yield_kg_day > 0) %>%   # keep usable rows only
  select(B.Code, A.Level.Name, Out.Subind, Yield_kg_day)


###########################################################################
##  2.  COMPUTE EMISSION-INTENSITY  (kg CH4 per kg *product*) #############
###########################################################################

ei_prod <- GE_combined %>%                              # absolute CH4 table
  select(B.Code, A.Level.Name = D.Item,
         CH4_kg_year, P.Product,T.Control) %>%
  inner_join(yield_d_clean, by = c("B.Code","A.Level.Name")) %>%
  mutate(
    Yield_kg_year = Yield_kg_day * 365,
    EI_CH4_prod   = CH4_kg_year / Yield_kg_year,
    Product       = if_else(Out.Subind == "Milk Yield", "Milk", "Meat")
  ) %>%
  filter(EI_CH4_prod > 0, EI_CH4_prod < 10)             # drop negatives / crazy outliers

#########################################
##  3.  QUICK DIAGNOSTIC PLOTS ##########
#########################################

# ── A) Box-and-jitter by species & product ───────────────────────────────
ggplot(ei_prod, aes(x = P.Product, y = EI_CH4_prod,
                    fill = P.Product)) +
  geom_boxplot(outlier.shape = NA, alpha = .35) +
  geom_jitter(width = .15, alpha = .65, size = 2) +
  scale_y_log10("Emission-intensity (kg CH₄ kg⁻¹ product)",
                breaks = scales::log_breaks(),
                labels = scales::comma_format()) +
  facet_wrap(~ Product, nrow = 1) +
  labs(title = "Enteric-methane intensity of milk and meat production",
       x = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

# ── B) Relationship between yield and EI ────────────────────────────────
ggplot(ei_prod,
       aes(x = Yield_kg_day,
           y = EI_CH4_prod,
           colour = P.Product)) +
  geom_point(alpha = .7, size = 2.3) +
  geom_smooth(method = "lm", se = TRUE, linetype = "dashed") +
  scale_x_log10("Yield (kg animal⁻¹ day⁻¹, log scale)",
                breaks = scales::log_breaks(),
                labels = scales::comma_format()) +
  scale_y_log10("Emission-intensity (kg CH₄ kg⁻¹ product)",
                breaks = scales::log_breaks(),
                labels = scales::comma_format()) +
  facet_wrap(~ Product, nrow = 1) +
  labs(title  = "Higher-yielding animals emit less CH₄ per kg of product",
       colour = "Species") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom")

length(unique(ei_prod$B.Code))

```

## Analysis based on practices 

```{r}
# ── 1) get the detailed “added‐ingredient” changes per diet ────────────────
practice_changes <- classify_practice_changes(ingredients)
# returns cols: B.Code, A.Level.Name, added, D.Type, compare_to, other_diet
practice_changes_classified <- practice_changes %>%
  mutate(
    D.Type.Class = case_when(
      str_detect(D.Type, regex("concentrate", ignore_case = TRUE)) ~ "Concentrate",
      str_detect(D.Type, regex("supplement", ignore_case = TRUE)) ~ "Supplement",
      TRUE ~ D.Type
    ))%>%
       mutate(
    D.Item = case_when(
      str_detect(D.Item, regex("Salt", ignore_case = TRUE)) ~ "Salt",
      TRUE ~ D.Item
    )
  )

# Step 2: Create practice_summary using the cleaned D.Type.Class
practice_summary <- practice_changes_classified %>% 
  group_by(B.Code, A.Level.Name, compare_to) %>%     # ← add compare_to here
  summarise(
    practice_type = first(na.omit(practice_type)),

    added_items   = paste(unique(D.Item[change_type == "added"   ]), 
                          collapse = "; "),
    removed_items = paste(unique(D.Item[change_type == "removed" ]), 
                          collapse = "; "),

    added_types   = paste(unique(D.Type.Class[change_type == "added"   ]), 
                          collapse = "; "),
    removed_types = paste(unique(D.Type.Class[change_type == "removed" ]), 
                          collapse = "; "),
    .groups = "drop"
  )




```


```{r}

# Option A: if GE_combined still has diet in D.Item
GE_with_practice <- GE_combined %>%
  rename(A.Level.Name = D.Item) %>%
  left_join(practice_summary, by = c("B.Code", "A.Level.Name"))

ei_plot_df_expanded <- GE_with_practice %>%
  left_join(
    ei_prod %>% select(B.Code, A.Level.Name, EI_CH4_prod, Product),
    by = c("B.Code", "A.Level.Name")
  ) %>%
  filter(T.Control %in% c("no", "false", "0", FALSE)) %>%
  filter(!is.na(EI_CH4_prod)) %>%
  filter(!is.na(practice_type)) %>%
  separate_rows(added_types, sep = ";\\s*") # Splits on "; " and expands rows

# Now plot
ggplot(ei_plot_df_expanded, aes(x = added_types, y = EI_CH4_prod, fill= practice_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.4) +
  geom_jitter(width = 0.2, size = 2, color = "black", alpha = 0.6) +
  facet_wrap(~ Product, scales = "fixed") +
  scale_y_log10("Emission intensity (kg CH₄ / kg product)",
                breaks = log_breaks(), labels = comma_format()) +
  labs(
    title = "CH₄ Emission Intensity by Ingredient Category",
    x     = "Ingredient type",
    fill  = "Practice"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, hjust = 1))

```
salt to eb salt 

```{r}
# Total number of plotted points
n_total <- nrow(ei_plot_df_expanded)

# Count by comparison type
comparison_counts <- ei_plot_df_expanded %>%
  count(compare_to, practice_type)

# View the results
print(paste("Total points in plot:", n_total))
print(comparison_counts)

```

look non era feed and other ingredients those could maybe classified other ways 

```{r}
non_era_ing <- ingredients %>%                 # ⬅️ all “Non-ERA Feed”
  filter(D.Type == "Non-ERA Feed") %>% 
  pull(D.Item) %>% 
  unique()                                      # character vector

other_ing   <- ingredients %>%                 # ⬅️ all “Other ingredients”
  filter(D.Type == "Other Feed" |              #  ← or whatever your label is
         D.Type == "Other Ingredient") %>%     #  ← add more ORs if needed
  pull(D.Item) %>% 
  unique()
```



```{r}
ctrl_tbl <- ei_prod %>% 
  filter(T.Control %in% c("yes", TRUE)) %>%   # keep ONLY control diets
  distinct(B.Code, .keep_all = TRUE) %>%      # make sure one control per study
  select(B.Code, ctrl_EI = EI_CH4_prod)       # keep just the two columns


## ── Build ONE tidy table of category–substitutions ──────────────────────────
subs_df <- GE_with_practice %>%                       # practice info
  left_join(ei_prod  %>% select(B.Code, A.Level.Name,
                                EI_CH4_prod, Product, T.Control),
            by = c("B.Code","A.Level.Name")) %>% 
  distinct(B.Code, A.Level.Name, .keep_all = TRUE) %>%       # de-dup
  filter(compare_to == "control",        # only trt vs. *its* control
         !is.na(practice_type)) %>% 
  left_join(ctrl_tbl, by = "B.Code") %>%                     # add ctrl EI
  mutate(delta_EI = EI_CH4_prod - ctrl_EI) %>%               # + = ↑ EI
  ## explode to one row per *category pair* -------------------------------
  separate_rows(added_types,   sep = ";\\s*") %>% 
  separate_rows(removed_types, sep = ";\\s*") %>% 
  transmute(B.Code, Product,
            removed = trimws(removed_types),
            added   = trimws(added_types),
            delta_EI)%>%
  filter(!is.na(delta_EI))


```
```{r}

heat_bubble <- subs_df %>% 
  group_by(removed, added) %>% 
  summarise(n      = n(),
            mean_d = mean(delta_EI, na.rm = TRUE),
            .groups = "drop") %>% 
  filter(!is.na(removed) & !is.na(added))

ggplot(heat_bubble, aes(removed, added)) +
  geom_point(aes(size = n, fill = mean_d),
             shape = 21, colour = "grey30") +
  scale_size(range = c(2, 14), breaks = c(1, 3, 5, 10)) +
  scale_fill_gradient2(low = "#2b83ba", high = "#d7191c",
                       mid = "white", midpoint = 0,
                       name = "ΔEI") +
  labs(title = "How category substitutions shift EI",
       x = "Category removed", y = "Category added") +
  theme_minimal(base_size = 12) +
  theme(axis.text.x  = element_text(angle = 45, hjust = 1),
        panel.grid   = element_blank())

```

```{r}
library(ggpubr)   # gives geom_errorbarh if you prefer base ggplot

forest_df <- subs_df %>% 
  group_by(removed, added) %>% 
  summarise(n   = n(),
            d   = mean(delta_EI, na.rm = TRUE),
            se  = sd(delta_EI)/sqrt(n),
            lo  = d - 1.96*se,
            hi  = d + 1.96*se,
            .groups = "drop") %>% 
  filter(n >= 3)   # keep only pairs with ≥3 trials

ggplot(forest_df,
       aes(y = reorder(paste(removed, "→", added), d),
           x = d)) +
  geom_point() +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  scale_x_continuous("Mean ΔEI  (kg CH₄ / kg product)") +
  labs(y = NULL, title = "Average effect of common substitutions") +
  theme_minimal(base_size = 11)


```
```{r}
add_effect <- subs_df %>%
  filter(removed == "" | is.na(removed)) %>%     # pure additions
  group_by(added) %>%
  summarise(n = n(),
            mean_d = mean(delta_EI),
            .groups = "drop")

ggplot(add_effect, aes(x = reorder(added, mean_d), y = mean_d,
                       fill = mean_d > 0)) +
  geom_col() + coord_flip() +
  scale_fill_manual(values = c("#2b83ba", "#d7191c"),
                    guide = "none") +
  labs(x = NULL, y = "ΔEI", title = "Effect of adding a single category") +
  theme_minimal()

```


```{r}

################################################################################
##  Join “changes” with emission-intensity of BOTH diets in every comparison ##
###############################################################################

## keep only the columns we need from ei_prod
ei_core <- ei_prod %>% 
  select(B.Code,
         A.Level.Name,           # diet name
         EI_CH4_prod,            # kg CH4 · kg-product⁻¹
         CH4_kg_year,            # (optional, if you want absolute CH4 later)
         Product, P.Product,     # meat / milk & species
         T.Control,              # yes / no
         Yield_kg_day, Yield_kg_year)   # keep yields if you like

ei_pairs <- changes %>% 
  ## ---- attach intensity for diet_A -----------------------------------------
  left_join(ei_core,
            by = c("B.Code", "diet_A" = "A.Level.Name"),
            suffix = c("", "_A")) %>%          # tmp suffix to avoid clashes
  rename(EI_A = EI_CH4_prod) %>%               # keep only the field we need
  ## ---- attach intensity for diet_B -----------------------------------------
  left_join(ei_core,
            by = c("B.Code", "diet_B" = "A.Level.Name"),
            suffix = c("_A", "_B")) %>%        # second join, new suffix
  rename(EI_B = EI_CH4_prod) %>%               
  ## ---- derive deltas & practice magnitude -----------------------------------
  mutate(
    delta_EI_prod = EI_B - EI_A,               # +  ⇒ diet_B is worse
    abs_change    = coalesce(total_pct_added,    0) +
                    coalesce(total_pct_removed,  0) +
                    coalesce(total_pct_increase, 0) +
                    coalesce(total_pct_decrease, 0),
    pair_label    = paste(diet_A, "→", diet_B)
  ) %>% 
  ## keep only rows where we have both intensities
  filter(!is.na(delta_EI_prod))
            # keep only pairs with EI on both sides

```
```{r}
###################################
##  Practice-vs-EI plotting chunk
###################################

library(tidyverse)
library(patchwork)               # install.packages("patchwork") once

## ───────────────── 0 · Prep  ────────────────────────────────────────────────
ei_pairs_plot <- ei_pairs %>% 
  # 0·a  choose one species/product label
  mutate(Product = coalesce(Product_B, Product_A,
                            P.Product_B, P.Product_A)) %>% 
  # 0·b  compute yield delta before selecting
  mutate(delta_Yield = Yield_kg_day_B - Yield_kg_day_A) %>% 
  select(abs_change,
         delta_EI_prod,
         delta_Yield,
         practice_type,
         Product) %>% 
  filter(!is.na(Product),
         !is.na(delta_EI_prod))

## ───────────────── 1 · scatter, 2 · violin, 3 · trade-off ──────────────────
p1 <- ggplot(ei_pairs_plot,
             aes(abs_change, delta_EI_prod, colour = practice_type)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = .7, size = 2) +
  geom_smooth(method = "lm", se = TRUE, colour = "black",
              formula = y ~ x) +
  scale_x_continuous("% of diet DM changed") +
  scale_y_continuous("Δ EI (kg CH₄ / kg product)") +
  facet_wrap(~ Product) +
  theme_bw() +
  theme(legend.position = "bottom")

p2 <- ggplot(ei_pairs_plot,
             aes(practice_type, delta_EI_prod, fill = practice_type)) +
  geom_violin(trim = FALSE, alpha = .35, colour = NA) +
  geom_boxplot(width = .18, outlier.shape = NA) +
  stat_summary(fun = median, geom = "point",
               size = 2, colour = "black") +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey40") +
  scale_y_continuous("Δ EI (kg CH₄ / kg product)") +
  facet_wrap(~ Product, nrow = 1) +
  labs(x = NULL, title = "Emission-intensity shift by practice type") +
  theme_bw() +
  theme(legend.position = "none")

p3 <- ggplot(ei_pairs_plot,
             aes(delta_Yield, delta_EI_prod,
                 colour = practice_type,
                 size   = abs_change)) +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey60") +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey60") +
  geom_point(alpha = .7) +
  scale_x_continuous("Δ Yield (kg animal⁻¹ d⁻¹)",
                     labels = scales::comma) +
  scale_y_continuous("Δ EI (kg CH₄ / kg product)",
                     labels = scales::comma) +
  scale_size(range = c(2,6), name = "% moved") +
  facet_wrap(~ Product) +
  labs(title = "Productivity vs. emission-intensity trade-offs") +
  theme_minimal() +
  theme(legend.position = "bottom")

## ───────────────── 2 · show or combine ──────────────────────────────────────
p1
p2
p3
# (p1 | p2 | p3) + plot_layout(guides = "collect")   # uncomment for one panel


```
```{r}
library(tidyverse)

## ── 1 · tidy table -----------------------------------------------------------
plot_df <- ei_pairs %>% 
  filter(practice_type %in% c("addition", "substitution")) %>%     # only add / sub
  mutate(abs_change = total_pct_added + total_pct_removed +
                     total_pct_increase + total_pct_decrease) %>%  # magnitude
  
  separate_rows(added_by_type, sep = ";\\s*") %>%                  # explode list
  
  mutate(
    Ingredient_Type = str_remove(added_by_type, ":.*$") %>% str_trim(),
    Ingredient_Type = case_when(                                   # collapse names
      str_detect(Ingredient_Type, regex("supplement",  TRUE)) ~ "Supplement",
      str_detect(Ingredient_Type, regex("concentrate", TRUE)) ~ "Concentrate",
      TRUE                                                   ~ Ingredient_Type
    )
  ) %>% 
  filter(Ingredient_Type != "", Ingredient_Type != "Entire Diet") %>% 
  
  mutate(Product = coalesce(!!!syms(names(.)[str_detect(names(.), "Product")]))) %>% 
  
  select(Product, practice_type, Ingredient_Type,
         abs_change, delta_EI_prod)

## ── 2 · bubble panel ---------------------------------------------------------
ggplot(plot_df,
       aes(abs_change, delta_EI_prod,
           colour = Ingredient_Type,
           shape  = practice_type,
           size   = abs_change)) +
  geom_hline(yintercept = 0, linetype = 2, colour = "grey50") +
  geom_point(alpha = .75) +
  scale_size(range = c(3,10), guide = "none") +
  scale_x_continuous("% of diet DM changed") +
  scale_y_continuous("Δ EI (kg CH₄ / kg product)",
                     labels = scales::comma) +
  facet_wrap(~ Product) +
  labs(title  = "Effect of adding / substituting single ingredient categories",
       colour = "Ingredient category",
       shape  = "Practice") +
  theme_bw() +
  theme(legend.position = "right")


```
```{r}
library(tidyverse)
library(ggalt)     # install.packages("ggalt")  for geom_dumbbell()


## ── 1 · tidy substitutions table --------------------------------------------
subs_db <- ei_pairs %>%
  filter(practice_type == "substitution") %>% 
  
  # explode the semicolon lists
  separate_rows(removed_by_type, added_by_type, sep = ";\\s*") %>% 
  
  # trim & strip trailing numbers, then normalise categories
  mutate(
    removed = removed_by_type %>% str_remove(":.*$") %>% str_trim() %>% str_to_lower(),
    added   = added_by_type   %>% str_remove(":.*$") %>% str_trim() %>% str_to_lower(),
    
    removed = case_when(
      str_detect(removed, "supplement")   ~ "Supplement",
      str_detect(removed, "concentrate")  ~ "Concentrate",
      TRUE                                ~ str_to_title(removed)
    ),
    added = case_when(
      str_detect(added, "supplement")     ~ "Supplement",
      str_detect(added, "concentrate")    ~ "Concentrate",
      TRUE                                ~ str_to_title(added)
    )
  ) %>% 
  
  # drop blanks & “Entire Diet”
  filter(removed != "", added != "",
         removed != "Entire Diet", added != "Entire Diet") %>% 
  
  # pair label + product label
  mutate(
    pair_id = paste(removed, "→", added),
    Product = coalesce(!!!syms(names(.)[str_detect(names(.), "Product")]))
  )

## ── 2 · dumb-bell plot -------------------------------------------------------
subs_db <- subs_db %>%                         # reuse the table
  mutate(effect = case_when(
    delta_EI_prod <= -0.02 ~ "Lower EI",
    delta_EI_prod >=  0.02 ~ "Higher EI",
    TRUE                   ~ "Neutral"
  ))

library(tidyverse)

## subs_db already built & classified with “effect” ----
## left  dot  = original EI  (grey)
## right dot  = new EI       (colour by effect)

ggplot(subs_db,
       aes(y = reorder(pair_id, delta_EI_prod))) +

  ## segment
  geom_segment(aes(x = 0, xend = delta_EI_prod),
               colour = "grey70") +

  ## left (before) dot
  geom_point(aes(x = 0),
             colour = "grey60",
             size   = 2) +

  ## right (after) dot – coloured by effect
  geom_point(aes(x = delta_EI_prod,
                 colour = effect),
             size = 2) +

  scale_colour_manual(values = c(
    "Lower EI"  = "#2b83ba",   # blue  = improvement
    "Neutral"   = "grey70",
    "Higher EI" = "#d7191c"    # red   = worse
  )) +

  facet_wrap(~ Product, scales = "free_y") +
  geom_vline(xintercept = 0, linetype = 2, colour = "grey50") +
  labs(x = "Δ EI (kg CH₄ / kg product)",
       y = NULL,
       colour = "Effect",
       title = "Direction & magnitude of each substitution") +
  theme_minimal()


```

add intensity of the substitution. how are subtitution done ? removed and added werid concentrate thing. 
do it by species ? how many pooints ? 