---
title: "8_Emission_Ingredient_classification"
output: html_document
date: "2025-09-26"
---

```{r}
library(ERAg)
library(stringr)
library(knitr)
library(dplyr)
library(DT)
library(readxl)
library(tidyr)
library(arrow)
library(ggplot2)
library(data.table)
library(purrr)
library(janitor)
library(ggalt)

ingredients <- readRDS("~/ERL/Emission_Analysis/data/ingredients_harmonized_subset.rds")
merged_metadata <- readRDS("~/ERL/Emission_Analysis/data/merged_metadata_nutrition_inference.rds")
diet_percentages <- read.csv("~/ERL/Emission_Analysis/data/diet_percentages.csv")


```

## 18. Ingredient Classification methods 


| **Method** | **R function** | **Core concept** | **Variables needed** | **One‑sentence rule** |
|------------|----------------|------------------|----------------------|-----------------------|
| **1** | `method1()` | Exact‑window filter | GE, Ash, CP, EE, NDF<br>(*all present*) | Assign to the *first* category in which **every variable** falls inside its strict baseline window. |
| **2** | `method2()` | Exact ± 10 % filter | Same five | Same as M1 but each numeric bound is widened ± 10 %. |
| **3** | `method3()` | Priority‑variable filter (strict) | Only the **priority** variables per category:<br> • I → NDF<br> • II → EE<br> • III–V → CP & NDF<br> • VI → Ash | If all priority variables exist *and* meet their strict windows, assign that category; missing non‑priority data are ignored. |
| **4** | `method4()` | Priority‑variable ± 10 % | Same | Priority windows widened ± 10 %; non‑priority values (if present) still use strict windows and can only disqualify. |
| **5** | `method5()` | Hierarchical decision tree (strict) | Any subset of GE, Ash, CP, EE, NDF | Gates tested in order:<br>`EE ≥ 80 → II` → else `Ash ≥ 50 → VI` → else `CP ≥ 35 & NDF ≤ 30 → III` → else `NDF ≤ 20 → I` → else `NDF ≥ 60 → V` → else `40 ≤ NDF ≤ 50 → IV` → else *unclassified*. |
| **6** | `method6()` | Hierarchical tree ± 10 % | Same | Same gates as M5 but every numeric constant is multiplied by 0.9 (lower) and 1.1 (upper) before testing. |
| **7** | `method7()` | Hierarchical ± 10 % plus text overrides | Same numerics **+** `D.Item`, `D.Type` | Start with Method 6 result, then override:<br> • `oil|tallow|fat|grease` in *D.Item* → **II**<br> • protein‑meal regex (distillers grain, soybean meal, …) → **III**<br> • `D.Type` ∈ “Supplement” *or* matches “supplement” → **VI** |

*All concentration variables are in g·kg⁻¹ DM except GE (MJ·kg⁻¹ DM).*


## 18.1. Hierachical decision tree

```{r,echo=FALSE,warning=FALSE}
# --------------------------------------------------------------
# UPDATED FLOW‑CHART  (no NFC test)  –  Method 5 decision tree
# --------------------------------------------------------------
#  Install DiagrammeR if you do not have it:
#  install.packages("DiagrammeR")

library(DiagrammeR)

feed_tree_nfc_removed <- grViz("
digraph feed_tree {
  graph [layout = dot, rankdir = TB]

  # ----- Default styling -----
  node [shape = rectangle,
        style = filled,
        fillcolor = LightYellow,
        fontname = Helvetica,
        fontsize = 12]

  # ----- Decision nodes -----
  n1  [label = 'EE ≥ 80 ?']
  n3  [label = 'Ash ≥ 50 ?']
  n5  [label = 'CP ≥ 35 &\\nNDF ≤ 30 ?']
  n7  [label = 'NDF ≤ 20 ?']
  n9  [label = 'NDF ≥ 60 ?']
  n11 [label = '40 ≤ NDF ≤ 50 ?']

  # ----- Terminal nodes -----
  node [fillcolor = LightSkyBlue]
  n2  [label = 'Category II\\nHigh‑fat']
  n4  [label = 'Category VI\\nMineral/Vit']
  n6  [label = 'Category III\\nProtein conc.']
  n8  [label = 'Category I\\nHigh‑starch']
  n10 [label = 'Category V\\nPoor forage']
  n12 [label = 'Category IV\\nGood forage']
  n13 [label = 'Unclassified']

  # ----- Edges -----
  n1  -> n2  [label = 'Yes']
  n1  -> n3  [label = 'No']
  n3  -> n4  [label = 'Yes']
  n3  -> n5  [label = 'No']
  n5  -> n6  [label = 'Yes']
  n5  -> n7  [label = 'No']
  n7  -> n8  [label = 'Yes']
  n7  -> n9  [label = 'No']
  n9  -> n10 [label = 'Yes']
  n9  -> n11 [label = 'No']
  n11 -> n12 [label = 'Yes']
  n11 -> n13 [label = 'No']
}
")

# Render in the Viewer or export as needed
feed_tree_nfc_removed

# Example export:
# export_graph(feed_tree_nfc_removed, file_name = "feed_tree_method5.png", file_type = "PNG")


```


```{r, warning=FALSE,message=FALSE}

library(dplyr)
library(stringr)

Ing_classif <- merged_metadata$Animals.Diet.Comp

diets <- ingredients

Ing_classif <- Ing_classif %>%
  dplyr::filter(!.data$D.Item %in% c("Salt", "Water"))


# mapping: variant (keep name) → donor (copy nutrients from here)
nutrient_donor_map <- tribble(
  ~variant,                                        ~donor,
  "Leucaena leucocephala Dried Ground",            "Leucaena leucocephala Leaves Ground",
  "Sheanut Cake Boiled Dried",                     "Sheanut Cake Ground",
  "maize straw dried ensiled ground urea treated", "maize straw ensiled urea treated",
  "leucaena leucocephala leaves dried",            "leucaena leucocephala leaves ground",
  "gliciridia sepium ground",                      "gliciridia sepium leaves dried ground",
  "sunflower",                                     "sunflower seed",
  "moringa stenopetala leaves dried",              "moringa stenopetala dried",
  "sesbania sesban leaves dried",                  "sesbania sesban leaves and twigs dried",
  "urochloa hybrid mulato ii dried",               "urochloa mosambicensis dried",
  "gliciridia sepium leaves ground",               "gliciridia sepium leaves dried ground",
  "leucaena leucocephala leaves dried ground",     "leucaena leucocephala leaves ground",
  "wheat straw effective microbes",                "wheat straw",
  "maize & wheat offal","maize offal",
  "leucaena leucocephala dried","leucaena leucocephala leaves ground"

)

# Normalize to lowercase for matching
nutrient_donor_map <- nutrient_donor_map %>%
  mutate(across(everything(), tolower))

Ing_classif <- Ing_classif %>%
  mutate(D.Item_lc = tolower(D.Item))

# --- Step 1: extract donor profiles
donor_profiles <- Ing_classif %>%
  mutate(D.Item_lc = tolower(D.Item)) %>%
  filter(D.Item_lc %in% nutrient_donor_map$donor)

# --- Step 2: build replacement rows
replacement_rows <- nutrient_donor_map %>%
  rename(Donor_lc = donor, Variant_lc = variant) %>%
  inner_join(donor_profiles, by = c("Donor_lc" = "D.Item_lc")) %>%
  mutate(
    D.Item_lc = Variant_lc,        # replace with variant name
    D.Item    = str_to_title(Variant_lc)  # keep nicer capitalization
  )

# --- Step 3: remove old variant rows and add replacements
Ing_classif <- Ing_classif %>%
  filter(!D.Item_lc %in% nutrient_donor_map$variant) %>%  # drop old variants
  bind_rows(replacement_rows %>% select(names(Ing_classif)))  # add clones


```

# add items where nutrition estimated by models 
```{r}

# 0) Canonicalizer (match the one you use later)
if (!exists("canonicalize_item")) {
  clean_key <- function(x) x %>% str_trim() %>% str_squish() %>% str_to_lower()
  canonicalize_item <- function(x) {
    clean_key(x) %>%
      str_replace("^unspecified\\s+molasses$", "molasses") %>%
      str_replace_all("\\bbran\\b", "offal")
  }
}

# 1) Point to your CSV (edit the path as needed)
# csv_path <- file.choose()
csv_path <- "data/Ingredients_nutrition_unmatched_GPT.csv"  # <- change if saved elsewhere

# 2) Read & clean the CSV (strip tildes, parse numbers)
csv_raw <- readr::read_csv(csv_path, show_col_types = FALSE)


# 2) Clean + build canonical join key; parse numbers (robust to "NA", "~")
csv_clean <- csv_raw %>%
  mutate(
    D.Item_can = canonicalize_item(D.Item),
    GE  = suppressWarnings(readr::parse_number(as.character(GE))),
    Ash = suppressWarnings(readr::parse_number(as.character(Ash))),
    CP  = suppressWarnings(readr::parse_number(as.character(CP))),
    EE  = suppressWarnings(readr::parse_number(as.character(EE))),
    NDF = suppressWarnings(readr::parse_number(as.character(NDF)))
  ) %>%
  select(D.Item_can, GE, Ash, CP, EE, NDF)

# 3) Long-format replacement table (only non-missing values)
csv_long <- csv_clean %>%
  pivot_longer(
    cols = c(GE, Ash, CP, EE, NDF),
    names_to = "DC.Variable",
    values_to = "DC.Value_new"
  ) %>%
  filter(!is.na(DC.Value_new))

# 4) Prepare Ing_classif with same canonical key
Ing_classif <- Ing_classif %>%
  mutate(
    D.Item_can = canonicalize_item(D.Item),
    DC.Value = ifelse(
      DC.Variable %in% c("GE","Ash","CP","EE","NDF"),
      suppressWarnings(as.numeric(DC.Value)),
      DC.Value
    )
  )

# 5) Join and override ONLY where CSV provides a value
Ing_classif <- Ing_classif %>%
  left_join(csv_long, by = c("D.Item_can", "DC.Variable")) %>%
  mutate(
    overridden = !is.na(DC.Value_new),
    DC.Value   = if_else(overridden, DC.Value_new, DC.Value)
  ) %>%
  select(-DC.Value_new)

# (Optional) Quick report: how many values were actually overridden
override_counts <- Ing_classif %>%
  filter(overridden) %>%
  count(DC.Variable, name = "n_overridden")
print(override_counts)

# Drop helper flag
Ing_classif$overridden <- NULL
# ---- end CSV override block ----
```


```{r}
# --- Helpers ---------------------------------------------------------------
clean_key <- function(x) x %>% str_trim() %>% str_squish() %>% str_to_lower()

# Canonicalize names so duplicates collapse:
#   "unspecified molasses" → "molasses"
#   any "bran"             → "offal"
canonicalize_item <- function(x) {
  clean_key(x) %>%
    str_replace("^unspecified\\s+molasses$", "molasses") %>%
    str_replace_all("\\bbran\\b", "offal")
}

# Robust logical checks for columns stored as logical or "TRUE"/"FALSE"
is_false_or_na <- function(x) is.na(x) | x == FALSE | x == "FALSE"
is_true        <- function(x) !is.na(x) & (x == TRUE | x == "TRUE")

# Global D.Item → D.Type (built on canonical names)
type_lookup <- diets %>% 
  transmute(
    D.Item = canonicalize_item(D.Item),
    D.Type = str_trim(D.Type)
  ) %>%
  distinct(D.Item, .keep_all = TRUE)

# Canonicalize Ing_classif once
Ing_base <- Ing_classif %>%
  mutate(
    B.Code     = str_trim(B.Code),
    D.Item_raw = D.Item,
    D.Item     = canonicalize_item(D.Item)
  )

# Build (pivot-wide + attach D.Type; optional back-fill from global item means)
build_nutr_table <- function(df_long, type_lookup, backfill_by_item = TRUE) {
  target_vars <- c("CP","EE","NDF","Ash","GE")

  nutr <- df_long %>%
    filter(DC.Variable %in% target_vars) %>%
    mutate(DC.Value = suppressWarnings(as.numeric(DC.Value))) %>%
    pivot_wider(
      id_cols     = c(B.Code, D.Item),
      names_from  = DC.Variable,
      values_from = DC.Value,
      values_fn   = ~ mean(.x, na.rm = TRUE),
      values_fill = NA
    ) %>%
    left_join(type_lookup, by = "D.Item")

  if (backfill_by_item) {
    global_means <- df_long %>%
      filter(DC.Variable %in% target_vars) %>%
      mutate(DC.Value = suppressWarnings(as.numeric(DC.Value))) %>%
      group_by(D.Item, DC.Variable) %>%
      summarise(Value = mean(DC.Value, na.rm = TRUE), .groups = "drop") %>%
      pivot_wider(names_from = DC.Variable, values_from = Value)

    nutr <- nutr %>%
      left_join(global_means, by = "D.Item", suffix = c("", ".global")) %>%
      mutate(
        GE  = coalesce(GE,  GE.global),
        Ash = coalesce(Ash, Ash.global),
        CP  = coalesce(CP,  CP.global),
        EE  = coalesce(EE,  EE.global),
        NDF = coalesce(NDF, NDF.global)
      ) %>%
      select(-ends_with(".global"))
  }

  nutr
}

# Ingredients subset (non-entire, non-group) — WITH back-fill
Ing_ingredients_long <- Ing_base %>%
  filter(is_false_or_na(is_entire_diet)) %>%
  filter(is_false_or_na(is_group))

nutr_ing <- build_nutr_table(Ing_ingredients_long, type_lookup, backfill_by_item = TRUE) %>%
  mutate(scope = "ingredient") %>%
  relocate(scope, .before = B.Code)

# Diets subset (entire diets, non-group) — NO back-fill
Ing_diets_long <- Ing_base %>%
  filter(is_true(is_entire_diet)) %>%
  filter(is_false_or_na(is_group))

nutr_diet <- build_nutr_table(Ing_diets_long, type_lookup, backfill_by_item = FALSE) %>%
  mutate(scope = "diet") %>%
  relocate(scope, .before = B.Code)

# Combined (optional)
nutr_all <- bind_rows(nutr_ing, nutr_diet)

# Diagnostics: items without a mapped D.Type
unmatched_ing  <- anti_join(nutr_ing,  type_lookup, by = "D.Item")
unmatched_diet <- anti_join(nutr_diet, type_lookup, by = "D.Item")

# ── Canonical mean per ingredient (collapse to one row per D.Item) ──────────
canon_ing <- nutr_ing %>%
  group_by(D.Item) %>%
  summarise(
    GE  = mean(GE,  na.rm = TRUE),
    Ash = mean(Ash, na.rm = TRUE),
    CP  = mean(CP,  na.rm = TRUE),
    EE  = mean(EE,  na.rm = TRUE),
    NDF = mean(NDF, na.rm = TRUE),
    D.Type = first(D.Type),
    .groups = "drop"
  )

# Numeric windows (g/kg DM; GE in MJ/kg DM)
win <- list(
  GE  = list(I=c(17,19), II=c(30,38), III=c(18,20), IV=c(16,18), V=c(15,17)),
  Ash = list(I=c(20,40),   II=c(0,10),   III=c(50,80),   IV=c(70,100),  V=c(60,100), VI=c(500,1000)),
  CP  = list(I=c(80,120),  II=c(0,50),   III=c(350,500), IV=c(120,200), V=c(30,80)),
  EE  = list(I=c(20,50),   II=c(80,990), III=c(20,60),   IV=c(10,30),   V=c(0,20)),
  NDF = list(I=c(100,200), II=c(0,50),   III=c(100,300), IV=c(400,500), V=c(600,750))
)
expand10 <- function(x) c(x[1]*0.9, x[2]*1.1)
inside   <- function(x, rng) !is.na(x) & x >= rng[1] & x <= rng[2]

# Priority variables
priority <- list(
  I  = c("NDF"),
  II = c("EE"),
  III= c("CP","NDF"),
  IV = c("NDF","CP"),
  V  = c("NDF","CP"),
  VI = c("Ash")
)

# M1 strict all-five
method1 <- function(df, w = win){
  df %>% mutate(
    M1_Category = case_when(
      inside(GE,w$GE$I)   & inside(Ash,w$Ash$I)  & inside(CP,w$CP$I)   & inside(EE,w$EE$I)   & inside(NDF,w$NDF$I)   ~ "I",
      inside(GE,w$GE$II)  & inside(Ash,w$Ash$II) & inside(CP,w$CP$II)  & inside(EE,w$EE$II)  & inside(NDF,w$NDF$II)  ~ "II",
      inside(GE,w$GE$III) & inside(Ash,w$Ash$III)& inside(CP,w$CP$III) & inside(EE,w$EE$III) & inside(NDF,w$NDF$III) ~ "III",
      inside(GE,w$GE$IV)  & inside(Ash,w$Ash$IV) & inside(CP,w$CP$IV)  & inside(EE,w$EE$IV)  & inside(NDF,w$NDF$IV)  ~ "IV",
      inside(GE,w$GE$V)   & inside(Ash,w$Ash$V)  & inside(CP,w$CP$V)   & inside(EE,w$EE$V)   & inside(NDF,w$NDF$V)   ~ "V",
      inside(Ash,w$Ash$VI) ~ "VI",
      TRUE ~ NA_character_
    )
  )
}

# M2 ±10% all-five
method2 <- function(df){
  w10 <- map(win, ~ map(.x, expand10))
  method1(df, w10) %>% rename(M2_Category = M1_Category)
}

# Priority helper
method_priority <- function(df, w, out_name){
  df %>% rowwise() %>% mutate(
    "{out_name}" := {
      dat <- c(CP=CP, EE=EE, NDF=NDF, Ash=Ash, GE=GE)
      cat <- NA_character_
      for (cl in names(priority)) {
        req <- priority[[cl]]
        if (all(!is.na(dat[req])) &&
            all(map2_lgl(dat[req], req, \(v,r) inside(v, w[[r]][[cl]])))) {
          cat <- cl; break
        }
      }
      cat
    }
  ) %>% ungroup()
}
method3 <- function(df) method_priority(df, win,  "M3_Category")
method4 <- function(df){ w10 <- map(win, ~ map(.x, expand10)); method_priority(df, w10, "M4_Category") }

# Decision tree
tree_decide <- function(df, w){
  df %>% mutate(
    Category = case_when(
      EE  >= w$EE$II[1]                         ~ "II",
      Ash >= w$Ash$VI[1]                        ~ "VI",
      CP  >= w$CP$III[1] & NDF <= w$NDF$III[2]  ~ "III",
      NDF <= w$NDF$I[2]                         ~ "I",
      NDF >= w$NDF$V[1]                         ~ "V",
      NDF >= w$NDF$IV[1] & NDF <= w$NDF$IV[2]   ~ "IV",
      TRUE                                      ~ NA_character_
    )
  )
}
method5 <- function(df) tree_decide(df, win) %>% rename(M5_Category = Category)
method6 <- function(df){ w10 <- map(win, ~ map(.x, expand10)); tree_decide(df, w10) %>% rename(M6_Category = Category) }

# Method 7: start from M6 and apply name/type overrides
method7 <- function(df){
  out <- if ("M6_Category" %in% names(df)) df else method6(df)
  out$M7_Category <- out$M6_Category

  # normalize item names
  item_norm <- canonicalize_item(out$D.Item)

  # --- Category II: fats/oils ---
  oil_pat <- "\\b(oil|tallow|grease|lard)\\b"
  is_oil  <- str_detect(item_norm, regex(oil_pat, ignore_case = TRUE))
  out$M7_Category[is_oil] <- "II"

  # --- Category III: protein sources ---
  protein_pat <- "\\b(distillers?.*grain|soybean ground|canola meal|fish ground|blood ground)\\b"
  is_pmeal    <- str_detect(item_norm, regex(protein_pat, ignore_case = TRUE))
  is_fish     <- str_detect(item_norm, regex("\\bfish\\b", ignore_case = TRUE)) &
                 !str_detect(item_norm, regex("\\bfish\\s*oil\\b", ignore_case = TRUE))
  is_animprot <- str_detect(item_norm, regex("\\b(feather|larvae|snail|termite|shrimp|hydrolysate)\\b",
                                             ignore_case = TRUE))
  out$M7_Category[is_pmeal | is_fish | is_animprot] <- "III"

  # --- Category I: byproducts (bread waste only) ---
  is_bread <- str_detect(item_norm, regex("\\bbread waste\\b", ignore_case = TRUE))
  out$M7_Category[is_bread] <- "I"

  # --- Category VI: supplements / additives ---
  sup_types   <- c("Supplement", "Non-ERA Feed")
  vi_type_hit <- !is.na(out$D.Type) &
                 (out$D.Type %in% sup_types |
                  str_detect(out$D.Type, regex("supplement", ignore_case = TRUE)))
  is_shell    <- str_detect(item_norm, regex("\\boyster\\s*shells?\\b", ignore_case = TRUE))
  is_cellulose <- str_detect(
    item_norm,
    regex("\\b(cellulose|micro\\s*-?crystalline\\s+cellulose|carboxy\\s*methyl\\s*cellulose|\\bMCC\\b|\\bCMC\\b)\\b",
          ignore_case = TRUE)
  )
  is_nonfeed <- str_detect(item_norm,
                           regex("\\b(cement|polyethylene|humate|probiotic|solution|mixture|axtra xap)\\b",
                                 ignore_case = TRUE))
  out$M7_Category[vi_type_hit | is_shell | is_cellulose | is_nonfeed] <- "VI"

  out
}



# Run all methods so we can inspect coverage; M7 uses M6 base + overrides
classif_ing <- canon_ing %>%   # ← use canonical means for INGREDIENTS
  method1() %>% method2() %>% method3() %>% method4() %>%
  method5() %>% method6() %>% method7()

classif_diet <- nutr_diet %>%  # ← keep DIETS per-row (no mean collapse)
  method1() %>% method2() %>% method3() %>% method4() %>%
  method5() %>% method6() %>% method7()



```

```{r}
# Helper to build coverage/counts/samples/unclassified per table
build_reports <- function(df){
  method_cols <- grep("^M[1-7]_Category$", names(df), value = TRUE)

  coverage_tbl <- purrr::map_dfr(method_cols, function(mc){
    tibble(
      Method   = sub("_Category$", "", mc),
      Total    = nrow(df),
      Assigned = sum(!is.na(df[[mc]])),
      Coverage = round(Assigned / Total * 100, 1)
    )
  })

  counts_tbl <- df %>%
    pivot_longer(all_of(method_cols), names_to = "Method", values_to = "Category") %>%
    filter(!is.na(Category)) %>%
    mutate(Method = sub("_Category$", "", Method)) %>%
    count(Method, Category, name = "n") %>%
    arrange(Method, desc(n))

  set.seed(123)
sample_tbl <- df %>%
  pivot_longer(all_of(method_cols), names_to = "Method", values_to = "Category") %>%
  filter(!is.na(Category)) %>%
  mutate(Method = sub("_Category$", "", Method)) %>%
  group_by(Method, Category) %>%
  # take distinct items per (Method,Category), sample up to 10 per group
  group_modify(function(.x, .y) {
    .x %>%
      distinct(D.Item, .keep_all = TRUE) %>%
      slice_sample(n = min(10, nrow(.)))
  }) %>%
  # still grouped by Method,Category here → summarise keeps those cols
  summarise(Examples = paste(D.Item, collapse = "; "), .groups = "drop") %>%
  arrange(Method, Category)


  unclassified_tbl <- df %>%
    pivot_longer(all_of(method_cols), names_to = "Method", values_to = "Category") %>%
    filter(is.na(Category)) %>%
    mutate(Method = sub("_Category$", "", Method)) %>%
    count(Method, D.Item, sort = TRUE) %>%
    group_by(Method) %>%
    slice_max(n, n = 10, with_ties = FALSE) %>%
    ungroup()

  list(coverage = coverage_tbl, counts = counts_tbl,
       samples = sample_tbl, unclassified = unclassified_tbl)
}

rep_ing  <- build_reports(classif_ing)
rep_diet <- build_reports(classif_diet)

# Display (ingredients)
datatable(classif_ing %>% arrange(M7_Category),
          caption = "INGREDIENTS: Classified rows (sorted by M7)",
          rownames = FALSE, options = list(pageLength = 30, dom = "tip"))
datatable(rep_ing$coverage,    caption = "INGREDIENTS: Coverage by method",    options = list(pageLength = 10, dom = "tip"))
datatable(rep_ing$counts,      caption = "INGREDIENTS: Counts per category",   options = list(pageLength = 25, dom = "tip"))
datatable(rep_ing$samples,     caption = "INGREDIENTS: Random examples",       options = list(pageLength = 25, dom = "tip"))
datatable(rep_ing$unclassified,caption = "INGREDIENTS: Top-10 unclassified",   options = list(pageLength = 25, dom = "tip"))

# Display (diets)
datatable(classif_diet %>% arrange(M7_Category),
          caption = "DIETS: Classified rows (sorted by M7)",
          rownames = FALSE, options = list(pageLength = 30, dom = "tip"))
datatable(rep_diet$coverage,    caption = "DIETS: Coverage by method",   options = list(pageLength = 10, dom = "tip"))
datatable(rep_diet$counts,      caption = "DIETS: Counts per category",  options = list(pageLength = 25, dom = "tip"))
datatable(rep_diet$samples,     caption = "DIETS: Random examples",      options = list(pageLength = 25, dom = "tip"))
datatable(rep_diet$unclassified,caption = "DIETS: Top-10 unclassified",  options = list(pageLength = 25, dom = "tip"))

```


```{r}
# Canonicalize D.Item and A.Level.Name in ingredients before joining
ingredients_joinable <- ingredients %>%
  mutate(
    D.Item = canonicalize_item(D.Item),
    A.Level.Name = canonicalize_item(A.Level.Name)
  )

# Simplify classification tables (already canonicalized)
classif_ing_simple <- classif_ing %>%
  select(D.Item, Ingredient_Category = M7_Category)

classif_diet_simple <- classif_diet %>%
  rename(A.Level.Name = D.Item) %>%
  select(A.Level.Name, Diet_Category = M7_Category)

# Join the classifications
ingredients<- ingredients_joinable %>%
  left_join(classif_ing_simple,  by = "D.Item")%>%
  distinct()

# --- Final polish: map Roman numerals -> full names in ingredients ---
roman_to_name <- function(x) dplyr::recode(
  x,
  "I"   = "High-starch",
  "II"  = "High-fat",
  "III" = "Protein",
  "IV"  = "Good forage",
  "V"   = "Poor forage",
  "VI"  = "Mineral/Vitamin",
  .default = x,
  .missing = NA_character_
)

# Update the ingredient database
ingredients <- ingredients %>%
  dplyr::mutate(
    Ingredient_Category = roman_to_name(Ingredient_Category))
    # If you also keep a diet-level tag, uncomment:
    # , Diet_Category = roman_to_name(Diet_Category)
  

```


```{r}
# ----------------------------
# Completeness within selected papers
# ----------------------------

# # Selected papers (already built above)
# selected_bcodes <- GE_data   # a data.frame with B.Code from your earlier step
# 
# # Ensure per-row INGREDIENT classifications exist (ingredients, by B.Code)
# classif_ing_rows <- nutr_ing %>%
#   method1() %>% method2() %>% method3() %>% method4() %>%
#   method5() %>% method6() %>% method7()
# 
# # Choose which method’s category to test for completeness
# method_col <- "M7_Category"
# 
# # --- Diet completeness (all diet rows classified for a paper) ---
# diet_status <- classif_diet %>%
#   semi_join(selected_bcodes, by = "B.Code") %>%
#   group_by(B.Code) %>%
#   summarise(
#     diet_n             = n(),
#     diet_n_classified  = sum(!is.na(.data[[method_col]])),
#     diets_all_classified = diet_n > 0 & (diet_n_classified == diet_n),
#     .groups = "drop"
#   )
# 
# # --- Ingredient completeness (all ingredient rows classified for a paper) ---
# ing_status <- classif_ing_rows %>%
#   semi_join(selected_bcodes, by = "B.Code") %>%
#   group_by(B.Code) %>%
#   summarise(
#     ing_n                = n(),
#     ing_n_classified     = sum(!is.na(.data[[method_col]])),
#     ingredients_all_classified = ing_n > 0 & (ing_n_classified == ing_n),
#     .groups = "drop"
#   )
# 
# # Combine statuses; treat missing as FALSE (paper may have diets but no ingredients, or vice versa)
# paper_status <- full_join(diet_status, ing_status, by = "B.Code") %>%
#   mutate(
#     diets_all_classified       = replace_na(diets_all_classified, FALSE),
#     ingredients_all_classified = replace_na(ingredients_all_classified, FALSE),
#     either_complete            = diets_all_classified | ingredients_all_classified
#   )
# 
# # Attach species for those selected papers (Cattle/Sheep/Goat)
# paper_status <- paper_status %>%
#   left_join(
#     merged_metadata$`Prod.Out` %>%
#       filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
#       distinct(B.Code, P.Product),
#     by = "B.Code"
#   )
# 
# # ---------- Totals ----------
# n_papers_diets_all        <- sum(paper_status$diets_all_classified)
# n_papers_ingredients_all  <- sum(paper_status$ingredients_all_classified)
# n_papers_either           <- sum(paper_status$either_complete)
# 
# n_papers_diets_all
# n_papers_ingredients_all
# n_papers_either
# 
# # ---------- By species ----------
# summary_by_product <- paper_status %>%
#   group_by(P.Product) %>%
#   summarise(
#     Papers_total            = n_distinct(B.Code),
#     Papers_diets_all        = sum(diets_all_classified),
#     Papers_ingredients_all  = sum(ingredients_all_classified),
#     Papers_either           = sum(either_complete),
#     .groups = "drop"
#   )
# 
# # View as a table
# kable(summary_by_product, caption = "Papers with complete diet/ingredient classification (within selected papers)")

```

```{r}
ingredients <-ingredients%>%
  filter(D.Item!="salt")

unmatched_ing_m7 <- ingredients %>%
  filter(is.na(Ingredient_Category)) %>%
  distinct(D.Item) %>%
  left_join(canon_ing, by = "D.Item") %>%
  arrange(D.Item)
```

### k means for ingredients 
```{r}


# ---------- Windows & priority (your spec) ----------
win <- list(
  GE  = list(I=c(17,19), II=c(30,38), III=c(18,20), IV=c(16,18), V=c(15,17)),
  Ash = list(I=c(20,40),   II=c(0,10),   III=c(50,80),   IV=c(70,100),  V=c(60,100), VI=c(500,1000)),
  CP  = list(I=c(80,120),  II=c(0,50),   III=c(350,500), IV=c(120,200), V=c(30,80)),
  EE  = list(I=c(20,50),   II=c(80,990), III=c(20,60),   IV=c(10,30),   V=c(0,20)),
  NDF = list(I=c(100,200), II=c(0,50),   III=c(100,300), IV=c(400,500), V=c(600,750))
)
priority <- list(
  I  = c("NDF","CP"),
  II = c("EE"),
  III= c("CP","NDF"),
  IV = c("NDF","CP"),
  V  = c("NDF","CP"),
  VI = c("Ash")
)

expand10 <- function(x) c(x[1]*0.9, x[2]*1.1)
mid      <- function(r) mean(r, na.rm = TRUE)
w10 <- map(win, ~ map(.x, expand10))  # ±10% windows (Method 6 spirit)

# ---------- 1) Build the working table from *your* dataset ----------
# Expecting unmatched_ing_m7 to already contain D.Item (+ optionally D.Type) and nutrients
unmatched <- unmatched_ing_m7 %>%
  select(any_of(c("D.Item","D.Type","GE","Ash","CP","EE","NDF"))) %>%
  group_by(across(any_of(c("D.Item","D.Type")))) %>%
  summarise(across(c(GE, Ash, CP, EE, NDF),
                   ~ suppressWarnings(mean(as.numeric(.), na.rm = TRUE))),
            .groups = "drop") %>%
  mutate(across(c(GE, Ash, CP, EE, NDF), ~ ifelse(is.nan(.), NA_real_, .))) %>%
  # drop rows where ALL five nutrients are NA
  filter(!(is.na(GE) & is.na(Ash) & is.na(CP) & is.na(EE) & is.na(NDF))) %>%
  mutate(
    n_values = rowSums(!is.na(across(c(GE, Ash, CP, EE, NDF)))),
    Missing  = pmap_chr(select(., GE, Ash, CP, EE, NDF), ~ {
      nm <- c("GE","Ash","CP","EE","NDF"); vv <- c(...); miss <- nm[is.na(vv)]
      if (length(miss)==0) "—" else paste(miss, collapse = ", ")
    })
  )

# ---------- 2) Priority-based score with HARD GATES ----------
# Standardized violation (0 if inside; >0 if outside by fraction of window width)
std_violation <- function(x, rng) {
  if (is.na(x) || any(is.na(rng))) return(NA_real_)
  L <- rng[1]; U <- rng[2]; W <- U - L; if (W <= 0) return(NA_real_)
  if (x < L) (L - x)/W else if (x > U) (x - U)/W else 0
}

cats <- names(priority)
missing_penalty <- 0.75  # penalty if a priority var is missing

scores <- matrix(NA_real_, nrow = nrow(unmatched), ncol = length(cats))
colnames(scores) <- cats

# Hard gates (fixes the “all go to II” issue):
gate_ii_lb  <- w10$EE$II[1]   # 72 (0.9 * 80)
gate_vi_lb  <- w10$Ash$VI[1]  # 450 (0.9 * 500)

for (j in seq_along(cats)) {
  cat <- cats[j]
  req <- priority[[cat]]

  # Per-variable violations for this category
  vlist <- lapply(req, function(v) {
    x   <- unmatched[[v]] %>% as.numeric()
    rng <- w10[[v]][[cat]]
    sapply(x, std_violation, rng = rng)
  })
  vmat <- do.call(cbind, vlist)

  # Replace NA violations (missing values) with bounded penalty
  vmat[is.na(vmat)] <- missing_penalty

  # Mean violation across this category’s priority vars
  s <- rowMeans(vmat)

  # Apply gates:
  if (cat == "II") {
    ee <- unmatched$EE
    s[is.na(ee) | ee < gate_ii_lb] <- Inf
  }
  if (cat == "VI") {
    ash <- unmatched$Ash
    s[is.na(ash) | ash < gate_vi_lb] <- Inf
  }

  scores[, j] <- s
}

best_idx   <- apply(scores, 1, which.min)
second_idx <- apply(scores, 1, function(v) order(v, na.last = TRUE)[2])

Assigned_Category    <- cats[best_idx]
Distance_Closest     <- scores[cbind(seq_len(nrow(scores)), best_idx)]
Second_Best_Category <- cats[second_idx]
Distance_Second      <- scores[cbind(seq_len(nrow(scores)), second_idx)]

# How many priority vars actually present for the chosen cat?
n_used_priority <- map2_int(seq_len(nrow(unmatched)), best_idx, function(i, j) {
  req <- priority[[cats[j]]]
  sum(!is.na(unmatched[i, req]))
})
Priority_vars_used <- map2_chr(seq_len(nrow(unmatched)), best_idx, function(i, j) {
  req <- priority[[cats[j]]]; have <- req[!is.na(unmatched[i, req])]
  if (length(have)==0) "—" else paste(have, collapse = ", ")
})
Priority_vars_missing <- map2_chr(seq_len(nrow(unmatched)), best_idx, function(i, j) {
  req <- priority[[cats[j]]]; miss <- req[is.na(unmatched[i, req])]
  if (length(miss)==0) "—" else paste(miss, collapse = ", ")
})

results <- unmatched %>%
  mutate(
    Assigned_Category    = Assigned_Category,
    Distance_Closest     = round(Distance_Closest, 3),
    Second_Best_Category = Second_Best_Category,
    Distance_Second      = round(Distance_Second, 3),
    n_used_priority      = n_used_priority,
    Priority_vars_used   = Priority_vars_used,
    Priority_vars_missing= Priority_vars_missing
  ) %>%
  arrange(Assigned_Category, Distance_Closest, desc(n_values), D.Item)

DT::datatable(
  results %>% select(D.Item, any_of("D.Type"), GE, Ash, CP, EE, NDF,
                     n_values, Missing,
                     Assigned_Category, Distance_Closest,
                     Second_Best_Category, Distance_Second,
                     n_used_priority, Priority_vars_used, Priority_vars_missing),
  caption = "M7-unclassified (from unmatched_ing_m7) → priority-window assignment (±10%), with gates for II/VI",
  options = list(pageLength = 50, dom = "tip"),
  rownames = FALSE
)

# ---------- 3) Quick plots (centroids midpoints + item segments) ----------
centroids_full <- tibble(
  Category = c("I","II","III","IV","V","VI"),
  GE  = c(mid(win$GE$I),  mid(win$GE$II),  mid(win$GE$III),  mid(win$GE$IV),  mid(win$GE$V),  NA_real_),
  Ash = c(mid(win$Ash$I), mid(win$Ash$II), mid(win$Ash$III), mid(win$Ash$IV), mid(win$Ash$V), mid(win$Ash$VI)),
  CP  = c(mid(win$CP$I),  mid(win$CP$II),  mid(win$CP$III),  mid(win$CP$IV),  mid(win$CP$V),  NA_real_),
  EE  = c(mid(win$EE$I),  mid(win$EE$II),  mid(win$EE$III),  mid(win$EE$IV),  mid(win$EE$V),  NA_real_),
  NDF = c(mid(win$NDF$I), mid(win$NDF$II), mid(win$NDF$III), mid(win$NDF$IV), mid(win$NDF$V), NA_real_)
)

plot_pair <- function(items_df, cents_df, xvar, yvar) {
  items_xy <- items_df %>% filter(!is.na(.data[[xvar]]), !is.na(.data[[yvar]]))
  cents_xy <- cents_df %>% filter(!is.na(.data[[xvar]]), !is.na(.data[[yvar]])) %>%
    transmute(Category, cx = .data[[xvar]], cy = .data[[yvar]])
  items_xy <- items_xy %>% left_join(cents_xy, by = c("Assigned_Category" = "Category"))

  ggplot() +
    geom_point(data = items_xy,
               aes(x = .data[[xvar]], y = .data[[yvar]], color = Assigned_Category),
               size = 2, alpha = 0.9) +
    geom_segment(data = items_xy %>% filter(!is.na(cx), !is.na(cy)),
                 aes(x = .data[[xvar]], y = .data[[yvar]], xend = cx, yend = cy, color = Assigned_Category),
                 alpha = 0.4, linewidth = 0.6) +
    geom_point(data = cents_xy, aes(x = cx, y = cy), shape = 4, size = 4, stroke = 1.1) +
    geom_text(data = cents_xy, aes(x = cx, y = cy, label = Category), vjust = -1, fontface = "bold") +
    labs(x = xvar, y = yvar, color = "Assigned",
         title = paste(xvar, "vs", yvar, "— items and category centroids")) +
    theme_minimal()
}

print(plot_pair(results, centroids_full, "CP", "NDF"))
print(plot_pair(results, centroids_full, "EE", "NDF"))

```


```{r}
# ---- Assign the first (best) category back to `ingredients` ----
# Roman → full name map (keep if you already have it)
if (!exists("roman_to_name")) {
  roman_to_name <- function(x) dplyr::recode(
    x,
    "I"   = "High-starch",
    "II"  = "High-fat",
    "III" = "Protein",
    "IV"  = "Good forage",
    "V"   = "Poor forage",
    "VI"  = "Mineral/Vitamin",
    .default = x, .missing = NA_character_
  )
}

# Join keys (prefer D.Item + D.Type if you have both)
join_keys <- intersect(c("D.Item","D.Type"), names(ingredients))

# Make the assignment table unique on the join keys (prevents many-to-many)
assign_tbl <- results %>%
  mutate(Assigned_Category_Name = roman_to_name(Assigned_Category)) %>%
  select(any_of(c("D.Item","D.Type")), Assigned_Category, Assigned_Category_Name) %>%
  group_by(across(all_of(join_keys))) %>%
  summarise(
    Assigned_Category      = dplyr::first(Assigned_Category),
    Assigned_Category_Name = dplyr::first(Assigned_Category_Name),
    .groups = "drop"
  )

# Keep a backup of existing labels the first time you run this
if (!"Ingredient_Category_orig" %in% names(ingredients)) {
  ingredients <- ingredients %>%
    mutate(Ingredient_Category_orig = Ingredient_Category)
}

# Ensure Category_Source exists (so mutate below can reference it)
if (!"Category_Source" %in% names(ingredients)) {
  ingredients <- ingredients %>% mutate(Category_Source = NA_character_)
}

# Join and fill ONLY where Ingredient_Category is NA
ingredients <- ingredients %>%
  # dplyr >= 1.1: relationship arg suppresses warning if truly many-to-one
  left_join(assign_tbl, by = join_keys, relationship = "many-to-one") %>%
  mutate(
    Ingredient_Category = coalesce(Ingredient_Category, Assigned_Category_Name),
    Category_Source     = ifelse(
      is.na(Ingredient_Category_orig) & !is.na(Assigned_Category_Name),
      "Nearest-window (priority, ±10%, gated II/VI)",
      Category_Source
    )
  ) %>%
  select(-Assigned_Category, -Assigned_Category_Name)

# Quick sanity check
filled_n <- sum(is.na(ingredients$Ingredient_Category_orig) & !is.na(ingredients$Ingredient_Category))
message("Filled missing Ingredient_Category for ", filled_n, " ingredient rows.")

```

```{r}
ingredients <-ingredients%>%
  filter(D.Item!="salt")

unmatched_ing_m7 <- ingredients %>%
  filter(is.na(Ingredient_Category)) %>%
  distinct(D.Item) %>%
  left_join(canon_ing, by = "D.Item") %>%
  arrange(D.Item)
```


## classifying diets 


```{r}
library(dplyr)
# ─────────────────────────────────────────────────────────────────────────────
# Reclassify ONLY unmatched diets, then merge back
# Requires in env:
#   unmatched_diet (B.Code, D.Item ...), diet_percentage, canon_ing, classif_ing,
#   classif_diet (initial classifications), canonicalize_item(), method1..method7()
# ─────────────────────────────────────────────────────────────────────────────

unmatched_diet <- classif_diet %>%
  # standardize the diet key
  # not yet classified by M7
  filter(is.na(M7_Category)) %>%
  select(any_of(c(
    "scope","B.Code","D.Item","CP","NDF","Ash","EE","GE","M7_Category"
  ))) %>%
  distinct()



unmatched_keys <- unmatched_diet %>%
  distinct(B.Code, D.Item)


# 1) Diet % for unmatched diets only (canonical keys + proportions)
diet_perc_unmatched <- diet_percentages %>%
  mutate(
    w_prop       = suppressWarnings(as.numeric(Percentage_of_Diet))/100
  ) %>%
  semi_join(unmatched_keys, by = c("B.Code","A.Level.Name" = "D.Item"))

# 2) Build whole-diet nutrients when ALL ingredients have GE/Ash/CP/EE/NDF
diet_perc_nutr <- diet_perc_unmatched %>%
  left_join(classif_diet %>% select(D.Item, GE, Ash, CP, EE, NDF), by = "D.Item")

diet_ok <- diet_perc_nutr %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    n_ing = n(),
    all_complete = all(!is.na(GE) & !is.na(Ash) & !is.na(CP) & !is.na(EE) & !is.na(NDF) & !is.na(w_prop)),
    .groups = "drop"
  ) %>% filter(all_complete)

diet_nutrients_from_perc_unmatched <- diet_perc_nutr %>%
  semi_join(diet_ok, by = c("B.Code","A.Level.Name")) %>%
  group_by(B.Code, A.Level.Name) %>%
  summarise(
    wsum   = sum(w_prop, na.rm = TRUE),
    GE     = sum(GE  * w_prop, na.rm = TRUE) / wsum,
    Ash    = sum(Ash * w_prop, na.rm = TRUE) / wsum,
    CP     = sum(CP  * w_prop, na.rm = TRUE) / wsum,
    EE     = sum(EE  * w_prop, na.rm = TRUE) / wsum,
    NDF    = sum(NDF * w_prop, na.rm = TRUE) / wsum,
    .groups = "drop"
  ) %>%
  mutate(D.Item = A.Level.Name) %>%
  select(B.Code, D.Item, GE, Ash, CP, EE, NDF)

# 3) M1–M7 on unmatched diets, preferring computed nutrients where available
nutr_diet_unmatched <- unmatched_keys %>%
  left_join(diet_nutrients_from_perc_unmatched, by = c("B.Code","D.Item"))

classif_diet_unmatched <- nutr_diet_unmatched %>%
  method1() %>% method2() %>% method3() %>% method4() %>%
  method5() %>% method6() %>% method7() %>%
  mutate(Patch_Method = if_else(!is.na(M7_Category), "M7_numeric_from_perc", NA_character_))

# 4) Fallback for any still NA: most prevalent ingredient category (no thresholds)
ing_cat_map <- classif_ing %>%
  select(D.Item, Ingredient_Category=M7_Category) %>%
  distinct()

topcat_unmatched <- diet_perc_unmatched %>%
  left_join(ing_cat_map, by = "D.Item") %>%
  filter(!is.na(Ingredient_Category), !is.na(w_prop)) %>%
  group_by(B.Code, A.Level.Name, Ingredient_Category) %>%
  summarise(prop = sum(w_prop, na.rm = TRUE), .groups = "drop") %>%
  group_by(B.Code, A.Level.Name) %>%
  slice_max(prop, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    B.Code,
    D.Item = A.Level.Name,
    M7_Category_fallback = Ingredient_Category,
    Patch_Method_fallback = "by_category_dominance",
    Diet_Category_prop_top = prop
  )

# Apply fallback only where M7 still NA
classif_diet_unmatched <- classif_diet_unmatched %>%
  left_join(topcat_unmatched, by = c("B.Code","D.Item")) %>%
  mutate(
    used_fallback = is.na(M7_Category) & !is.na(M7_Category_fallback),
    M7_Category   = coalesce(M7_Category, M7_Category_fallback),
    Patch_Method  = coalesce(Patch_Method, if_else(used_fallback, Patch_Method_fallback, NA_character_))
  ) %>%
  select(-used_fallback, -any_of(c("M7_Category_fallback","Patch_Method_fallback")))

# --- ensure method column exists and stamp initial M7 rows once ---
if (!"Diet_Category_method" %in% names(classif_diet)) {
  classif_diet <- classif_diet %>%
    mutate(Diet_Category_method = if_else(!is.na(M7_Category), "M7_numeric", NA_character_))
}

# 5) Merge the patch back WITHOUT overriding pre-classified diets
classif_diet_2nd <- classif_diet %>%
  left_join(
    classif_diet_unmatched %>%
      select(B.Code, D.Item,
             M7_Category_patch = M7_Category,
             Patch_Method,
             Diet_Category_prop_top),
    by = c("B.Code","D.Item")
  ) %>%
  mutate(
    used_patch  = is.na(M7_Category) & !is.na(M7_Category_patch),
    M7_Category = coalesce(M7_Category, M7_Category_patch),
    Diet_Category_method = case_when(
      used_patch & Patch_Method == "M7_numeric_from_perc"  ~ "M7_numeric_from_perc",
      used_patch & Patch_Method == "by_category_dominance" ~ "by_category_dominance",
      TRUE                                                 ~ Diet_Category_method
    )
  ) %>%
  select(-M7_Category_patch, -Patch_Method) %>%
  relocate(Diet_Category_method, .after = M7_Category)

roman_to_name <- function(x) dplyr::recode(
  x,
  "I"   = "High-starch",
  "II"  = "High-fat",
  "III" = "Protein",
  "IV"  = "Good forage",
  "V"   = "Poor forage",
  "VI"  = "Mineral/Vitamin",
  .default = x, .missing = NA_character_
)

classif_diet_2nd <- classif_diet_2nd %>%
  mutate(across(dplyr::matches("^M[1-7]_Category$"), roman_to_name))

# (Optional) quick coverage report for the patch run
patched_n <- sum(classif_diet_2nd$used_patch, na.rm = TRUE)
message("Patched (newly classified) diets: ", patched_n)
classif_diet$used_patch <- NULL


```
category dominance

```{r}

# ─────────────────────────────────────────────────────────────────────────────
# Coverage & method breakdown (with consistent keys)
#   Uses: diet_percentages, nutr_diet, classif_diet, classif_diet_2nd
#   Assumes: canonicalize_item() exists; if not, define a minimal one here
# ─────────────────────────────────────────────────────────────────────────────

# Fallback canonicalizer if not already defined
if (!exists("canonicalize_item")) {
  clean_key <- function(x) gsub("\\s+", " ", trimws(tolower(x)))
  canonicalize_item <- function(x) {
    x <- clean_key(x)
    x <- gsub("^unspecified\\s+molasses$", "molasses", x)
    x <- gsub("\\bbran\\b", "offal", x)
    x
  }
}

# --- Universe of diets (union of % table and nutrient table), canonicalized ---
all_diets <- dplyr::full_join(
  diet_percentages %>%
    dplyr::transmute(B.Code, DietKey = canonicalize_item(A.Level.Name)) %>%
    dplyr::distinct(),
  nutr_diet %>%
    dplyr::transmute(B.Code, DietKey = canonicalize_item(D.Item)) %>%
    dplyr::distinct(),
  by = c("B.Code","DietKey")
) %>% dplyr::distinct()

# --- Stage flags (all keys canonicalized) ---
# Stage 1: initial M7 (pre-patch)
stage1_flags <- classif_diet %>%
  dplyr::transmute(B.Code,
                   DietKey = canonicalize_item(D.Item),
                   stage1  = !is.na(M7_Category))

# Stage 2: classified via M7 after building diet nutrients from ingredient %
stage2_flags <- classif_diet_2nd %>%
  dplyr::transmute(B.Code,
                   DietKey = canonicalize_item(D.Item),
                   stage2  = Diet_Category_method == "M7_numeric_from_perc")

# Stage 3: classified via dominant ingredient category (fallback)
stage3_flags <- classif_diet_2nd %>%
  dplyr::transmute(B.Code,
                   DietKey = canonicalize_item(D.Item),
                   stage3  = Diet_Category_method == "by_category_dominance")

# --- Combine & compute incremental + cumulative coverage ---
coverage <- all_diets %>%
  dplyr::left_join(stage1_flags, by = c("B.Code","DietKey")) %>%
  dplyr::left_join(stage2_flags, by = c("B.Code","DietKey")) %>%
  dplyr::left_join(stage3_flags, by = c("B.Code","DietKey")) %>%
  dplyr::mutate(dplyr::across(c(stage1, stage2, stage3),
                              ~ tidyr::replace_na(., FALSE))) %>%
  dplyr::summarise(
    diets_total          = dplyr::n(),
    # Stage 1
    classified_stage1    = sum(stage1),
    pct_stage1           = round(100 * classified_stage1 / diets_total, 1),
    # Stage 2 (incremental = not stage1, but stage2)
    inc_stage2           = sum(!stage1 & stage2),
    cum_stage2           = classified_stage1 + inc_stage2,
    pct_cum_stage2       = round(100 * cum_stage2 / diets_total, 1),
    # Stage 3 (incremental = not in stage1/2, but stage3)
    inc_stage3           = sum(!stage1 & !stage2 & stage3),
    cum_final            = cum_stage2 + inc_stage3,
    pct_cum_final        = round(100 * cum_final / diets_total, 1)
  )

print(coverage)

# --- Final method breakdown (use the final table; keys canonicalized) --------
final_method_breakdown <- all_diets %>%
  dplyr::left_join(
    classif_diet_2nd %>%
      dplyr::transmute(B.Code,
                       DietKey = canonicalize_item(D.Item),
                       M7_Category, Diet_Category_method),
    by = c("B.Code","DietKey")
  ) %>%
  dplyr::mutate(method = dplyr::case_when(
    is.na(M7_Category)                              ~ "Unclassified",
    Diet_Category_method == "by_category_dominance" ~ "Fallback: dominance",
    Diet_Category_method == "M7_numeric_from_perc"  ~ "M7 from % nutrients",
    TRUE                                            ~ "M7 initial"
  )) %>%
  dplyr::count(method, name = "n") %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(dplyr::desc(n))

print(final_method_breakdown)

# --- Optional: quick diagnostics for key mismatches --------------------------
# cat("Stage1 keys missing in all_diets: ",
#     nrow(dplyr::anti_join(stage1_flags, all_diets, by = c("B.Code","DietKey"))), "\n")
# cat("Stage2 keys missing in all_diets: ",
#     nrow(dplyr::anti_join(stage2_flags, all_diets, by = c("B.Code","DietKey"))), "\n")
# cat("Stage3 keys missing in all_diets: ",
#     nrow(dplyr::anti_join(stage3_flags, all_diets, by = c("B.Code","DietKey"))), "\n")


```


##k means diet

```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Nearest-window (priority, ±10%, gated II/VI) for remaining unclassified diets
# Works after you have `classif_diet_2nd` (post M7 + % nutrients + dominance).
# ─────────────────────────────────────────────────────────────────────────────

# Reuse windows/priority/helpers if already defined; otherwise define them
if (!exists("win")) {
  win <- list(
    GE  = list(I=c(17,19), II=c(30,38), III=c(18,20), IV=c(16,18), V=c(15,17)),
    Ash = list(I=c(20,40),   II=c(0,10),   III=c(50,80),   IV=c(70,100),  V=c(60,100), VI=c(500,1000)),
    CP  = list(I=c(80,120),  II=c(0,50),   III=c(350,500), IV=c(120,200), V=c(30,80)),
    EE  = list(I=c(20,50),   II=c(80,990), III=c(20,60),   IV=c(10,30),   V=c(0,20)),
    NDF = list(I=c(100,200), II=c(0,50),   III=c(100,300), IV=c(400,500), V=c(600,750))
  )
}
if (!exists("priority")) {
  priority <- list(
    I  = c("NDF","CP"),
    II = c("EE"),
    III= c("CP","NDF"),
    IV = c("NDF","CP"),
    V  = c("NDF","CP"),
    VI = c("Ash")
  )
}
expand10 <- function(x) c(x[1]*0.9, x[2]*1.1)
w10 <- purrr::map(win, ~ purrr::map(.x, expand10))  # ±10% windows

std_violation <- function(x, rng) {
  if (is.na(x) || any(is.na(rng))) return(NA_real_)
  L <- rng[1]; U <- rng[2]; W <- U - L; if (W <= 0) return(NA_real_)
  if (x < L) (L - x)/W else if (x > U) (x - U)/W else 0
}

# 1) Pull remaining unclassified diets that have *some* nutrient info
diets_left <- classif_diet_2nd %>%
  dplyr::filter(is.na(M7_Category)) %>%
  dplyr::mutate(dplyr::across(c(GE, Ash, CP, EE, NDF),
                              ~ suppressWarnings(as.numeric(.)))) %>%
  # drop rows where ALL five are NA
  dplyr::filter(!(is.na(GE) & is.na(Ash) & is.na(CP) & is.na(EE) & is.na(NDF))) %>%
  dplyr::select(B.Code, D.Item, GE, Ash, CP, EE, NDF) %>%
  dplyr::distinct()

# If nothing to do, exit gracefully
if (nrow(diets_left) > 0) {

  # 2) Score each diet against categories (priority variables + gates)
  cats <- names(priority)
  missing_penalty <- 0.75
  gate_ii_lb  <- w10$EE$II[1]   # require EE high enough for cat II
  gate_vi_lb  <- w10$Ash$VI[1]  # require Ash high enough for cat VI

  scores <- matrix(NA_real_, nrow = nrow(diets_left), ncol = length(cats))
  colnames(scores) <- cats

  for (j in seq_along(cats)) {
    cat <- cats[j]
    req <- priority[[cat]]

    vlist <- lapply(req, function(v) {
      x   <- as.numeric(diets_left[[v]])
      rng <- w10[[v]][[cat]]
      vapply(x, std_violation, rng = rng, FUN.VALUE = numeric(1))
    })
    vmat <- do.call(cbind, vlist)
    vmat[is.na(vmat)] <- missing_penalty
    s <- rowMeans(vmat)

    if (cat == "II") {
      ee <- diets_left$EE
      s[is.na(ee) | ee < gate_ii_lb] <- Inf
    }
    if (cat == "VI") {
      ash <- diets_left$Ash
      s[is.na(ash) | ash < gate_vi_lb] <- Inf
    }
    scores[, j] <- s
  }

  best_idx <- apply(scores, 1, which.min)
  second_idx <- apply(scores, 1, function(v) order(v, na.last = TRUE)[2])

  diet_results <- diets_left %>%
    dplyr::mutate(
      Assigned_Category    = cats[best_idx],
      Distance_Closest     = round(scores[cbind(seq_len(nrow(scores)), best_idx)], 3),
      Second_Best_Category = cats[second_idx],
      Distance_Second      = round(scores[cbind(seq_len(nrow(scores)), second_idx)], 3)
    )

  # 3) Merge back *without overriding* any previously assigned categories
  classif_diet_3rd <- classif_diet_2nd %>%
    dplyr::left_join(
      diet_results %>% dplyr::select(B.Code, D.Item, Assigned_Category),
      by = c("B.Code","D.Item")
    ) %>%
    dplyr::mutate(
      used_nearest = is.na(M7_Category) & !is.na(Assigned_Category),
      M7_Category  = dplyr::coalesce(M7_Category, Assigned_Category),
      Diet_Category_method = dplyr::case_when(
        used_nearest ~ "nearest_window_diet",
        TRUE         ~ Diet_Category_method
      )
    ) %>%
    dplyr::select(-Assigned_Category)

  # 4) Optional: add a full-text label for readability
  roman_to_name <- function(x) dplyr::recode(
    x,
    "I"   = "High-starch",
    "II"  = "High-fat",
    "III" = "Protein",
    "IV"  = "Good forage",
    "V"   = "Poor forage",
    "VI"  = "Mineral/Vitamin",
    .default = x, .missing = NA_character_
  )
  classif_diet_3rd <- classif_diet_3rd %>%
    dplyr::mutate(M7_Category_full = roman_to_name(M7_Category))

  # 5) Replace your working table
  classif_diet_2nd <- classif_diet_3rd

  # 6) Quick message
  newly_classified <- sum(classif_diet_2nd$Diet_Category_method == "nearest_window_diet", na.rm = TRUE)
  message("Nearest-window (diet) classified additional diets: ", newly_classified)
} else {
  message("No remaining unclassified diets with any nutrient info to process.")
}

```
```{r}
# ─────────────────────────────────────────────────────────────────────────────
# Coverage summary after all methods (initial M7, % nutrients, dominance, nearest)
# ─────────────────────────────────────────────────────────────────────────────

# Universe of diets (union of % table and nutrient table)
all_diets <- dplyr::full_join(
  diet_percentages %>%
    dplyr::transmute(B.Code, DietKey = canonicalize_item(A.Level.Name)) %>%
    dplyr::distinct(),
  nutr_diet %>%
    dplyr::transmute(B.Code, DietKey = D.Item) %>%
    dplyr::distinct(),
  by = c("B.Code","DietKey")
) %>% dplyr::distinct()

# Pull method tags from the FINAL table
flags <- all_diets %>%
  dplyr::left_join(
    classif_diet_2nd %>%
      dplyr::transmute(
        B.Code,
        DietKey = D.Item,
        method = Diet_Category_method,
        classified = !is.na(M7_Category)
      ),
    by = c("B.Code","DietKey")
  ) %>%
  dplyr::mutate(
    stage_init = method == "M7_numeric",
    stage_perc = method == "M7_numeric_from_perc",
    stage_dom  = method == "by_category_dominance",
    stage_near = method == "nearest_window_diet"
  ) %>%
  tidyr::replace_na(list(
    stage_init = FALSE, stage_perc = FALSE, stage_dom = FALSE, stage_near = FALSE,
    classified = FALSE
  ))

# Incremental & cumulative coverage
coverage_all <- flags %>%
  dplyr::summarise(
    diets_total    = dplyr::n(),
    # Stage 1 (initial only)
    classified_s1  = sum(stage_init),
    pct_s1         = round(100 * classified_s1 / diets_total, 1),

    # Stage 2 (new, not in stage 1)
    inc_s2         = sum(!stage_init & stage_perc),
    cum_s2         = classified_s1 + inc_s2,
    pct_cum_s2     = round(100 * cum_s2 / diets_total, 1),

    # Stage 3 (new, not in stage 1/2)
    inc_s3         = sum(!stage_init & !stage_perc & stage_dom),
    cum_s3         = cum_s2 + inc_s3,
    pct_cum_s3     = round(100 * cum_s3 / diets_total, 1),

    # Stage 4 (new, not in stage 1/2/3)
    inc_s4         = sum(!stage_init & !stage_perc & !stage_dom & stage_near),
    cum_final      = cum_s3 + inc_s4,
    pct_cum_final  = round(100 * cum_final / diets_total, 1)
  )

print(coverage_all)

# Final method breakdown (how many diets ended up with each path)
final_method_breakdown <- all_diets %>%
  dplyr::left_join(
    classif_diet_2nd %>% dplyr::select(B.Code, DietKey = D.Item, M7_Category, Diet_Category_method),
    by = c("B.Code","DietKey")
  ) %>%
  dplyr::mutate(method_readable = dplyr::case_when(
    is.na(M7_Category)                          ~ "Unclassified",
    Diet_Category_method == "M7_numeric"        ~ "M7 initial",
    Diet_Category_method == "M7_numeric_from_perc"  ~ "M7 from % nutrients",
    Diet_Category_method == "by_category_dominance" ~ "Fallback: dominance",
    Diet_Category_method == "nearest_window_diet"   ~ "Nearest-window diet",
    TRUE ~ Diet_Category_method
  )) %>%
  dplyr::count(method_readable, name = "n") %>%
  dplyr::mutate(pct = round(100 * n / sum(n), 1)) %>%
  dplyr::arrange(dplyr::desc(n))

print(final_method_breakdown)

```


## papers with ing classified 

```{r}

# 
# # 1) Per-diet completeness: all ingredients classified?
# diet_class_status <- ingredients %>%
#   semi_join(GE_data, by = "B.Code") %>%
#   filter(!is.na(A.Level.Name), !is.na(D.Item)) %>%
#   # (Optionally exclude entire-diet placeholder rows if present)
#   filter(is.na(D.Type) | D.Type != "Entire Diet") %>%
#   group_by(B.Code, A.Level.Name) %>%
#   summarise(
#     n_ing          = n(),
#     n_classified   = sum(!is.na(Ingredient_Category)),
#     all_classified = n_ing > 0 & (n_ing == n_classified),
#     .groups = "drop"
#   )
# 
# # 2) Papers with at least TWO diets fully classified
# papers_2plus <- diet_class_status %>%
#   group_by(B.Code) %>%
#   summarise(n_complete_diets = sum(all_classified), .groups = "drop") %>%
#   filter(n_complete_diets >= 2)
# 
# # 3) Attach product and count per product (Cattle / Sheep / Goat)
# papers_2plus_by_product <- papers_2plus %>%
#   left_join(
#     merged_metadata$`Prod.Out` %>%
#       filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
#       distinct(B.Code, P.Product),
#     by = "B.Code"
#   ) %>%
#   count(P.Product, name = "Papers_with_≥2_fully_classified_diets")
# 
# # Show the summary table
# knitr::kable(
#   papers_2plus_by_product %>% arrange(P.Product),
#   caption = "Papers per product with ≥2 diets where all ingredients are classified"
# )
# 
# # (Optional) Show the actual B.Codes and how many complete diets each has
# knitr::kable(
#   papers_2plus %>%
#     left_join(
#       merged_metadata$`Prod.Out` %>%
#         filter(P.Product %in% c("Cattle", "Sheep", "Goat")) %>%
#         distinct(B.Code, P.Product),
#       by = "B.Code"
#     ) %>%
#     arrange(P.Product, desc(n_complete_diets), B.Code),
#   caption = "B.Codes with number of fully classified diets (within GE papers)"
# )

```

```{r}
write.csv(ingredients, "~/ERL/Emission_Analysis/data/ingredients_classified.csv")
write.csv(classif_diet, "~/ERL/Emission_Analysis/data/classif_diet.csv")
write.csv(unmatched_diet, "~/ERL/Emission_Analysis/data/unmatched_diet.csv")
write.csv(classif_ing_simple, "~/ERL/Emission_Analysis/data/classif_ing_simple.csv")

```

